### 2022-01-26 mr-review.Rmd
initial commit

---
title: "Mendelian randomization for causal inference in Genome-wide association studies"
author: "Frederick J. Boehm & Xiang Zhou"
date: "1/26/2022"
output: 
  - bookdown::word_document2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Great interest has developed in inferring causal relationships between complex 
traits in observational human genetics studies.

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via mendelian randomization.

Methods for Mendelian randomization have appeared at a rapid pace since the 
appearance of the first published GWAS.

Now, GWAS and Mendelian randomization for Biobank-scale data offer the possibility
of discovering even more causal relationships among complex traits. The large
sample sizes offer the possibility of uncovering associations of modest strength 
that might be undetectable in smaller sample sizes.

# Mendelian Randomization with one Instrument

Mendelian randomization with one instrument uses a single genetic marker 
as the proxy ("instrument" or "instrumental variable") for the "exposure", or 
putative cause, in a causal inference study. When the genetic marker is a single
nucleotide polymorphism on an autosome, the cohort is partitioned into three 
classes, two homozygotes and a heterozygote, based on genotypes for that marker.

## Assumptions of Mendelian Randomization

Three assumptions are central to Mendelian Randomization.

# Mendelian Randomization with multiple instruments

# Horizontal pleiotropy

Horizontal pleiotropy refers to...

# Correlated horizontal pleiotropy

# Applications and Recent Findings

# Future Directions

### 2022-01-26 article-counts-per-year-figure.Rmd
renamed file

---
title: "Citations figure"
author: "Fred Boehm & Xiang Zhou"
date: "1/26/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

We need a bar plot of counts per year for a search with keywords "Mendelian randomization".

```{r, eval = FALSE}
remotes::install_github("akshaynagpal/RgScholar")
```

```{r}
RgScholar::google_Scholar(user_query = "mendelian randomization", year_low = 2021)
```

### 2022-01-27 mr-review.Rmd
added data
added bib file

@@
 author: "Frederick J. Boehm & Xiang Zhou"
-date: "1/26/2022"
+date: "`r lubridate::now()`"
+bibliography: exported.bib
 output: 
+  - bookdown::html_document2
   - bookdown::word_document2
@@
 
-Methods for Mendelian randomization have appeared at a rapid pace since the 
-appearance of the first published GWAS.
+Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
+appearance of the first GWAS.
 
@@
 of discovering even more causal relationships among complex traits. The large
-sample sizes offer the possibility of uncovering associations of modest strength 
+sample sizes in biobank studies offer the possibility of uncovering associations of modest strength 
 that might be undetectable in smaller sample sizes.
@@
 
-# Future Directions
+# Summary and Outlook

### 2022-01-27 mr-review.Rmd
added a citation
started materials on OMR

@@
 Mendelian randomization with one instrument uses a single genetic marker 
-as the proxy ("instrument" or "instrumental variable") for the "exposure", or 
-putative cause, in a causal inference study. When the genetic marker is a single
+as the proxy ("instrument" or "instrumental variable") for the "exposure" in a causal inference study. When the genetic marker is a single
 nucleotide polymorphism on an autosome, the cohort is partitioned into three 
@@
 Three assumptions are central to Mendelian Randomization.
+
+The first, often termed the
 
@@
 
+@yuan_testing_2020
+
 # Applications and Recent Findings
 
+## Omnigenic MR
+
+Recent research on the omnigenic hypothesis, which posits 
+that every SNP's effect on a trait is nonzero, has informed
+MR methods development [@Boyle2017]. @wang_mendelian_nodate 
+recognized the limitations of previous MR methods in the 
+context of the omnigenic hypothesis. To accommodate the possibility
+that a trait is influenced by many genome-wide small effects,
+
 # Summary and Outlook
+
+# References

### 2022-01-27 article-counts-per-year-figure.Rmd
added ggplot2 code

@@
 ---
-title: "Citations figure"
-author: "Fred Boehm & Xiang Zhou"
+title: "Article counts per year figure"
+author: "Fred Boehm"
 date: "1/26/2022"
@@
 
-```{r, eval = FALSE}
-remotes::install_github("akshaynagpal/RgScholar")
+I manually recorded the article counts per year for two searches: 
+1. genome-wide association study and 2. mendelian randomization.
+
+```{r}
+library(dplyr)
+dat <- readxl::read_xlsx(path = "data/publication-counts.xlsx", sheet = "counts")
+dat %>%
+  ggplot2::ggplot(ggplot2::aes(x = year, y = mrcount)) + ggplot2::geom_bar(stat = "identity")
 ```
@@
 ```{r}
-RgScholar::google_Scholar(user_query = "mendelian randomization", year_low = 2021)
+dat %>%
+  ggplot2::ggplot(ggplot2::aes(x = year, y = gwascount)) + ggplot2::geom_bar(stat = "identity")
+
 ```

### 2022-01-27 mr-review.Rmd
many updates

@@
 Mendelian randomization with one instrument uses a single genetic marker 
-as the proxy ("instrument" or "instrumental variable") for the "exposure" in a causal inference study. When the genetic marker is a single
-nucleotide polymorphism on an autosome, the cohort is partitioned into three 
-classes, two homozygotes and a heterozygote, based on genotypes for that marker.
+as the proxy ("instrument" or "instrumental variable") for the "exposure" and asks 
+whether variation in the genotypes at the genetic marker causes variation in the 
+"outcome" variable, which is typically a second complex trait. When the genetic marker 
+is a single nucleotide polymorphism on an autosome, the cohort is partitioned into three 
+classes, two homozygote classes and a single heterozygote class, based on genotypes 
+for that marker. When the three assumptions of mendelian randomization are satisfied, 
+estimation of the causal effect of the exposure on the outcome is possible.
 
@@
 Three assumptions are central to Mendelian Randomization.
+Satisfying them is necessary to infer causal effects in a MR study.
+Careful evaluation of the assumptions is needed.
+Briefly, the first assumption states that the exposure is associated with the outcome.
+The second requires independence, conditional on the exposure and all measured and unmeasured confounders, between the genetic marker and the outcome.
+The last of the three assumptions states that the genetic marker is independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome.
+We next discuss these in more detail and consider their evaluation in MR studies.
 
-The first, often termed the
+The first assumption, about the association between the genetic marker and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].
+
+The second assumption, often termed the exclusion restriction, can also be stated as 
+the need for the genetic marker to affect the outcome only through the exposure. In a 
+causal diagram, this means that the only path from the marker to the outcome is that 
+containing the exposure.
+
+When the three MR assumptions hold, one may use instrumental variable methods, 
+with the genetic marker data, the exposure data, and the outcome data, to estimate the 
+effect of the exposure on the outcome.
 

### 2022-01-28 mr-review.Rmd
added more text

@@
 
+Recognition of complex traits' diverse genetic architectures has fueled development of MR methods with multiple instruments.
+
 # Horizontal pleiotropy
 
-Horizontal pleiotropy refers to...
+Horizontal pleiotropy refers to
 
@@
 
-# Applications and Recent Findings
+# Recent Findings
 
@@
 context of the omnigenic hypothesis. To accommodate the possibility
-that a trait is influenced by many genome-wide small effects,
+that a trait is influenced by many genome-wide, small effects,
+
+## Multiple outcomes in MR
+
+Investigators have found that combining information from multiple  related traits can lead to more informative MR results.
+
+One approach to combining information across traits is to extend the univariate MR statistical methods to their multivariate analogs.
+
+# Applications of MR in GWAS
 

### 2022-01-28 mr-review.Rmd
more refs added
updated table of methods

@@
 
-Horizontal pleiotropy refers to
+Horizontal pleiotropy refers to the presence of multiple pathways from a genetic 
+variant to the outcome. One of these multiple pathways may contain the exposure variable.
+The fact that there are multiple pathways from the genetic variant to the outcome violates
+the MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
+in GWAS. This finding motivated the development of MR-PRESSO and inspired 
+further research into methods to accommodate horizontal pleiotropy in MR studies [@verbanck_detection_2018].
+* What does MR-PRESSO do? What are its limitations?
 

### 2022-01-28 mr-review.Rmd
more text...

@@
 
+We begin our discussion of Mendelian randomization by considering the simplest case, where there is a single genetic marker and a single outcome variable.
 Mendelian randomization with one instrument uses a single genetic marker
@@
 
+The outcome variable can be binary, a count, or a continuous variable.
+
 ## Assumptions of Mendelian Randomization
@@
 Satisfying them is necessary to infer causal effects in a MR study.
-Careful evaluation of the assumptions is needed.
+Careful evaluation of the assumptions is needed in MR studies.
 Briefly, the first assumption states that the exposure is associated with the outcome.
@@
 
-The first assumption, about the association between the genetic marker and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].
+If the first assumption, about the association between the genetic marker and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].
 
@@
 
-When the three MR assumptions hold, one may use instrumental variable methods, 
-with the genetic marker data, the exposure data, and the outcome data, to estimate the 
-effect of the exposure on the outcome.
+The third assumption, sometimes called the ____, ensures that ___.
+
+When the three MR assumptions hold, one may use instrumental variable statistical methods, 
+with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the causal effect of the exposure on the outcome.
+
+## Assumption violations
+
+There are several ways that real data may violate the three MR assumptions. These include ___, ____, and ___.
+
+Recognition that the assumptions may be violated has inspired methods advances that 
+enable assumption relaxations under certain conditions.
+
+## Statistical models for MR with one instrument and one outcome
+
+## Statistical methods for MR with one instrument and one outcome
+
+The approaches to MR can be classified into two categories: two-stage methods and 
+likelihood-based methods. Both two-stage strategies and likelihood-based strategies 
+make parametric assumptions for the identification of causal effects [@burgess_review_2017].
+
+Two-stage methods use two regressions. In the first stage, the exposure is regressed on
+the instrument to obtain fitted values. In the second stage, the outcome is regressed on the fitted values for the 
+exposure, which result from the first regression. Note that the uncertainty in the fitted values of 
+the first regression is not considered when performing the second regression.
 

### 2022-01-31 mr-review.Rmd
reworded a sentence
added csl

@@
 bibliography: exported.bib
+csl: elsevier-with-titles.csl
 output: 
@@
 Great interest has developed in inferring causal relationships between complex 
-traits in observational human genetics studies.
+traits in observational human genetics studies. Throughout the manuscript, we 
+use the term "complex trait" to refer to a trait that doesn't appear to be 
+inherited in a Mendelian fashion. Typically, traits such as gene expression levels
+are inherited in a polygenic manner, with many genes influencing the trait.
 
 Genome-wide association studies (GWAS) present unique opportunities for 
-discovery of causal relationships via mendelian randomization.
+discovery of causal relationships via mendelian randomization. A genome-wide association study 
+interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
+SNP is a genetic marker defined by nucleotide base identity at a single genomic position.
 
@@
 
-Now, GWAS and Mendelian randomization for Biobank-scale data offer the possibility
+Now, GWAS and Mendelian randomization for biobank-scale data offer the possibility
 of discovering even more causal relationships among complex traits. The large
@@
 
-Three assumptions are central to Mendelian Randomization.
-Satisfying them is necessary to infer causal effects in a MR study.
-Careful evaluation of the assumptions is needed in MR studies.
-Briefly, the first assumption states that the exposure is associated with the outcome.
-The second requires independence, conditional on the exposure and all measured and unmeasured confounders, between the genetic marker and the outcome.
-The last of the three assumptions states that the genetic marker is independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome.
-We next discuss these in more detail and consider their evaluation in MR studies.
+Three assumptions are central to Mendelian Randomization. Satisfying them is necessary to infer 
+causal effects in a MR study. Careful evaluation of the assumptions is needed in MR studies. Briefly, 
+the first assumption states that the exposure is associated with the outcome. The second requires 
+independence, conditional on the exposure and all measured and unmeasured confounders, between the 
+genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
+independent of all (measured and unmeasured) confounders of the relationship between exposure and 
+outcome. We next discuss these in more detail and consider their evaluation in MR studies.
 
-If the first assumption, about the association between the genetic marker and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].
+If the first assumption, about the association between the genetic marker and the exposure, holds, 
+but this association is weak, then an amplification of biases, such as those due to violations of 
+assumptions 2 and 3, may result [@vanderweele_methodological_2014].
 
@@
 When the three MR assumptions hold, one may use instrumental variable statistical methods, 
-with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the causal effect of the exposure on the outcome.
+with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
+causal effect of the exposure on the outcome.
 
@@
 
-Investigators have found that combining information from multiple  related traits can lead to more informative MR results.
+Investigators have found that combining information from multiple related traits can lead to richer MR results.
 

### 2022-01-31 mr-review-outline.Rmd
made a separate file for outlining

--- 2022-01-28 mr-review.Rmd
@@
 title: "Mendelian randomization for causal inference in Genome-wide association studies"
-author: "Frederick J. Boehm & Xiang Zhou"
+author: "Frederick J. Boehm"
 date: "`r lubridate::now()`"
 bibliography: exported.bib
+csl: elsevier-with-titles.csl
 output: 
@@
 
-Great interest has developed in inferring causal relationships between complex 
-traits in observational human genetics studies.
-
-Genome-wide association studies (GWAS) present unique opportunities for 
-discovery of causal relationships via mendelian randomization.
-
-Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
-appearance of the first GWAS.
-
-Now, GWAS and Mendelian randomization for Biobank-scale data offer the possibility
-of discovering even more causal relationships among complex traits. The large
-sample sizes in biobank studies offer the possibility of uncovering associations of modest strength 
-that might be undetectable in smaller sample sizes.
-
 # Mendelian Randomization with one Instrument
-
-Mendelian randomization with one instrument uses a single genetic marker 
-as the proxy ("instrument" or "instrumental variable") for the "exposure" and asks 
-whether variation in the genotypes at the genetic marker causes variation in the 
-"outcome" variable, which is typically a second complex trait. When the genetic marker 
-is a single nucleotide polymorphism on an autosome, the cohort is partitioned into three 
-classes, two homozygote classes and a single heterozygote class, based on genotypes 
-for that marker. When the three assumptions of mendelian randomization are satisfied, 
-estimation of the causal effect of the exposure on the outcome is possible.
 
@@
 
-Three assumptions are central to Mendelian Randomization.
-Satisfying them is necessary to infer causal effects in a MR study.
-Careful evaluation of the assumptions is needed.
-Briefly, the first assumption states that the exposure is associated with the outcome.
-The second requires independence, conditional on the exposure and all measured and unmeasured confounders, between the genetic marker and the outcome.
-The last of the three assumptions states that the genetic marker is independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome.
-We next discuss these in more detail and consider their evaluation in MR studies.
+## Assumption violations
 
-The first assumption, about the association between the genetic marker and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].
+## Statistical models for MR with one instrument and one outcome
 
-The second assumption, often termed the exclusion restriction, can also be stated as 
-the need for the genetic marker to affect the outcome only through the exposure. In a 
-causal diagram, this means that the only path from the marker to the outcome is that 
-containing the exposure.
-
-When the three MR assumptions hold, one may use instrumental variable methods, 
-with the genetic marker data, the exposure data, and the outcome data, to estimate the 
-effect of the exposure on the outcome.
+## Statistical methods for MR with one instrument and one outcome
 
@@
 
-Recognition of complex traits' diverse genetic architectures has fueled development of MR methods with multiple instruments.
-
 # Horizontal pleiotropy
 
-Horizontal pleiotropy refers to the presence of multiple pathways from a genetic 
-variant to the outcome. One of these multiple pathways may contain the exposure variable.
-The fact that there are multiple pathways from the genetic variant to the outcome violates
-the MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
-in GWAS. This finding motivated the development of MR-PRESSO and inspired 
-further research into methods to accommodate horizontal pleiotropy in MR studies [@verbanck_detection_2018].
-* What does MR-PRESSO do? What are its limitations?
-
 # Correlated horizontal pleiotropy
-
-@yuan_testing_2020
 
@@
 
-Recent research on the omnigenic hypothesis, which posits 
-that every SNP's effect on a trait is nonzero, has informed
-MR methods development [@Boyle2017]. @wang_mendelian_nodate 
-recognized the limitations of previous MR methods in the 
-context of the omnigenic hypothesis. To accommodate the possibility
-that a trait is influenced by many genome-wide, small effects,
-
 ## Multiple outcomes in MR
-
-Investigators have found that combining information from multiple  related traits can lead to more informative MR results.
-
-One approach to combining information across traits is to extend the univariate MR statistical methods to their multivariate analogs.
 

### 2022-01-31 mr-review.Rmd
made a separate file for outlining

@@
 use the term "complex trait" to refer to a trait that doesn't appear to be 
-inherited in a Mendelian fashion. Typically, traits such as gene expression levels
+inherited in a Mendelian fashion. Typically, such complex traits such as gene expression levels
 are inherited in a polygenic manner, with many genes influencing the trait.
@@
 Mendelian randomization with one instrument uses a single genetic marker
-as the proxy ("instrument" or "instrumental variable") for the "exposure" and asks 
+as the proxy ("instrument" or "instrumental variable") for the "exposure", a complex trait, and asks 
 whether variation in the genotypes at the genetic marker causes variation in the 
@@
 
-Recognition of complex traits' diverse genetic architectures has fueled development of MR methods with multiple instruments.
+Recognition of complex traits' diverse genetic architectures has fueled development of MR methods with multiple instruments [@Boyle2017].
 

### 2022-01-31 mr-review-outline.Rmd
switched outline to md output format

@@
 output: 
+  - bookdown::github_document2
   - bookdown::html_document2

### 2022-01-31 article-counts-per-year-figure.Rmd
magrittr to replace dplyr

@@
 ```{r}
-library(dplyr)
+library(magrittr)
 dat <- readxl::read_xlsx(path = "data/publication-counts.xlsx", sheet = "counts")

### 2022-02-01 mr-review-outline.Rmd
minor updates to Rmd outline
added topic sentences to outline

@@
 
+Great interest has developed in inferring causal relationships between complex 
+traits in observational human genetics studies.
+
+Genome-wide association studies (GWAS) present unique opportunities for 
+discovery of causal relationships via mendelian randomization.
+
+Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
+appearance of the first GWAS.
+
+Now, GWAS and Mendelian randomization for biobank-scale data offer the possibility
+of discovering even more causal relationships among complex traits.
+
 # Mendelian Randomization with one Instrument
+
+We begin our discussion of Mendelian randomization by considering the simplest case, where there is a single genetic marker and a single outcome variable.
+
+A MR study can be performed with one sample or two samples.
 
@@
 
+Three assumptions are central to Mendelian Randomization.
+
 ## Assumption violations
+
+Several real world settings may give rise to violations of one or more of the three assumptions.
 
@@
 
+Statistical models for MR with one instrument can be formulated as two linear regressions.
+
 ## Statistical methods for MR with one instrument and one outcome
 
+Estimation and testing for unknown parameters proceeds in one of several methods.
+
+## Sensitivity analysis for MR with one instrument and one outcome
+
 # Mendelian Randomization with multiple instruments
+
+Recognizing that many
 
@@
 
+1. greater sample sizes and richer, molecular phenotyping will continue to enable 
+causal inferences in observational human genetics studies
+
+1. ??Experimental validation in animal models?? PerturbSeq?? CRISPR manipulation of putative causal genes??
+
+1. Robustness to assumptions violations
+
 # References

### 2022-02-01 mr-review-outline.Rmd
minor text changes

@@
 
-Estimation and testing for unknown parameters proceeds in one of several methods.
+Estimation and testing for unknown parameters proceeds in one of several approaches.
 
@@
 
+Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify the three MR assumptions.
+
 # Mendelian Randomization with multiple instruments
 
-Recognizing that many
+Recognizing that many genetic markers collectively may affect a trait, investigators 
+adapted single-instrument MR methods to accommodate multiple genetic markers.
 
 # Horizontal pleiotropy
+
+Linkage disequilibrium, which refers to genotype correlations for genetic markers 
+that are near each other, manifests as horizontal pleiotropy in MR studies.
 
@@
 ## Multiple outcomes in MR
+
+## MR with summary statistics (Morrison et al.)
 

### 2022-02-02 article-counts-per-year-figure.Rmd
rendered Rmd to get figs

--- 2022-01-27 article-counts-per-year-figure.Rmd
@@
   ggplot2::ggplot(ggplot2::aes(x = year, y = mrcount)) + ggplot2::geom_bar(stat = "identity")
+ggplot2::ggsave("figures/mr-counts.png")
 ```
@@
   ggplot2::ggplot(ggplot2::aes(x = year, y = gwascount)) + ggplot2::geom_bar(stat = "identity")
+ggplot2::ggsave("figures/gwas-counts.png")
 
 ```
+
+```{r}
+dat %>%
+  tidyr::pivot_longer(!year) %>%
+  ggplot2::ggplot(ggplot2::aes(x = year, y = value, colour = name)) + ggplot2::geom_point()
+ggplot2::ggsave("figures/mr-gwas-counts-scatter.png")
+```

### 2022-02-02 mr-review.Rmd
updated ms

@@
 Great interest has developed in inferring causal relationships between complex 
-traits in observational human genetics studies. Throughout the manuscript, we 
+traits in observational human genetics studies. Discovery of such relationships is crucial to 
+enhancing our understanding of the biology of health and disease. Two-sample 
+Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
+genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
+inferring causal effects of an exposure variable on an outcome variable. They are termed 
+"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
+two non-overlapping cohorts of subjects. Importantly, the SNPs are measured in both sets of subjects.
+
+Throughout the manuscript, we 
 use the term "complex trait" to refer to a trait that doesn't appear to be 
-inherited in a Mendelian fashion. Typically, such complex traits such as gene expression levels
+inherited in a Mendelian fashion. Typically, such complex traits, for example, gene expression levels,
 are inherited in a polygenic manner, with many genes influencing the trait.
@@
 Genome-wide association studies (GWAS) present unique opportunities for 
-discovery of causal relationships via mendelian randomization. A genome-wide association study 
+discovery of causal relationships via MR. A GWAS
 interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
@@
 
+```{r, article-counts, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
+knitr::include_graphics("figures/mr-gwas-counts-scatter.png")
+```
+
 Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
-appearance of the first GWAS.
+appearance of the first GWAS (Figure \@ref(fig:article-counts)).
 
@@
 that might be undetectable in smaller sample sizes.
+
+~~Investigating causal relationship among complex traits and identifying causal risk
+factors are an important first step towards understanding the biology of diseases. A
+common statistical tool for performing such causal inference in observational studies is
+Mendelian randomization (MR). MR is a form of instrumental variable analysis that
+uses SNPs to serve as instruments for inferring the causal effect of an exposure variable
+on an outcome variable(1). MR requires only summary statistics from genome-wide
+association studies (GWASs) and is often performed in a two-sample study setting
+where the exposure variable and the outcome variable are measured in two separate
+studies(2). With the abundant availability of GWAS summary statistics, numerous MR
+analyses are being carried out, identifying important causal risk factors for various
+common diseases. These MR studies are facilitated by many recently developed MR
+methods that include the inverse variance weighted (IVW) method, MR-Egger(3),
+median-based regression(4), BWMR(5), RAPS(6), MRMix(7), CAUSE(8), to name a
+few. Different MR methods differ in their modeling assumptions and inference
+algorithms, but the majority of them encounter two important modeling and algorithmic
+challenges that have so far limited the effectiveness of MR analysis.
+First, almost all existing MR methods rely on a pre-selected set of independent
+SNPs to serve as instruments for MR analysis. The instruments are selected to be
+independent from each other to ensure the validity of the statistical inference framework
+used in many common MR methods such as IVW. The independent SNPs are often
+selected through linkage disequilibrium (LD) clumping, a procedure that first ranks
+SNPs based on their marginal association evidence with the exposure variable and then
+retains SNPs that are not in high LD with the SNPs on top of the ranking list. Using LD
+clumping to select SNPs may be suboptimal, however, as the selected SNPs may only
+epresent tagging SNPs that are in LD with the causal SNPs rather than the causal ones
+themselves. Using tagging SNPs instead of the causal ones to serve as instruments can
+reduce the power of MR analysis. In addition, perhaps more importantly, selecting
+independent SNPs for MR analysis may not be ideal either, as complex traits can be
+influenced by multiple causal SNPs residing in the same local region that are in
+potential LD with each other. Consequently, selecting independent SNPs may only
+capture a small proportion of the phenotypic variance in the exposure variable, again
+leading to a loss of power in the subsequent MR analysis (1, 2, 9, 10). Indeed, in the
+parallel research field of transcriptome-wide association studies, it has been well
+documented that incorporating correlated SNPs can substantially improve analysis
+power than using independent SNPs only(11-14). Therefore, incorporating correlated
+SNPs and developing effective approaches to select instruments among them are
+important to fully captivate the potential of MR.
+Second, only a limited number of MR methods model horizontal pleiotropy and
+even fewer can effectively control for it during MR analysis(15). Horizontal pleiotropy
+occurs when the SNP instruments exhibit effects on the outcome through pathways
+other than the exposure. Horizontal pleiotropy has been widely observed in complex
+trait analysis(13, 15) and often comes in two distinct types. The first type of horizontal
+pleiotropy arises through paths independent of the exposure, with the resulting
+horizontal pleotropic effects being independent of the SNP effects on the exposure. The
+second type of horizontal pleiotropy arises through unobserved exposure-outcome
+confounders and induces correlation between the horizontal pleotropic effects and the
+SNP effects on the exposure. The presence of either type of horizontal pleiotropy
+violates standard MR modeling assumptions and can lead to biased causal effect
+estimates and increased false discoveries. Early MR analyses control for horizontal~~
 
@@
 
-## Assumption violations
-
-There are several ways that real data may violate the three MR assumptions. These include ___, ____, and ___.
-
-Recognition that the assumptions may be violated has inspired methods advances that 
-enable assumption relaxations under certain conditions.
-
 ## Statistical models for MR with one instrument and one outcome
@@
 the first regression is not considered when performing the second regression.
+
+## Assumption violations
+
+There are several ways that real data may violate the three MR assumptions. These include .
+
+Recognition that the assumptions may be violated has inspired methods advances that 
+enable assumption relaxations under certain conditions.
+
+## Sensitivity analysis for MR with one instrument and one outcome
+
+Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
+the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
+into how the results might differ were we able to verify the MR assumptions.
 

### 2022-02-03 mr-review.Rmd
deleted word
added refs to intro

@@
 Great interest has developed in inferring causal relationships between complex 
-traits in observational human genetics studies. Discovery of such relationships is crucial to 
+traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
 enhancing our understanding of the biology of health and disease. Two-sample 
@@
 genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
-inferring causal effects of an exposure variable on an outcome variable. They are termed 
+inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].
+They are termed
 "two-sample" MR studies because the exposure variable and the outcome variable are measured on 
-two non-overlapping cohorts of subjects. Importantly, the SNPs are measured in both sets of subjects.
+two non-overlapping cohorts of subjects. Importantly, the SNP instruments are measured in both sets of subjects.
 
-Throughout the manuscript, we 
-use the term "complex trait" to refer to a trait that doesn't appear to be 
-inherited in a Mendelian fashion. Typically, such complex traits, for example, gene expression levels,
-are inherited in a polygenic manner, with many genes influencing the trait.
+Recent developments have enabled performance of two-sample MR studies with only 
+GWAS summary statistics [@burgess2015using].
 

### 2022-02-03 mr-review.Rmd
added to 1-instrument, 1-outcome methods

--- 2022-01-31 mr-review.Rmd
@@
 Great interest has developed in inferring causal relationships between complex 
-traits in observational human genetics studies. Throughout the manuscript, we 
-use the term "complex trait" to refer to a trait that doesn't appear to be 
-inherited in a Mendelian fashion. Typically, such complex traits such as gene expression levels
-are inherited in a polygenic manner, with many genes influencing the trait.
+traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
+enhancing our understanding of the biology of health and disease. Two-sample 
+Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
+genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
+inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].
+They are termed
+"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
+two non-overlapping cohorts of subjects. Importantly, the SNP instruments are measured in both sets of subjects.
 
 Genome-wide association studies (GWAS) present unique opportunities for 
-discovery of causal relationships via mendelian randomization. A genome-wide association study 
+discovery of causal relationships via MR. A GWAS
 interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
@@
 
+```{r, article-counts, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
+knitr::include_graphics("figures/mr-gwas-counts-scatter.png")
+```
+
 Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
-appearance of the first GWAS.
+appearance of the first GWAS (Figure \@ref(fig:article-counts)).
+
+Recent developments have enabled execution of two-sample MR studies with only 
+GWAS summary statistics [@burgess2015using].
+These methods are especially pertinent now that there is an abundance of freely available GWAS summary data [@buniello2019nhgri].
+GWAS summary statistics typically consist of an identifier and a p-value for every SNP in the data set [@buniello2019nhgri].
 
@@
 
-We begin our discussion of Mendelian randomization by considering the simplest case, where there is a single genetic marker and a single outcome variable.
+We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where there is a single genetic marker and a single outcome variable.
 Mendelian randomization with one instrument uses a single genetic marker
@@
 "outcome" variable, which is typically a second complex trait. When the genetic marker 
-is a single nucleotide polymorphism on an autosome, the cohort is partitioned into three 
+is a SNP on an autosome, the cohort is partitioned into three 
 classes, two homozygote classes and a single heterozygote class, based on genotypes 
@@
 Three assumptions are central to Mendelian Randomization. Satisfying them is necessary to infer 
-causal effects in a MR study. Careful evaluation of the assumptions is needed in MR studies. Briefly, 
+causal effects in an observational MR study. Careful evaluation of the assumptions is needed. Briefly, 
 the first assumption states that the exposure is associated with the outcome. The second requires 
@@
 
-The second assumption, often termed the exclusion restriction, can also be stated as 
+The second assumption, often termed the exclusion restriction assumption, can also be stated as 
 the need for the genetic marker to affect the outcome only through the exposure. In a 
@@
 
-The third assumption, sometimes called the ____, ensures that ___.
+The third assumption, sometimes called the , ensures that .
 
@@
 
-## Assumption violations
+## Statistical models for MR with one instrument and one outcome
 
-There are several ways that real data may violate the three MR assumptions. These include ___, ____, and ___.
-
-Recognition that the assumptions may be violated has inspired methods advances that 
-enable assumption relaxations under certain conditions.
-
-## Statistical models for MR with one instrument and one outcome
+A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
+outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
+First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes). The resulting 
+fitted values for the exposure variables are the independent variable in the second linear regression, 
+where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
+coefficient obtained from the second regression analysis.
 

### 2022-02-03 mr-review.Rmd
reorganized a paragraph

@@
 of discovering even more causal relationships among complex traits. The large
-sample sizes in biobank studies offer the possibility of uncovering associations of modest strength 
-that might be undetectable in smaller sample sizes.
+sample sizes in biobank studies, often greater than 100,000 subjects, offer the possibility of uncovering associations of modest strength 
+that might be undetectable in smaller studies.
 
@@
 
-## Statistical models for MR with one instrument and one outcome
-
-A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
-outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
-First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes). The resulting 
-fitted values for the exposure variables are the independent variable in the second linear regression, 
-where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
-coefficient obtained from the second regression analysis.
-
-## Statistical methods for MR with one instrument and one outcome
+## Statistical models and methods for MR with one instrument and one outcome
 
@@
 the first regression is not considered when performing the second regression.
+
+A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
+outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
+First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes). The resulting 
+fitted values for the exposure variables are the independent variable in the second linear regression, 
+where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
+coefficient obtained from the second regression analysis.
 

### 2022-02-03 mr-review.Rmd
added ukb ref

@@
 output: 
+  - bookdown::pdf_document2
   - bookdown::html_document2
@@
 
-Now, GWAS and Mendelian randomization for biobank-scale data offer the possibility
+Modern GWAS for biobank-scale data offer the possibility
 of discovering even more causal relationships among complex traits. The large
-sample sizes in biobank studies, often greater than 100,000 subjects, offer the possibility of uncovering associations of modest strength 
-that might be undetectable in smaller studies.
+sample sizes in biobank studies, often greater than 100,000 
+subjects, offer the possibility of uncovering associations of modest strength 
+that might be undetectable in smaller studies. For example, the UK 
+Biobank Study involves more than 500,000 subjects. Each UKB subject's 
+data consists of millions of SNP genotypes and measurements for 
+thousands of traits [@bycroft_uk_2018].
 

### 2022-02-04 mr-review.Rmd
abbreviated MR

@@
 
-Three assumptions are central to Mendelian Randomization. Satisfying them is necessary to infer 
+Three assumptions are central to MR. Satisfying them is necessary to infer 
 causal effects in an observational MR study. Careful evaluation of the assumptions is needed. Briefly, 
@@
 
-Recognition of complex traits' diverse genetic architectures has fueled development of MR methods with multiple instruments [@Boyle2017].
+Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR methods with multiple instruments [@Boyle2017].
 
@@
 
-Investigators have found that combining information from multiple related traits can lead to richer MR results.
+Investigators have found that combining information from multiple related 
+traits can lead to richer MR results.
 
-One approach to combining information across traits is to extend the univariate MR statistical methods to their multivariate analogs.
+One approach to combining information across traits is to extend the 
+univariate MR statistical methods to their multivariate analogs.
 

### 2022-02-04 mr-review.Rmd
added 1st dag

@@
 These methods are especially pertinent now that there is an abundance of freely available GWAS summary data [@buniello2019nhgri].
-GWAS summary statistics typically consist of an identifier and a p-value for every SNP in the data set [@buniello2019nhgri].
+GWAS summary statistics typically consist of an identifier and a p-value for every SNP in the data set, along with the sample size [@buniello2019nhgri].
 
@@
 
-We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where there is a single genetic marker and a single outcome variable.
-Mendelian randomization with one instrument uses a single genetic marker
+We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where 
+there is a single genetic marker and a single outcome variable. Mendelian randomization with one 
+instrument uses a single genetic marker 
 as the proxy ("instrument" or "instrumental variable") for the "exposure", a complex trait, and asks 
@@
 
-There are several ways that real data may violate the three MR assumptions. These include .
+There are several ways that experimental data may violate one or more of the three MR assumptions.
+Let's first consider the exclusion restriction assumption. It states that the genetic marker 
+must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
+a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
+if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
+possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
+affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].
 
-Recognition that the assumptions may be violated has inspired methods advances that 
-enable assumption relaxations under certain conditions.
+```{r, dag1, fig.cap = "Scenario where the genetic variant affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}
+g1 <- dagitty::dagitty( "dag {
+	G -> W -> X -> Y
+	W -> Y
+}")
+plot(dagitty::graphLayout(g1))
+```
+
+Recognition of the many ways that the MR assumptions may be violated has both inspired methods advances that 
+enable assumption relaxations under certain conditions and motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
 

### 2022-02-04 mr-review.Rmd
added eqs

@@
 Three assumptions are central to MR. Satisfying them is necessary to infer 
-causal effects in an observational MR study. Careful evaluation of the assumptions is needed. Briefly, 
-the first assumption states that the exposure is associated with the outcome. The second requires 
+causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
+The first assumption states that the exposure is associated with the outcome. The second requires 
 independence, conditional on the exposure and all measured and unmeasured confounders, between the 
@@
 independent of all (measured and unmeasured) confounders of the relationship between exposure and 
-outcome. We next discuss these in more detail and consider their evaluation in MR studies.
+outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].
 
@@
 
-The third assumption, sometimes called the , ensures that .
+The third assumption ensures that the genetic marker is
+independent of all confounders of the exposure-outcome relationship.
 
@@
 with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
-causal effect of the exposure on the outcome.
+causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
+
+To understand the origin of the term "Mendelian randomization", it is helpful to 
+recall Mendel's inheritance laws. Under the assumption that alleles segregate 
+randomly from parent to offspring, the offspring genotypes are unlikely to be 
+associated with confounders of the exposure-outcome relationship. Additionally,
+reverse causation is ruled out by the observation that germ-line genotypes are 
+fixed at conception and, thus, precede other observed variables.
 
@@
 
-Two-stage methods use two regressions. In the first stage, the exposure is regressed on
-the instrument to obtain fitted values. In the second stage, the outcome is regressed on the fitted values for the 
-exposure, which result from the first regression. Note that the uncertainty in the fitted values of 
-the first regression is not considered when performing the second regression.
-
 A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
 outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
-First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes). The resulting 
-fitted values for the exposure variables are the independent variable in the second linear regression, 
+First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
+fitted values for the exposure variable are the independent variable in the second linear regression, 
 where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
 coefficient obtained from the second regression analysis.
+
+\begin{equation}
+  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
+  (\#eq:lm1)
+\end{equation}
+
+where the $\epsilon_i$ are assumed to be iid normal random variables with mean zero.
+
+\begin{equation}
+  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
+  (\#eq:lm2)
+\end{equation}
+
+where the $\tau_i$ are assumed to be iid normal random variables with mean zero and $\hat x_i$ denotes the fitted value for $x_i$ obtained from Equation \@ref(eq:lm1).
+
+Note that the uncertainty in the fitted values from 
+the first regression is not considered when performing the second regression.
+For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is incorrect in two-stage calculations. This observation led researchers to develop likelihood-based MR methods.
 
@@
 
-```{r, dag1, fig.cap = "Scenario where the genetic variant affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}
+```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}
 g1 <- dagitty::dagitty( "dag {
@@
 
-Recognition of the many ways that the MR assumptions may be violated has both inspired methods advances that 
-enable assumption relaxations under certain conditions and motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
+Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that 
+enable assumption relaxations under certain conditions and has motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
 

### 2022-02-04 mr-review.Rmd
added more ref

@@
 where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
-coefficient obtained from the second regression analysis.
+coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).
 
@@
 Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that 
-enable assumption relaxations under certain conditions and has motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
+enable assumption relaxations and has motivated use of sensitivity analysis to quantify the impact of 
+possible assumption violations.
 
@@
 
-Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR methods with multiple instruments [@Boyle2017].
+Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR methods with multiple instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019].
 

### 2022-02-05 mr-review.Rmd
added section header
wrote about sensitivity analysis

@@
 
+@burgess_sensitivity_2017 divided the suite of methods for sensitivity analysis 
+into two categories: 1) those that support the validity of causal inferences in a MR 
+study and 2) those that question that validity. Because not all confounding variables are known or measured, assumptions 2 and 3 are not fully verifiable 
+in MR studies. Recognition of this fact has inspired the development of 
+sensitivity analysis tools. While we can't assess whether a SNP instrument is 
+associated with every confounder of the exposure-outcome association, it is 
+possible to examine the associations of the SNP instrument with the measured 
+covariates. While absence of such associations doesn't guarantee satisfaction 
+of the assumption, presence of SNP-covariate associations must be investigated
+carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].
+
 # Mendelian Randomization with multiple instruments
@@
 Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR methods with multiple instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019].
+
+## Statistical models and methods for MR with multiple instruments and one outcome
 

### 2022-02-05 mr-review.Rmd
carriage return
minor addns

@@
 
-Methods for Mendelian randomization have appeared at a rapid and accelerating pace since the 
-appearance of the first GWAS (Figure \@ref(fig:article-counts)).
+Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
+appearance of the early GWAS (Figure \@ref(fig:article-counts)).
+Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted median estimation [@bowden_consistent_2016], Bayesian weighted MR [@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix [@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID [@yuan_likelihood_2021] are all recent methods developments.
+These methods differ in their approaches to two challenges:
+
+1. instrument selection (from among the available SNPs)  
+1. modeling and controlling for horizontal pleiotropy
+
+Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
+be statistically independent. This restriction to 
+independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
+selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
+are ranked by the strength of their marginal association with the exposure 
+variable. The LD clumping procedure then selects SNPs that are not in high LD 
+with the top-ranking SNPs.
+
+LD clumping for choosing instrument SNPs can be problematic for several 
+reasons. First, the chosen SNPs may not be causal for the outcome. This can 
+happen, for instance, by choosing a noncausal SNP that is in high LD with a 
+causal SNP.
+
+@verbanck_detection_2018
 
 Recent developments have enabled execution of two-sample MR studies with only 
-GWAS summary statistics [@burgess2015using].
-These methods are especially pertinent now that there is an abundance of freely available GWAS summary data [@buniello2019nhgri].
-GWAS summary statistics typically consist of an identifier and a p-value for every SNP in the data set, along with the sample size [@buniello2019nhgri].
+GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is 
+an abundance of freely available GWAS summary data [@buniello2019nhgri]. GWAS summary statistics 
+typically consist of an identifier and a p-value for every SNP in the data set, along with the sample 
+size [@buniello2019nhgri].
 
@@
 into two categories: 1) those that support the validity of causal inferences in a MR 
-study and 2) those that question that validity. Because not all confounding variables are known or measured, assumptions 2 and 3 are not fully verifiable 
+study and 2) those that question that validity. Because not all confounding 
+variables are known or measured, assumptions 2 and 3 are not fully verifiable 
 in MR studies. Recognition of this fact has inspired the development of 

### 2022-02-06 mr-review.Rmd
added text around multi-SNP methods

@@
 the first regression is not considered when performing the second regression.
-For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is incorrect in two-stage calculations. This observation led researchers to develop likelihood-based MR methods.
+For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is 
+incorrect in two-stage calculations. This observation led researchers to develop 
+likelihood-based MR methods.
 
@@
 There are several ways that experimental data may violate one or more of the three MR assumptions.
-Let's first consider the exclusion restriction assumption. It states that the genetic marker 
+Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
 must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
@@
 the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
-into how the results might differ were we able to verify the MR assumptions.
+into how the results might differ if the assumptions are violated.
 
@@
 
-Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR methods with multiple instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019].
+Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
+methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019].
 
 ## Statistical models and methods for MR with multiple instruments and one outcome
+
+In the case of continuous exposures and outcomes, we extend the methods and models for 
+single-SNP MR to accommodate multiple SNPs. For two stage analyses, we perform 
+a sequence of two linear regressions, as in the single-SNP MR models (Equations \@ref(eq:lm3) and \@ref(eq:lm4)). 
+For an analysis
+with $S$ SNP instruments, one performs two stage analysis by first regressing the 
+exposure $n$ vector $X$ on the $n$ by $S$ matrix of SNP instruments. Taking the fitted exposure values 
+from this regression, the second regression involves regressing the outcome variable $Y$ on 
+the fitted values.
+
+\begin{equation}
+  X  = \alpha Z + \epsilon
+  (\#eq:lm3)
+\end{equation}
+
+where $\epsilon$ is assumed to be a vector of iid normal random variables with mean zero.
+
+\begin{equation}
+  Y = \beta \hat X + \tau
+  (\#eq:lm4)
+\end{equation}
 
@@
 in GWAS. This finding motivated the development of MR-PRESSO and inspired 
-further research into methods to accommodate horizontal pleiotropy in MR studies [@verbanck_detection_2018].
-* What does MR-PRESSO do? What are its limitations?
+further research into methods to model and to account for horizontal pleiotropy [@verbanck_detection_2018].
 

### 2022-02-07 mr-review.Rmd
added liml text

@@
 
-The outcome variable can be binary, a count, or a continuous variable.
+The outcome variable can be binary, a count, or a continuous variable. For brevity and simplicity, 
+we focus on continuous outcomes below.
 
@@
 
-Three assumptions are central to MR. Satisfying them is necessary to infer 
+Three assumptions are central to MR. Satisfying them enables inference of 
 causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
@@
 but this association is weak, then an amplification of biases, such as those due to violations of 
-assumptions 2 and 3, may result [@vanderweele_methodological_2014].
+assumptions 2 and 3, may result [@vanderweele_methodological_2014]. @burgess_avoiding_2011, in analyses 
+of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
+the strength of association between the SNP and exposure.
 
@@
 likelihood-based MR methods.
+
+Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
+estimates with their many desirable properties. Limited information maximum likelihood from econometrics 
+is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020].
+
+Limited information maximum likelihood model equations are shown below (Equations \@ref(eq:lm5) and \@ref(eq:lm6)).
+
+\begin{equation}
+  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
+  (\#eq:lm5)
+\end{equation}
+
+\begin{equation}
+  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
+  (\#eq:lm6)
+\end{equation}
+
+They differ from the two-stage model because the two error terms, $\epsilon_i$ and $\tau_i$, have a bivariate normal
+distribution.
+
+CoMM, along with its extension for summary statistics, CoMM-S2, uses a collaborative mixed model to 
+relate exposure and outcome variables to the SNP genotypes [@yang_comm-s2_2020]. Model fitting then 
+proceeds via a parameter-expanded EM algorithm [@yang_comm_2019;@liu_parameter_1998].
+
+We discuss PMR-Egger below in the context of methods that deal with horizontal pleiotropy.
 
@@
 
+probabilistic Mendelian randomization (PMR-Egger) for two-sample MR with correlated SNPs models [@yuan_testing_2020]
+
 # Correlated horizontal pleiotropy

### 2022-02-07 mr-review.Rmd
text following MRAID paper

@@
 
-We discuss PMR-Egger below in the context of methods that deal with horizontal pleiotropy.
+We discuss PMR-Egger below in the context of methods that model horizontal pleiotropy.
 
@@
 
+Choosing a subset of the available SNPs to serve as instruments is one consideration that arises 
+in MR studies. The solution to this task differs among published methods. Early methods chose a 
+set of independent SNPs. However, this approach can be problematic when there are multiple causal 
+SNPs in LD with each other. In this setting, discarding some of the causal SNPs may only capture a small
+proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
+this is the finding that including multiple correlated SNPs in transcriptome-wide association
+studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].
+
 ## Statistical models and methods for MR with multiple instruments and one outcome
 
-In the case of continuous exposures and outcomes, we extend the methods and models for 
-single-SNP MR to accommodate multiple SNPs. For two stage analyses, we perform 
+In the case of continuous exposures and outcomes, researchers have taken several approaches to extend the methods and models for 
+single-SNP MR to accommodate multiple SNPs. For two stage analyses, one may perform 
 a sequence of two linear regressions, as in the single-SNP MR models (Equations \@ref(eq:lm3) and \@ref(eq:lm4)). 
@@
 The fact that there are multiple pathways from the genetic variant to the outcome violates
-the MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
+standard MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
 in GWAS. This finding motivated the development of MR-PRESSO and inspired 
 further research into methods to model and to account for horizontal pleiotropy [@verbanck_detection_2018].
+
+Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
+instrumental SNPs that might be associated with the outcome [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
+More recent methods have attempted to model horizontal pleiotropy [@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021].
+CAUSE [@morrison_mendelian_2020] mdoels two varieties of horizontal pleiotropy:
 

### 2022-02-08 mr-review.Rmd
added text about correlated horizontal pleiotropy

--- 2022-02-06 mr-review.Rmd
@@
 Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
-methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019].
+methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015].
 
@@
 exposure $n$ vector $X$ on the $n$ by $S$ matrix of SNP instruments. Taking the fitted exposure values 
-from this regression, the second regression involves regressing the outcome variable $Y$ on 
-the fitted values.
+from this regression, the second stage involves regressing the outcome variable $Y$ on 
+the fitted values, $\hat X$.
 
@@
 
+Researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
+exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
+SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
+in the presence of unobserved exposure-outcome confounding. It induces correlation between 
+horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
+violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
+discoveries [@yuan_likelihood_2021].
+
+Early MR methods account manage horizontal pleiotropy by removing instrumental SNPs 
+that may be associated with the outcome variable [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
+
 # Correlated horizontal pleiotropy
 
-@yuan_testing_2020
+The second type of horizontal pleiotropy, from above, is sometimes called "correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs when 
+CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy.
 

### 2022-02-08 mr-review.Rmd
merge fixes

--- 2022-02-07 mr-review.Rmd
@@
 
+Researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
+exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
+SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
+in the presence of unobserved exposure-outcome confounding. It induces correlation between 
+horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
+violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
+discoveries [@yuan_likelihood_2021].
 Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
@@
 
+Early MR methods account manage horizontal pleiotropy by removing instrumental SNPs 
+that may be associated with the outcome variable [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
+
 # Correlated horizontal pleiotropy
 
-@yuan_testing_2020
+The second type of horizontal pleiotropy, from above, is sometimes called "correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs when 
+CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy.
 

### 2022-02-09 mr-review.Rmd
added two new section headers
noted
minor fixes

@@
 
-The approaches to MR can be classified into two categories: two-stage methods and 
+Many approaches to MR can be classified into two categories: two-stage methods and 
 likelihood-based methods. Both two-stage strategies and likelihood-based strategies 
@@
 
-where the $\epsilon_i$ are assumed to be iid normal random variables with mean zero.
+where the $\epsilon_i$ are assumed to be independent and identically distributed normal random variables with mean zero.
 
@@
 
-where the $\tau_i$ are assumed to be iid normal random variables with mean zero and $\hat x_i$ denotes the fitted value for $x_i$ obtained from Equation \@ref(eq:lm1).
+where the $\tau_i$ are assumed to be independent and identically distributed  normal random variables with mean zero and $\hat x_i$ denotes the fitted value for $x_i$ obtained from Equation \@ref(eq:lm1).
 
@@
 
-where $\epsilon$ is assumed to be a vector of iid normal random variables with mean zero.
+where $\epsilon$ is assumed to be a vector of independent and identically distributed normal random variables with mean zero.
 
@@
 in GWAS. This finding motivated the development of MR-PRESSO and inspired 
-further research into methods to model and to account for horizontal pleiotropy [@verbanck_detection_2018].
+further research into methods to model horizontal pleiotropy [@verbanck_detection_2018].
 
@@
 
+# Software
+
+# Non-continuous outcomes
+
+Binary, survival, count - see discussion section of MRAID paper.
+
 # Summary and Outlook

### 2022-02-09 mr-review.Rmd
tweaks and added more to correlated horizontal pleio section

@@
 are ranked by the strength of their marginal association with the exposure 
-variable. The LD clumping procedure then selects SNPs that are not in high LD 
-with the top-ranking SNPs.
+variable. The LD clumping procedure then selects high ranking SNPs that are not in high LD with each other.
 
@@
 reasons. First, the chosen SNPs may not be causal for the outcome. This can 
-happen, for instance, by choosing a noncausal SNP that is in high LD with a 
-causal SNP.
+happen, for instance, by using a noncausal SNP that is in high LD with a 
+causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
+causal SNPs may be in high LD together. LD clumping is likely to omit one or more
+of the causal SNPs.
 
@@
 
-Choosing a subset of the available SNPs to serve as instruments is one consideration that arises 
+Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
 in MR studies. The solution to this task differs among published methods. Early methods chose a 
-set of independent SNPs. However, this approach can be problematic when there are multiple causal 
+set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
 SNPs in LD with each other. In this setting, discarding some of the causal SNPs may only capture a small
@@
 Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
-instrumental SNPs that might be associated with the outcome [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
-More recent methods have attempted to model horizontal pleiotropy [@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021].
-CAUSE [@morrison_mendelian_2020] mdoels two varieties of horizontal pleiotropy:
+instrumental SNPs that might be associated with the outcome 
+[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
+More recent methods have attempted to model horizontal pleiotropy 
+[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
+MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
+types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
+challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
+Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
+MRMix uses ...
+CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
+the causal effect parameter. Specifically, CAUSE calculates the expected log pointwise posterior density 
+between the two models. 
+MRAID models horizontal pleiotropy by ...
+@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
+biased. CAUSE, they found, yields overly conservative p-values.
 

### 2022-02-09 mr-review.Rmd
added several refs

@@
 discovery of causal relationships via MR. A GWAS
-interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait. A 
-SNP is a genetic marker defined by nucleotide base identity at a single genomic position.
+interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
+In the nearly 17 years since publication of early studies, researchers have reported 
+tens of thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].
 
@@
 Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
-appearance of the early GWAS (Figure \@ref(fig:article-counts)).
-Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted median estimation [@bowden_consistent_2016], Bayesian weighted MR [@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix [@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID [@yuan_likelihood_2021] are all recent methods developments.
-These methods differ in their approaches to two challenges:
+appearance of the early GWAS (Figure \@ref(fig:article-counts)). Inverse variance 
+weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
+median estimation [@bowden_consistent_2016], Bayesian weighted MR 
+[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
+[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
+[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
+their approaches to two challenges:
 
@@
 are ranked by the strength of their marginal association with the exposure 
-variable. The LD clumping procedure then selects high ranking SNPs that are not in high LD with each other.
+variable. The LD clumping procedure then outputs high ranking SNPs that are not in high LD with each other.
 
@@
 causal SNPs may be in high LD together. LD clumping is likely to omit one or more
-of the causal SNPs.
-
-@verbanck_detection_2018
+of the causal SNPs, which can diminish statistical power in MR analysis.
 
@@
 GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is 
-an abundance of freely available GWAS summary data [@buniello2019nhgri]. GWAS summary statistics 
-typically consist of an identifier and a p-value for every SNP in the data set, along with the sample 
-size [@buniello2019nhgri].
-
-Modern GWAS for biobank-scale data offer the possibility
-of discovering even more causal relationships among complex traits. The large
-sample sizes in biobank studies, often greater than 100,000 
-subjects, offer the possibility of uncovering associations of modest strength 
-that might be undetectable in smaller studies. For example, the UK 
-Biobank Study involves more than 500,000 subjects. Each UKB subject's 
-data consists of millions of SNP genotypes and measurements for 
-thousands of traits [@bycroft_uk_2018].
+an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
+ability to use published has led to a rapid increase in the number of available
+two-sample MR methods for GWAS summary statistics.
 
@@
 Investigators have found that combining information from multiple related 
-traits can lead to richer MR results.
+traits can lead to richer MR results [@liu_multi-trait_2021].
 
@@
 Binary, survival, count - see discussion section of MRAID paper.
+@allman_novel_2021 @burgess_improving_2012
 
@@
 
+Modern GWAS for biobank-scale data offer the possibility
+of discovering even more causal relationships among complex traits. The large
+sample sizes in biobank studies, often greater than 100,000 
+subjects, offer the possibility of uncovering associations of modest strength 
+that might be undetectable in smaller studies. For example, the UK 
+Biobank Study involves more than 500,000 subjects. Each UKB subject's 
+data consists of millions of SNP genotypes and measurements for 
+thousands of traits [@bycroft_uk_2018].
+
 # References

### 2022-02-09 mr-review.Rmd
added text

@@
 Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
-appearance of the early GWAS (Figure \@ref(fig:article-counts)). Inverse variance 
+publication of the early GWAS (Figure \@ref(fig:article-counts)). Inverse variance 
 weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
@@
 
-Recent developments have enabled execution of two-sample MR studies with only 
-GWAS summary statistics [@burgess2015using]. These methods are especially pertinent now that there is 
+Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
+the standard MR assumptions and can lead to biased 
+causal effect estimates and diminished statistical power. We discuss approaches to 
+modeling horizontal pleiotropy below.
+
+Recent methods developments have enabled execution of two-sample MR studies with only 
+GWAS summary statistics [@burgess2015using]. These strategies are especially pertinent now that there is 
 an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
-ability to use published has led to a rapid increase in the number of available
+ability to use published data has contributed to a rapid increase in the number of
 two-sample MR methods for GWAS summary statistics.
@@
 
-The outcome variable can be binary, a count, or a continuous variable. For brevity and simplicity, 
-we focus on continuous outcomes below.
+The outcome variable can be binary, a count, a time to event, or a continuous variable. For brevity and simplicity, 
+we focus on continuous outcomes before considering other classes of outcome variables in a later section.
 
@@
 of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
-the strength of association between the SNP and exposure.
+the strength of association between the SNP and exposure. We later discuss, in the context of 
+methods for multiple SNP instruments, how this finding, coupled with the recognition of 
+the polygenicity of many complex traits, has motivated methods to work with large numbers of weak SNP instruments [@ye_debiased_2021;@chao_consistent_2005].
 
-The second assumption, often termed the exclusion restriction assumption, can also be stated as 
+The second MR assumption, often termed the exclusion restriction assumption, can also be stated as 
 the need for the genetic marker to affect the outcome only through the exposure. In a 
@@
 with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
-causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
+causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008]. It's important 
+to recognize that assumptions 2 and 3, which require absence of two types of confounding, 
+can't be verified in observational studies. For this reason, researchers have developed 
+sensitivity analysis to consider the robustness of their inferences to possible violations 
+of assumptions 2 and 3.
 
@@
 
-A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
+A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
 outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
@@
 
-# Mendelian Randomization with multiple instruments
+# Mendelian Randomization with multiple SNP instruments
 
@@
 methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
-
 Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
@@
 with $S$ SNP instruments, one performs two stage analysis by first regressing the 
-exposure $n$ vector $X$ on the $n$ by $S$ matrix of SNP instruments. Taking the fitted exposure values 
+exposure $n$ vector $X$ on the $n$ by $S + 1$ matrix of SNP instruments (and an intercept). Taking the fitted exposure values 
 from this regression, the second stage involves regressing the outcome variable $Y$ on 
@@
 \begin{equation}
-  X  = \alpha Z + \epsilon
+  X  = Z\alpha + \epsilon
   (\#eq:lm3)
@@
 \begin{equation}
-  Y = \beta \hat X + \tau
+  Y = \hat X\beta + \tau
   (\#eq:lm4)
@@
 
-Horizontal pleiotropy refers to the presence of multiple pathways from a genetic 
-variant to the outcome. One of these multiple pathways may contain the exposure variable.
-The fact that there are multiple pathways from the genetic variant to the outcome violates
+Pleiotropy, where a single genetic variant affects multiple traits, has a long history of study in genetics and complex traits [@solovieff_pleiotropy_2013;@visscher_plethora_2016;@stearns_one_2010].
+Horizontal pleiotropy refers to the presence of multiple causal pathways from a genetic
+variant to the outcome. 
+The fact that there are pathways that exclude the exposure and connect the SNP to the outcome violates
 standard MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
@@
 CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
-the causal effect parameter. Specifically, CAUSE calculates the expected log pointwise posterior density 
+the causal effect parameter [@morrison_mendelian_2020]. Specifically, CAUSE calculates the expected log pointwise posterior density 
 between the two models. 
@@
 @yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
-biased. CAUSE, they found, yields overly conservative p-values.
+biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].
 

### 2022-02-10 mr-review.Rmd
added sensitivity analysis text

@@
 
-@burgess_sensitivity_2017 divided the suite of methods for sensitivity analysis 
-into two categories: 1) those that support the validity of causal inferences in a MR 
-study and 2) those that question that validity. Because not all confounding 
+Because not all confounding 
 variables are known or measured, assumptions 2 and 3 are not fully verifiable 
@@
 carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].
+
+@burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when 
+working with GWAS summary statistics. @burgess_sensitivity_2017 presented methods for both assessing the
+MR assumptions, to the extent possible, and performing robust analyses. For example, 
+they used measured covariates to assess for possible associations with the SNP 
+instruments. While they can't rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.
+
+To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of 
+C-reactive protein (CRP) levels on coronary artery disease risk with four genetic 
+variants in the CRP gene region and 17 other genetic variants that affect coronary 
+artery disease risk. They use measured covariates to probe for SNP-covariate 
+associations, and followed it up with scatter plots and Cochran's Q test on the causal 
+estimates to inquire about whether the 
+SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003]. 
+Additionally, @burgess_sensitivity_2017 suggests using a funnel plot, like those in the 
+meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011].
+Additionally, Egger regression can be useful in this
+setting [@egger_bias_1997;@bowden_mendelian_2015].
 

### 2022-02-11 article-counts-per-year-figure.Rmd
redid figure with pub counts

@@
   tidyr::pivot_longer(!year) %>%
-  ggplot2::ggplot(ggplot2::aes(x = year, y = value, colour = name)) + ggplot2::geom_point()
-ggplot2::ggsave("figures/mr-gwas-counts-scatter.png")
+  dplyr::filter(year != 2022) %>%
+  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
+  ggplot2::geom_bar(stat = "identity", 
+                    position=ggplot2::position_dodge()) + 
+  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
+  ggplot2::ylab("Annual publication count") + 
+  ggplot2::xlab("Year") + 
+  ggplot2::guides(fill = ggplot2::guide_legend("Search terms"))
+
+ggplot2::ggsave("figures/mr-gwas-counts.png")
 ```

### 2022-02-11 mr-review.Rmd
redid figure with pub counts

@@
 
-```{r setup, include=FALSE}
+```{r setup}
 knitr::opts_chunk$set(echo = TRUE)
+library(magrittr)
 ```
@@
 ```{r, article-counts, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
-knitr::include_graphics("figures/mr-gwas-counts-scatter.png")
+knitr::include_graphics("figures/mr-gwas-counts.png")
 ```
@@
 ```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}
-g1 <- dagitty::dagitty( "dag {
-	G -> W -> X -> Y
-	W -> Y
-}")
-plot(dagitty::graphLayout(g1))
+
 ```
@@
 
-## Statistical models and methods for MR with multiple instruments and one outcome
+## Statistical models and methods for MR with multiple instruments and one continuous outcome
 
@@
 
+# Binary, Count, and Time-to-event outcomes
+
+Binary, survival, count - see discussion section of MRAID paper.
+@allman_novel_2021 @burgess_improving_2012
+
 # Recent Findings
@@
 
-# Non-continuous outcomes
-
-Binary, survival, count - see discussion section of MRAID paper.
-@allman_novel_2021 @burgess_improving_2012
-
 # Summary and Outlook

### 2022-02-11 mr-review.Rmd
minor rewording of a sentence

@@
 
-Researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
+In the short time since publication of @verbanck_detection_2018, researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
 exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
@@
 CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
-the causal effect parameter [@morrison_mendelian_2020]. Specifically, CAUSE calculates the expected log pointwise posterior density 
+the causal effect parameter [@morrison_mendelian_2020]. Specifically, CAUSE calculates the expected pointwise log posterior density 
 between the two models. 

### 2022-02-12 mr-review.Rmd
text about CAUSE

@@
 
+and $\tau$ is a vector of independent and identically distributed normal random errors with mean zero.
+
 # Horizontal pleiotropy
@@
 
-In the short time since publication of @verbanck_detection_2018, researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
+In the short time since publication of @verbanck_detection_2018, researchers have recognized two 
+types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
 exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
@@
 Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
-MRMix uses ...
 CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
@@
 between the two models. 
-MRAID models horizontal pleiotropy by ...
 @yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
@@
 CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy. 
+Their approaches to modeling correlated horizontal pleiotropy differ in important ways that, ultimately, 
+affect performance. 
+CAUSE allows for a proportion of variants to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
+models as an effect on a shared, unobserved factor. The remaining variants are independent of this 
+unobserved factor. Every variant can have a nonzero pleiotropic effect on the outcome, and these 
+pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, 
+is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on 
+adaptive shrinkage from @stephens_false_2016. Finally, @morrison_mendelian_2020 compares two models 
+to determine whether the GWAS summary statistics are consistent with a causal effect of the
+exposure on the outcome. Specifically, they estimate the difference in the expected log pointwise 
+posterior density for the model with causal effect fixed at zero and the model that permits a nonzero 
+causal effect [@vehtari_practical_2017].
 

### 2022-02-12 mr-review.Rmd
echo = FALSE
added more about mendel

@@
 
-```{r setup}
-knitr::opts_chunk$set(echo = TRUE)
+```{r setup, echo = FALSE}
+knitr::opts_chunk$set(echo = FALSE)
 library(magrittr)
@@
 To understand the origin of the term "Mendelian randomization", it is helpful to 
-recall Mendel's inheritance laws. Under the assumption that alleles segregate 
+recall Mendel's inheritance laws [@mendel1865versuche]. Under the assumption that alleles segregate 
 randomly from parent to offspring, the offspring genotypes are unlikely to be 
 associated with confounders of the exposure-outcome relationship. Additionally,
-reverse causation is ruled out by the observation that germ-line genotypes are 
-fixed at conception and, thus, precede other observed variables.
+reverse causation is unanticipated since germ-line genotypes are 
+fixed at conception and, thus, precede realization of other observed variables. Reverse causation 
+occurs when the outcome at an early time point has a causal effect on the exposure variable. While 
+reverse causation is often assumed to not occur in MR study designs, @burgess_are_2021 points out 
+several scenarios where this assumption may not hold [@davey_smith_mendelian_2003].
 
@@
 
-# Binary, Count, and Time-to-event outcomes
+# Binary, Count, and Time-to-event outcomes and exposures
 
@@
 @allman_novel_2021 @burgess_improving_2012
+
+While we've focused on continuous outcomes and continuous exposures, some
+data are more naturally treated as binary, count, or time-to-event variables. MR methods 
+for non-continuous outcomes and exposures is an active area of research. @burgess_improving_2012 and @allman_novel_2021 
+developed MR strategies for binary exposures and binary outcomes. @burgess_improving_2012 extended 
+methods from @wald_fitting_1940 by fitting two logistic regressions, one for the outcome and one for 
+the exposure. They then calculated the causal effect estimator as the ratio of the two logistic 
+regression coefficient estimators. @allman_novel_2021 devised a new MR method for binary outcome and 
+binary exposure by drawing on @gauvreau_analysis_1997 and treating the exposure and outcome as correlated binary variables.
+Writing the likelihood for this model, they implemented an iterative optimization algorithm to simultaneously infer the causal effect parameter and other parameters.
 

### 2022-02-13 mr-review.Rmd
added count and survival outcome methods

@@
 The second type of horizontal pleiotropy, from above, is sometimes called "correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs when
+
 CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy. 
@@
 
-Binary, survival, count - see discussion section of MRAID paper.
-@allman_novel_2021 @burgess_improving_2012
-
-While we've focused on continuous outcomes and continuous exposures, some
+While we've focused above on continuous outcomes and continuous exposures, some
 data are more naturally treated as binary, count, or time-to-event variables. MR methods 
@@
 regression coefficient estimators. @allman_novel_2021 devised a new MR method for binary outcome and 
-binary exposure by drawing on @gauvreau_analysis_1997 and treating the exposure and outcome as correlated binary variables.
-Writing the likelihood for this model, they implemented an iterative optimization algorithm to simultaneously infer the causal effect parameter and other parameters.
+binary exposure by drawing on @gauvreau_analysis_1997 and treating the exposure and outcome as correlated binary 
+random variables. They implemented an iterative optimization algorithm to 
+simultaneously infer the causal effect parameter and other parameters.
+
+Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017]. 
+@guo_body_2017, for example, studied post-diagnosis survival time in women with breast cancer. They 
+treated survival as a time-to-event outcome, which enabled them to model it with a proportional 
+hazards model [@cox_regression_1972] in two-stage MR with individual-level data. Similarly, 
+@he_causal_2018, in a study of cardiovascular disease, applied proportional hazards regression when 
+examining time to stroke as an outcome variable. We are unaware of MR studies that treat 
+the exposure variable as a time-to-event, but it's conceivable that future research will consider 
+time-to-event traits as exposures in MR analyses.
+
+Researchers have also drawn on other generalized linear models for analysis in MR 
+studies. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
+modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
+power, with simulations of count random variables and binary random variables. With growing interest
+in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
+there may be opportunities for methods innovations in a generalized linear models framework for MR studies.
+
+See also @tchetgen_tchetgen_genius_2021
 
@@
 
-# Software
-
 # Summary and Outlook

### 2022-02-14 mr-review.Rmd
added some biotech refs

@@
 In the nearly 17 years since publication of early studies, researchers have reported 
-tens of thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].
+thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].
 
@@
 @he_causal_2018, in a study of cardiovascular disease, applied proportional hazards regression when 
-examining time to stroke as an outcome variable. We are unaware of MR studies that treat 
+examining time to stroke as an outcome variable. In very recent work, @tchetgen_tchetgen_genius_2021
+reported methods, MR GENIUS, that are robust to violations of the exclusion restriction assumption (under 
+many assumed MR data generating processes). They considered binary, continuous, and time-to-event 
+outcomes within their framework. We are unaware of MR studies that treat 
 the exposure variable as a time-to-event, but it's conceivable that future research will consider 
@@
 in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
-there may be opportunities for methods innovations in a generalized linear models framework for MR studies.
-
-See also @tchetgen_tchetgen_genius_2021
+there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].
 

### 2022-02-14 mr-review.Rmd
minor update

@@
 causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
+Because assumptions 2 and 3, which involve absence of certain types of confounding, can't be verified, 
+practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
+way to assess robustness of estimates in the presence of possible assumption violations.
+
 The first assumption states that the exposure is associated with the outcome. The second requires 
@@
 For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is 
-incorrect in two-stage calculations. This observation led researchers to develop 
+incorrect in two-stage calculations. This and other observations led researchers to develop 
 likelihood-based MR methods.
@@
 Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
-estimates with their many desirable properties. Limited information maximum likelihood from econometrics 
-is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020].
-
-Limited information maximum likelihood model equations are shown below (Equations \@ref(eq:lm5) and \@ref(eq:lm6)).
-
-\begin{equation}
-  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
-  (\#eq:lm5)
-\end{equation}
-
-\begin{equation}
-  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
-  (\#eq:lm6)
-\end{equation}
-
-They differ from the two-stage model because the two error terms, $\epsilon_i$ and $\tau_i$, have a bivariate normal
-distribution.
+estimates with their many desirable properties [@fisher1925likelihood]. Limited information maximum likelihood from econometrics 
+is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and 
+PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020]. We discuss 
+PMR-Egger below in the context of methods that account for horizontal pleiotropy. Immediately below, we 
+discuss CoMM.
 
 CoMM, along with its extension for summary statistics, CoMM-S2, uses a collaborative mixed model to 
-relate exposure and outcome variables to the SNP genotypes [@yang_comm-s2_2020]. Model fitting then 
-proceeds via a parameter-expanded EM algorithm [@yang_comm_2019;@liu_parameter_1998].
-
-We discuss PMR-Egger below in the context of methods that model horizontal pleiotropy.
+relate exposure and outcome variables to the SNP genotypes [@yang_comm-s2_2020]. In this model, the 
+authors jointly model two linear mixed effects models, one relating the outcome to the SNPs and a second 
+tying the exposure to the SNPs. Model fitting then 
+proceeds via a parameter-expanded EM algorithm [@yang_comm_2019;@liu_parameter_1998]. @yang_comm_2019, 
+working with gene expression traits as exposures, found that CoMM better prioritized genetic variants 
+relative to other methods [@yang_comm_2019].
 
@@
 
-There are several ways that experimental data may violate one or more of the three MR assumptions.
+Now that we've discussed several MR methods, it is worthwhile to consider ways in which the standard MR assumptions may be violated. 
 Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
@@
 
-```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}
+```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X.
+Because W affecs Y through pathways that don't involve exposure X, this scenario violates the exclusion restriction assumption."}
+
+knitr::include_graphics()
 

### 2022-02-14 mr-review.Rmd
obsolete code deleted
added ref
fixed fisher ref

@@
 Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
-estimates with their many desirable properties [@fisher1925likelihood]. Limited information maximum likelihood from econometrics 
+estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
 is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and 
@@
 Because W affecs Y through pathways that don't involve exposure X, this scenario violates the exclusion restriction assumption."}
-
-knitr::include_graphics()
 
@@
 Researchers have also drawn on other generalized linear models for analysis in MR 
-studies. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
+studies [@nelder1972generalized]. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
 modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 

### 2022-02-15 mr-review.Rmd
added dagitty

@@
 ```{r setup, echo = FALSE}
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, message = FALSE)
 library(magrittr)
@@
 
-```{r, article-counts, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
-knitr::include_graphics("figures/mr-gwas-counts.png")
+```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
+library(magrittr)
+dat <- readxl::read_xlsx(path = "data/publication-counts.xlsx", sheet = "counts")
+dat %>%
+  tidyr::pivot_longer(!year) %>%
+  dplyr::filter(year < 2022) %>%
+  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
+  ggplot2::geom_bar(stat = "identity", 
+                    position=ggplot2::position_dodge()) + 
+  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
+  ggplot2::ylab("Annual publication count") + 
+  ggplot2::xlab("Year") + 
+  #ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
+  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
+  ggplot2::theme(text = ggplot2::element_text(size = 20))
+
+ggplot2::ggsave("figures/mr-gwas-counts.png")
+
 ```
@@
 ```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X.
-Because W affecs Y through pathways that don't involve exposure X, this scenario violates the exclusion restriction assumption."}
+Because W affects Y through a pathway that doesn't involve exposure X, this scenario violates the exclusion restriction assumption."}
+
+library(dagitty)
+library(ggdag)
+library(ggplot2)
+
+d1 <- dagitty('dag {
+bb="0,0,1,1"
+G [pos="0,0"]
+W [pos="0.5,0"]
+X [exposure, pos="1.0,0"]
+Y [outcome, pos="1.5,0"]
+G -> W
+W -> X
+W -> Y [pos="1,1"]
+X -> Y
+}')
+
+plot(d1)
+
+```
+
+```{r, dag2, fig.cap = "Scenario where the genetic variant G affects exposure X, which in turn affects Y. Unobserved confounder U affects both X and Y."}
+d1 <- dagitty('dag {
+bb="0,0,1,1"
+G [pos="0,0"]
+U [pos="0.75,10"]
+X [exposure, pos="0.5,0"]
+Y [outcome, pos="1.0,0"]
+G -> X
+U -> X
+U -> Y
+X -> Y
+}'
+)
+plot(d1)
 

### 2022-02-15 mr-review.Rmd
started writing applications section

@@
 bb="0,0,1,1"
-G [pos="0,0"]
-U [pos="0.75,10"]
-X [exposure, pos="0.5,0"]
-Y [outcome, pos="1.0,0"]
+G [pos="0.0,0.0"]
+U [pos="0.75,0.5"]
+X [exposure, pos="0.5,0.0"]
+Y [outcome, pos="1.0,0.0"]
 G -> X
@@
 variants in the CRP gene region and 17 other genetic variants that affect coronary 
-artery disease risk. They use measured covariates to probe for SNP-covariate 
+artery disease risk. They used measured covariates to probe for SNP-covariate 
 associations, and followed it up with scatter plots and Cochran's Q test on the causal 
@@
 SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003]. 
-Additionally, @burgess_sensitivity_2017 suggests using a funnel plot, like those in the 
+Additionally, @burgess_sensitivity_2017 suggested using a funnel plot, like those in the 
 meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011].
 Additionally, Egger regression can be useful in this
-setting [@egger_bias_1997;@bowden_mendelian_2015].
+setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].
 
@@
 and $\tau$ is a vector of independent and identically distributed normal random errors with mean zero.
+Note also that $\tau$ in Equation \@ref(eq:lm4) is independent of $\epsilon$ in Equation \@ref(eq:lm3).
 
@@
 
+Birthweight is thought to impact long-term susceptibility to adulthood diseases. This scientific 
+hypothesis, sometimes called the "fetal origins of adult diseases" motivated work by @zeng_causal_2019 in 
+which the authors used MR to identify causal relationships between lower birthweight and greater risks of 
+coronary artery disease, myocardial infarction, and type 2 diabetes mellitus [@barker_fetal_1990;@lucas_fetal_1999;@barker_fetal_2002]. Specifically, 
+@zeng_causal_2019 used 
+inverse variance-weighted MR to examine a total of 21 adult diseases and 38 other complex traits. Three causal associations met the Bonferroni multiple hypothesis testing threshold 
+[@bonferroni1936teoria]. The authors analyzed data from 37 GWAS with up to 340,000 subjects with European 
+ancestry [@zeng_causal_2019]. They estimated odds ratios of 1.34, 1.30, and 1.41 for coronary artery 
+disease, myocardial infarction, and type 2 diabetes mellitus, respectively. Sensitivity analyses revealed 
+the robustness of these findings to possible MR assumptions violations, including pleiotropy, maternal 
+effects, and reverse causation [@zeng_causal_2019]. Curiously, none of the 38 other complex traits 
+appeared to mediate the causal associations, perhaps due to limited sample sizes for these traits [@zeng_causal_2019]. @zeng_causal_2019 argue that further 
+studies, including consideration of socioeconomic status and incorporation of larger sample sizes, may 
+reveal mediating factors for these causal relationships and, thus, provide biological and public health insights.
+
+@zeng_causal_2019-1
+
+@zeng_birth_2019
+
 # Summary and Outlook

### 2022-02-16 mr-review.Rmd
added paragraph about zeng_causal_2019-1

@@
 
-@zeng_causal_2019-1
+Amyotrophic lateral sclerosis is an incurable and largely untreatable neurological condition [@rowland_amyotrophic_2001;@kiernan_amyotrophic_2011].
+Working with GWAS summary statistics, @zeng_causal_2019-1 found evidence of a causal relationship between
+type 2 diabetes mellitus and amyotrophic lateral sclerosis in both European and East Asian populations. 
+Interestingly, these effects differ in direction; in the European population, type 2 diabetes mellitus 
+exerted a neuroprotective effect and diminished the risk of amyotrophic lateral sclerosis, while in the 
+East Asian population type 2 diabetes mellitus increased the risk of amyotrophic lateral sclerosis. 
+@zeng_causal_2019-1 used inverse variance-weighted methods to estimate the causal effect from summary 
+statistics derived from approximately 660,000 type 2 diabetes mellitus European subjects and 81,000 
+amyotrophic lateral sclerosis European subjects. Smaller sample sizes (approximately 191,000 type 2 
+diabetes mellitus subjects and 4100 amyotrophic lateral sclerosis subjects) were available in the East 
+Asian cohorts. The authors used multiple SNP instruments with strong associations with type 2 diabetes 
+mellitus [@zeng_causal_2019-1]. The odds ratios of 0.93 and 1.28 in the Europeans and East Asians, 
+respectively, were robust against instrument selection, various model misspecifications, and estimation 
+biases [@zeng_causal_2019-1]. Egger regression and MR-PRESSO analyses were consistent with the absence of horizonal pleiotropy 
+[@zeng_causal_2019-1;@hemani_evaluating_2018;@verbanck_detection_2018;@ong2019implementing]. Future studies may reveal deeper biological insights into these observations and, ultimately, 
+contribute to treatments for amyotrophic lateral sclerosis.
 

### 2022-02-16 mr-review.Rmd
added refs from OMR paper

@@
 context of the omnigenic hypothesis. To accommodate the possibility
-that a trait is influenced by many genome-wide, small effects,
+that a trait is influenced by many SNPs, @wang_mendelian_nodate
 
@@
 # Applications of MR in GWAS
+
+MR analysis has revealed causal relationships and led to mechanistic insights for many 
+common diseases [@mancuso2017integrating;@parisinos2020genome;@zeng_causal_2019-2;@larsson2019smoking;@zheng2020phenome;@robinson2016insight;@luo2019association;@au2018impact;@zhao2019effect;@zeng_causal_2019;@zeng_causal_2019-1].
 

### 2022-02-17 mr-review.Rmd
eqns for MRAID

--- 2022-02-16 mr-review.Rmd
@@
 affect performance. 
-CAUSE allows for a proportion of variants to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
-models as an effect on a shared, unobserved factor. The remaining variants are independent of this 
-unobserved factor. Every variant can have a nonzero pleiotropic effect on the outcome, and these 
+CAUSE allows for a proportion of SNPs to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
+models as an effect on a shared, unobserved factor. The remaining SNPs are independent of this 
+unobserved factor. Every SNP can have a nonzero pleiotropic effect on the outcome, and these 
 pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, 
@@
 causal effect [@vehtari_practical_2017].
+
+MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
+the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref() and \@ref()).
+
+\begin{equation}
+  \hat{\beta_X} = \Sigma_1\beta + e_x
+  (\#eq:mraid1)
+\end{equation}
+
+\begin{equation}
+  \hat{\beta_Y} = \alpha \Sigma_2 \beta + \Sigma_2\eta_0 + \Sigma_2 \eta_1 + e_y
+  (\#eq:mraid2)
+\end{equation}
+
+The estimated marginal effects for the $p$ SNP instruments on the exposure are denoted by $\hat \beta_X$, while 
+$\hat\beta_Y$ represents the estimated effects of the same SNPs on the outcome. The $p$ by $p$ correlation matrices for 
+the SNP instruments are written as $\Sigma_1$ and $\Sigma_2$, respectively. $\Sigma_1$ and $\Sigma_2$ can be estimated with 1000 Genomes 
+Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry 
+[@the_1000_genomes_project_consortium_1000_2012]. The $p$-vector error terms, $e_x$ and $e_y$, follow multivariate normal 
+distributions with mean zero and variances $\Sigma_1\frac{\sigma_X^2}{n_1 - 1}$ and $\Sigma_2\frac{\sigma_Y^2}{n_2 - 1}$, 
+respectively.
 
@@
 # Applications of MR in GWAS
+
+MR analysis has revealed mechanistic insights for many 
+common diseases [@mancuso2017integrating;@parisinos2020genome;@zeng_causal_2019-2;@larsson2019smoking;@zheng2020phenome;@robinson2016insight;@luo2019association;@au2018impact;@zhao2019effect;@zeng_causal_2019;@zeng_causal_2019-1].
 

### 2022-02-17 mr-review.Rmd
added snp assumptions from mraid

@@
 [@the_1000_genomes_project_consortium_1000_2012]. The $p$-vector error terms, $e_x$ and $e_y$, follow multivariate normal 
-distributions with mean zero and variances $\Sigma_1\frac{\sigma_X^2}{n_1 - 1}$ and $\Sigma_2\frac{\sigma_Y^2}{n_2 - 1}$, 
+distributions with mean zero and variances $\Sigma_1\frac{\sigma_X^2}{n_1 - 1}$ 
+and $\Sigma_2\frac{\sigma_Y^2}{n_2 - 1}$, 
 respectively.
+@yuan_likelihood_2021 then constructs a Gibbs sampler to make likelihood-based inferences [@geman1984stochastic].
+To facilitate computations, the investigators make several assumptions about the collection of SNPs:
+
+1. Relatively small proportion of SNPs have nonzero effects on the exposure  
+1. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
+1. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs  
+1. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.
+
+By encoding these assumptions in prior distributions, @yuan_likelihood_2021 enables the inferential procedures to accommodate the observed data in the context of the assumptions on SNPs.
 

### 2022-02-17 mr-review.Rmd
updated timeline inputs

@@
 "two-sample" MR studies because the exposure variable and the outcome variable are measured on 
-two non-overlapping cohorts of subjects. Importantly, the SNP instruments are measured in both sets of subjects.
+two non-overlapping cohorts of subjects [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.
 
@@
 are ranked by the strength of their marginal association with the exposure 
-variable. The LD clumping procedure then outputs high ranking SNPs that are not in high LD with each other.
+variable.
 
@@
 modeling horizontal pleiotropy below.
+
+```{r timeline, fig.cap="Timeline of two-sample GWAS summary statistics MR methods"}
+readxl::read_xlsx("data/method-year.xlsx", sheet = "inputs")
+```
 
@@
 The estimated marginal effects for the $p$ SNP instruments on the exposure are denoted by $\hat \beta_X$, while 
-$\hat\beta_Y$ represents the estimated effects of the same SNPs on the outcome. The $p$ by $p$ correlation matrices for 
-the SNP instruments are written as $\Sigma_1$ and $\Sigma_2$, respectively. $\Sigma_1$ and $\Sigma_2$ can be estimated with 1000 Genomes 
+$\hat\beta_Y$ represents the estimated effects of the same SNPs on the outcome. The $p$ by $p$ SNP instrument correlation matrices for the exposure and outcome 
+are written as $\Sigma_1$ and $\Sigma_2$, respectively. $\Sigma_1$ and $\Sigma_2$ can be estimated with 1000 Genomes 
 Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry 

### 2022-02-18 mr-review.Rmd
deleted unused R code

@@
 
+# Abstract
+
+Complex trait genetics studies in humans have yielded thousands of new associations for  
+diverse traits. The next step in understanding biology of health and disease is to
+identify causal relationships among complex traits. Two-sample Mendelian randomization studies 
+in genome-wide association studies can uncover causal relationships 
+between complex traits. By leveraging shared SNP genotypes as instrumental variables, or proxies, for 
+the exposure complex trait, investigators can tease out causal effects from observational data, provided
+that three assumptions are satisfied. We discuss below the growth of genome-wide association studies 
+in parallel with development of Mendelian randomization methods.
+We argue that the recent availability of GWAS summary statistics for diverse complex traits has led to new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.
+
 # Introduction
@@
 Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
-genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments for 
+genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
 inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].
@@
 thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].
+Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and refined our understanding of
+human health and disease.
 
@@
 independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
-selected through linkage disequilibrium (LD) clumping. In this procedure, SNPs 
-are ranked by the strength of their marginal association with the exposure 
-variable.
+selected through linkage disequilibrium (LD) clumping.
 
-LD clumping for choosing instrument SNPs can be problematic for several 
+However, LD clumping for choosing instrument SNPs can be problematic for several 
 reasons. First, the chosen SNPs may not be causal for the outcome. This can 
@@
 causal effect estimates and diminished statistical power. We discuss approaches to 
-modeling horizontal pleiotropy below.
-
-```{r timeline, fig.cap="Timeline of two-sample GWAS summary statistics MR methods"}
-readxl::read_xlsx("data/method-year.xlsx", sheet = "inputs")
-```
+modeling and accounting for horizontal pleiotropy below.
 
@@
 The third assumption ensures that the genetic marker is
-independent of all confounders of the exposure-outcome relationship.
+independent of all confounders of the exposure-outcome relationship. Like the exclusion restriction
+assumption, we can't verify this assumption with observational data, so analysts need to 
+use tools like sensitivity analysis to ensure validity of findings.
 
@@
 with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
-causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008]. It's important 
-to recognize that assumptions 2 and 3, which require absence of two types of confounding, 
-can't be verified in observational studies. For this reason, researchers have developed 
-sensitivity analysis to consider the robustness of their inferences to possible violations 
-of assumptions 2 and 3.
+causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
 

### 2022-02-18 mr-review.Rmd
organizing text...

@@
 Many approaches to MR can be classified into two categories: two-stage methods and 
-likelihood-based methods. Both two-stage strategies and likelihood-based strategies 
-make parametric assumptions for the identification of causal effects [@burgess_review_2017].
-
-A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
-outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
+likelihood-based methods. A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
+outcome, one may perform "two-stage least squares", which involves two linear regressions. 
 First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
@@
 
-## Assumption violations
+# Assumption violations
 
@@
 
-## Sensitivity analysis for MR with one instrument and one outcome
+# Sensitivity analysis
 
@@
 
-probabilistic Mendelian randomization (PMR-Egger) for two-sample MR with correlated SNPs models [@yuan_testing_2020]
-
-Early MR methods account manage horizontal pleiotropy by removing instrumental SNPs 
-that may be associated with the outcome variable [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
-
 # Correlated horizontal pleiotropy
@@
 
-CAUSE and MRAID are two methods that aim to model and to account for correlated horizontal pleiotropy. 
-Their approaches to modeling correlated horizontal pleiotropy differ in important ways that, ultimately, 
+CAUSE and MRAID are two methods that aim to model correlated horizontal pleiotropy [@morrison_mendelian_2020;@yuan_likelihood_2021]. 
+Their approaches to modeling correlated horizontal pleiotropy differ in important ways that ultimately 
 affect performance. 
@@
 MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
-the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref() and \@ref()).
+the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref(mraid1) and \@ref(mraid2)).
 
@@
 
+Researchers have also drawn on other generalized linear models for analysis in MR 
+studies [@nelder1972generalized]. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
+modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
+power, with simulations of count random variables and binary random variables. With growing interest
+in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
+there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].
+
 Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017]. 
@@
 time-to-event traits as exposures in MR analyses.
-
-Researchers have also drawn on other generalized linear models for analysis in MR 
-studies [@nelder1972generalized]. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
-modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
-power, with simulations of count random variables and binary random variables. With growing interest
-in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
-there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].
 

### 2022-02-18 mr-review.Rmd
minor tweaks

@@
 common diseases [@mancuso2017integrating;@parisinos2020genome;@zeng_causal_2019-2;@larsson2019smoking;@zheng2020phenome;@robinson2016insight;@luo2019association;@au2018impact;@zhao2019effect;@zeng_causal_2019;@zeng_causal_2019-1].
+In some cases, a randomized clinical trial is not possible for ethical or other reasons.
+Below, we discuss recent applications of two-sample MR to GWAS summary statistics.
 
@@
 Amyotrophic lateral sclerosis is an incurable and largely untreatable neurological condition [@rowland_amyotrophic_2001;@kiernan_amyotrophic_2011].
-Working with GWAS summary statistics, @zeng_causal_2019-1 found evidence of a causal relationship between
+Working with GWAS summary statistics, @zeng_causal_2019-1 found evidence of causal relationships between
 type 2 diabetes mellitus and amyotrophic lateral sclerosis in both European and East Asian populations. 

### 2022-02-18 mr-review.Rmd
word fix

@@
 in parallel with development of Mendelian randomization methods.
-We argue that the recent availability of GWAS summary statistics for diverse complex traits has led to new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.
+We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.
 

### 2022-02-19 mr-review.Rmd
expanded OMR section

@@
 recognized the limitations of previous MR methods in the 
-context of the omnigenic hypothesis. To accommodate the possibility
-that a trait is influenced by many SNPs, @wang_mendelian_nodate
+context of the omnigenic hypothesis. In response, @wang_mendelian_nodate 
+developed a MR method that explicitly uses all genome-wide SNPs as instruments. 
+Using GWAS summary statistics as inputs, their method relies on a composite likelihood framework for scalable computation and allows for horizontal pleiotropy. @wang_mendelian_nodate used extensive 
+simulations, including those with model misspecifications, to characterize their method's statistical power and robustness. Finally, they applied the new method to 
+identify multiple complex traits that affect coronary artery disease and asthma.
+These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.
 

### 2022-02-19 mr-review.Rmd
deleted multiple outcomes section

@@
 outcome, one may perform "two-stage least squares", which involves two linear regressions. 
-First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
-fitted values for the exposure variable are the independent variable in the second linear regression, 
-where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
+First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)).
+We denote each subject's exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope.
+$z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.
+
+The resulting 
+fitted values for the exposure variable from Equation \@ref(eq:lm1) are the independent variable in the second linear regression, 
+where the outcome is the dependent variable. The causal effect estimate, then, is the 
 coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).
+Note also that the random errors in Equation \@ref(eq:lm2) are assumed independent of those in Equation \@ref(eq:lm1).
+Like the $\epsilon_i$ in Equation \@ref(eq:lm1), the $\tau_i$ are assumed independent and identically distributed normal random errors.
 
@@
 
-where the $\epsilon_i$ are assumed to be independent and identically distributed normal random variables with mean zero.
-
 \begin{equation}
@@
 \end{equation}
-
-where the $\tau_i$ are assumed to be independent and identically distributed  normal random variables with mean zero and $\hat x_i$ denotes the fitted value for $x_i$ obtained from Equation \@ref(eq:lm1).
 
@@
 is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and 
-PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020]. We discuss 
-PMR-Egger below in the context of methods that account for horizontal pleiotropy. Immediately below, we 
-discuss CoMM.
+PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020]. We next discuss CoMM and CoMM-S2.
 
@@
 
-Now that we've discussed several MR methods, it is worthwhile to consider ways in which the standard MR assumptions may be violated. 
+It is worthwhile to consider ways in which the standard MR assumptions may be violated. 
 Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
@@
 the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
-into how the results might differ if the assumptions are violated.
-
+into how the results might differ if the assumptions be violated.
 Because not all confounding 
@@
 context of the omnigenic hypothesis. In response, @wang_mendelian_nodate 
-developed a MR method that explicitly uses all genome-wide SNPs as instruments. 
+developed a MR method that uses all genome-wide SNPs as instruments. 
 Using GWAS summary statistics as inputs, their method relies on a composite likelihood framework for scalable computation and allows for horizontal pleiotropy. @wang_mendelian_nodate used extensive 
@@
 These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.
-
-## Multiple outcomes in MR
-
-Investigators have found that combining information from multiple related 
-traits can lead to richer MR results [@liu_multi-trait_2021].
-
-One approach to combining information across traits is to extend the 
-univariate MR statistical methods to their multivariate analogs.
 

### 2022-02-19 mr-review.Rmd
deleted "and one outcome" from section header

@@
 
-## Statistical models and methods for MR with one instrument and one outcome
+## Statistical models and methods for MR with one instrument
 
@@
 
-```{r, dag2, fig.cap = "Scenario where the genetic variant G affects exposure X, which in turn affects Y. Unobserved confounder U affects both X and Y."}
+```{r, dag2, fig.cap = "Scenario where the genetic variant G affects exposure X, which affects Y. Unobserved confounder U affects both X and Y."}
 d1 <- dagitty('dag {
@@
 
-The second type of horizontal pleiotropy, from above, is sometimes called "correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs when
+The second type of horizontal pleiotropy, from above, is sometimes called 
+"correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs 
+when a SNP affects both the exposure and a variable that confounds the 
+relationship between exposure and outcome (Figure \@ref(fig:dag3)).
+
+```{r, dag3, fig.cap = "Correlated horizontal pleiotropy occurs when the genetic variant G affects exposure X, which in turn affects Y, and G affects U, which in turn affects X and Y independently."}
+d1 <- dagitty('dag {
+bb="0,0,1,1"
+G [pos="0.0,0.0"]
+U [pos="0.75,0.5"]
+X [exposure, pos="0.5,0.0"]
+Y [outcome, pos="1.0,0.0"]
+G -> X
+G -> U
+U -> X
+U -> Y
+X -> Y
+}'
+)
+plot(d1)
+
+```
 

### 2022-02-20 mr-review.Rmd
added notes for conclusion section
added text about zeng 2019 birth weight

@@
 Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
-CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
-the causal effect parameter [@morrison_mendelian_2020]. Specifically, CAUSE calculates the expected pointwise log posterior density 
-between the two models. 
-@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
-biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].
 
@@
 
+@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
+biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].
+
 MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
@@
 
-1. Relatively small proportion of SNPs have nonzero effects on the exposure  
+1. Relatively small proportion of SNPs have nonz
+ero effects on the exposure  
 1. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
@@
 
-@zeng_birth_2019
+Birth weight is associated with childhood asthma, but whether there is a causal relationship 
+between the two remains unclear [@brooks_impact_2001]. To investigate this possibility, 
+@zeng_birth_2019 performed a MR analysis to assess the causal effect of birth weight on adult 
+asthma. They used data from two GWAS, with 140,000 subjects in the birth weight study and 62,000 
+subjects in the adult asthma study. With both the genetic risk score method and the 
+inverse variance-weighted method, @zeng_birth_2019 found no evidence of a causal effect of birth
+weight on adult asthma. Sensitivity analysis results offered reassurance that the 
+finding of no causal effect is robust. MR-PRESSO and MR-Egger found no evidence for bias from 
+SNP instrument outliers or horizontal pleiotropy. @zeng_birth_2019 concludes that confounding
+in other studies may explain the discrepancy between their findings and others' conclusions.
 
@@
 
+Other biobanks?
+
+EMR & GWAS?
+
 # References

### 2022-02-20 mr-review.Rmd
minor tweaks

@@
 
-## Statistical models and methods for MR with multiple instruments and one continuous outcome
+## Statistical models and methods for MR with multiple instruments
 
@@
 exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
-SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
+SNP-exposure relationships. The second type of horizontal pleiotropy manifests
 in the presence of unobserved exposure-outcome confounding. It induces correlation between 
@@
 biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].
-
-MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
+These observations motivated @yuan_likelihood_2021 to develop a new method, MRAID. MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
 the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref(mraid1) and \@ref(mraid2)).
@@
 
-1. Relatively small proportion of SNPs have nonz
-ero effects on the exposure  
+1. Relatively small proportion of SNPs have nonzero effects on the exposure  
 1. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
@@
 
-Researchers have also drawn on other generalized linear models for analysis in MR 
+Researchers have also leveraged other generalized linear models for analysis in MR 
 studies [@nelder1972generalized]. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
@@
 that might be undetectable in smaller studies. For example, the UK 
-Biobank Study involves more than 500,000 subjects. Each UKB subject's 
+Biobank Study involves more than 500,000 subjects. Each UK Biobank subject's 
 data consists of millions of SNP genotypes and measurements for 
@@
 
-Other biobanks?
-
-EMR & GWAS?
-
 # References

### 2022-02-21 mr-review.Rmd
revised multi-SNP text

@@
 set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
-SNPs in LD with each other. In this setting, discarding some of the causal SNPs may only capture a small
+SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
 proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
@@
 
-In the case of continuous exposures and outcomes, researchers have taken several approaches to extend the methods and models for 
-single-SNP MR to accommodate multiple SNPs. For two stage analyses, one may perform 
-a sequence of two linear regressions, as in the single-SNP MR models (Equations \@ref(eq:lm3) and \@ref(eq:lm4)). 
-For an analysis
-with $S$ SNP instruments, one performs two stage analysis by first regressing the 
-exposure $n$ vector $X$ on the $n$ by $S + 1$ matrix of SNP instruments (and an intercept). Taking the fitted exposure values 
-from this regression, the second stage involves regressing the outcome variable $Y$ on 
-the fitted values, $\hat X$.
-
-\begin{equation}
-  X  = Z\alpha + \epsilon
-  (\#eq:lm3)
-\end{equation}
-
-where $\epsilon$ is assumed to be a vector of independent and identically distributed normal random variables with mean zero.
-
-\begin{equation}
-  Y = \hat X\beta + \tau
-  (\#eq:lm4)
-\end{equation}
-
-and $\tau$ is a vector of independent and identically distributed normal random errors with mean zero.
-Note also that $\tau$ in Equation \@ref(eq:lm4) is independent of $\epsilon$ in Equation \@ref(eq:lm3).
+@smith_mendelian_2004 recognized that many genetic variants individually explain a 
+small proportion of the variation in a trait, and, thus, sufficient statistical 
+power would require sample sizes in the tens of thousands 
+[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
+resolved this issue by proposing use of multiple SNP instruments in a single MR 
+analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
+a collection of causal SNPs would have greater precision than an estimate derived 
+from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
+summary statistics was encouraged as a best practice [@nature2012asking]. 
+@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
+the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
+a strategy for determining a single causal effect from summary statistics for a 
+collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
+weighting" because each causal SNP's effect is weighted by the inverse of the variance 
+of its marginal (univariate) causal effect, and the overall causal effect is the sum of the weighted SNP
+causal effects.
 

### 2022-02-22 mr-review.Rmd
added ivw eq
minor updates to text

@@
 weighting" because each causal SNP's effect is weighted by the inverse of the variance 
-of its marginal (univariate) causal effect, and the overall causal effect is the sum of the weighted SNP
+of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
 causal effects.
+
+Specifically, @burgess_mendelian_2013 combined causal effect (ratio) estimates from $M$ independent SNPs by using Equation \@ref(eq:ivw).
+
+\begin{equation}
+  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
+  (\#eq:ivw)
+\end{equation}
+
+@burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
+analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core assumptions.
+
+@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted profile score to accomplish statistical inference in two-sample MR analyses with GWAS data.
 

### 2022-02-23 mr-review.Rmd
added gmm text

@@
 
-Many approaches to MR can be classified into two categories: two-stage methods and 
-likelihood-based methods. A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
+Approaches to MR can be classified into four categories: ratio of coefficients method, two-stage methods, 
+likelihood-based methods, and semiparametric methods.
+
+### Ratio of coefficients method
+
+The ratio of coefficients method, also known as the Wald method, estimates the 
+causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940].
+For a continuous outcome, the causal effect estimator, $\hat\gamma_{Wald}$ is defined in Equation \@ref(eq:wald).
+
+\begin{equation}
+  \hat\gamma_{Wald} = \frac{\hat\beta_{YZ}}{\hat\beta_{XZ}}
+  (\#eq:wald)
+\end{equation}
+
+In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope 
+estimates from the regressions of the outcome and exposure, respectively, on the 
+SNP instrument.
+Since the Wald method requires only the regression coefficients, it can be used with summary data.
+However, the Wald method doesn't accommodate multiple SNP instruments, which limits its use in the GWAS setting.
+
+### Two-stage methods
+
+A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
 outcome, one may perform "two-stage least squares", which involves two linear regressions. 
@@
 
+### Likelihood-based methods
+
 Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
 estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
-is the earliest approach for likelihood-based inference in MR [@davidson1993estimation]. CoMM and 
-PMR-Egger are two recent likelihood-based methods [@yang_comm_2019;@yuan_testing_2020]. We next discuss CoMM and CoMM-S2.
+is the earliest approach for likelihood-based inference in MR [@davidson1993estimation].
 
-CoMM, along with its extension for summary statistics, CoMM-S2, uses a collaborative mixed model to 
-relate exposure and outcome variables to the SNP genotypes [@yang_comm-s2_2020]. In this model, the 
-authors jointly model two linear mixed effects models, one relating the outcome to the SNPs and a second 
-tying the exposure to the SNPs. Model fitting then 
-proceeds via a parameter-expanded EM algorithm [@yang_comm_2019;@liu_parameter_1998]. @yang_comm_2019, 
-working with gene expression traits as exposures, found that CoMM better prioritized genetic variants 
-relative to other methods [@yang_comm_2019].
+Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations \@ref(eq:liml1) and \@ref(eq:liml2)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].
+
+\begin{equation}
+  x_i = \alpha_0 + \alpha_1 z_{i1} + \epsilon_{Xi}
+  (\#eq:liml1)
+\end{equation}
+
+\begin{equation}
+  y_i = \beta_0 + \beta_1 x_i + \epsilon_{Yi}
+  (\#eq:liml2)
+\end{equation}
+
+Limited information maximum likelihood is sometimes called the maximum likelihood 
+counterpart of two-stage least squares, and it yields the same causal estimate as 
+two-stage least squares and the ratio method when used with a single SNP 
+instrument. Additionally, the limited information maximum likelihood framework can accommodate $K$ SNP instruments by replacing $\alpha_1 z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.
+
+One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003].
+@burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation \@ref(eq:b1)).
+The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter, $\beta_1$ represents the effect between the true means for the exposure and the outcome.
+In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.
+
+\begin{eqnarray}
+\begin{pmatrix}
+X_i \\
+Y_i
+\end{pmatrix} \sim N_2\left(\begin{pmatrix} \xi_i\\
+\eta_i
+\end{pmatrix}, \Sigma
+\right)\nonumber \\
+\xi_i = \alpha_0 + \alpha_1 z_{i1}\nonumber \\
+\eta_i = \beta_0 + \beta_1 \xi_i
+  (\#eq:b1)
+\end{eqnarray}
+
+For each subject $i$, the exposure and outcome values come from a bivariate normal 
+distribution. They assumed that the mean of the exposure distribution is linear in 
+the SNP instrument and that the mean of the outcome distribution is a linear in the 
+mean exposure [@burgess_review_2017;@jones_choice_2012]. 
+@burgess_improving_2012 demonstrated that the Bayesian model, with weak 
+instruments, outperforms the limited information maximum likelihood model in terms 
+of frequentist coverage levels.
+
+### Semiparametric methods
+
+Semiparametric instrumental variable methods, which feature parametric and nonparametric components, 
+typically assume a parametric model connecting the exposure and outcome, but don't 
+impose distributional assumptions on the errors [@burgess_review_2017]. Compared to 
+fully parametric models, semiparametric models often are more robust to model 
+misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow 
+@burgess_review_2017 by discussing three semiparametric strategies, generalized 
+method of moments, continuous updating estimator, and G-estimation of structural 
+mean models.
+
+Generalized method of moments can be viewed as a more flexible form of two-stage 
+least squares that handles heteroscedastic errors and nonlinearity in the two 
+regressions [@burgess_review_2017]. Before specifying the model, we introduce 
+notation. $\mathbb{E}(Y|\text{do}(X = x))$ is the conditional expectation of $Y$ if 
+we forced $X$ to take value $x$ for every subject [@pearl2009causality].
+
+The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation \@ref(eq:gmm).
+
+\begin{eqnarray}
+  \sum_i (y_i - f(y_i;\beta)) = 0\nonumber \\
+  \sum_i z_i(y_i - f(y_i;\beta)) = 0
+  (\#eq:gmm)
+\end{eqnarray}
+
+The GMM estimate is the value of the vector $\beta$ that sastisfies Equation \@ref(eq:gmm), where $f(\mathbb{E}(Y|\text{do}(X = x))$
 

### 2022-02-23 mr-review.Rmd
g-estimation added

@@
 \begin{eqnarray}
-  \sum_i (y_i - f(y_i;\beta)) = 0\nonumber \\
-  \sum_i z_i(y_i - f(y_i;\beta)) = 0
+  \sum_i (y_i - f(x_i;\beta)) = 0\nonumber \\
+  \sum_i z_i(y_i - f(x_i;\beta)) = 0
   (\#eq:gmm)
@@
 
-The GMM estimate is the value of the vector $\beta$ that sastisfies Equation \@ref(eq:gmm), where $f(\mathbb{E}(Y|\text{do}(X = x))$
+The GMM estimate is the value of the vector $\beta$ that satisfies Equation \@ref(eq:gmm), where $f(x_i; \beta) = \mathbb{E}(Y|\text{do}(X = x))$.
+@palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations \@ref(eq:gmm).
+
+Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].  
+We follow @burgess_review_2017 by defining the potential outcome $Y(x)$ as the outcome value that we would have observed had we set the exposure value $X$ to $x$.
+For example, $Y(0)$ denotes the observed outcome had we set the exposure to zero instead of its observed value $x$.
+The structural mean model for a continuous outcome is displayed in Equation \@ref(eq:smm).
+
+\begin{equation}
+\mathbb{E}(Y(x) - Y(0) | X = x, Z = z) = \beta_1 x
+  (\#eq:smm)
+\end{equation}
+
+The causal effect parameter is $\beta_1$.
+@burgess_review_2017 derives the estimating equations for $\beta_1$, after noting that the conditional expectation, $\mathbb{E}(Y(0)| X = x, Z = z)$ is independent of $Z$ and reasoning that the causal effect is that value of $\beta_1$ that yields zero covariance between $Z$ and $\mathbb{E}(Y(0)| X = x, Z = z)$ (Equations \@ref(eq:smm1)).
+
+\begin{equation}
+\sum_i (z_{ik} - \bar z_{\cdot k})(y_i - \beta_1 x_i) = 0
+  (\#eq:smm1)
+\end{equation}
+
+Note that $k$ indexes SNP instruments and ranges from 1 to $K$.
+
+# Mendelian Randomization with multiple SNP instruments
+
+Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
+methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
+Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
+in MR studies. The solution to this task differs among published methods. Early methods chose a 
+set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
+SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
+proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
+this is the finding that including multiple correlated SNPs in transcriptome-wide association
+studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].
+
+## Statistical models and methods for MR with multiple instruments
+
+@smith_mendelian_2004 recognized that many genetic variants individually explain a 
+small proportion of the variation in a trait, and, thus, sufficient statistical 
+power would require sample sizes in the tens of thousands 
+[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
+resolved this issue by proposing use of multiple SNP instruments in a single MR 
+analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
+a collection of causal SNPs would have greater precision than an estimate derived 
+from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
+summary statistics was encouraged as a best practice [@nature2012asking]. 
+@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
+the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
+a strategy for determining a single causal effect from summary statistics for a 
+collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
+weighting" because each causal SNP's effect is weighted by the inverse of the variance 
+of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
+causal effects.
+
+Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from $M$ independent SNPs by using Equation \@ref(eq:ivw).
+
+\begin{equation}
+  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
+  (\#eq:ivw)
+\end{equation}
+
+@burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
+analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core assumptions.
+
+@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted profile score to accomplish statistical inference in two-sample MR analyses with GWAS data.
 
@@
 setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].
-
-# Mendelian Randomization with multiple SNP instruments
-
-Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
-methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
-Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
-in MR studies. The solution to this task differs among published methods. Early methods chose a 
-set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
-SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
-proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
-this is the finding that including multiple correlated SNPs in transcriptome-wide association
-studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].
-
-## Statistical models and methods for MR with multiple instruments
-
-@smith_mendelian_2004 recognized that many genetic variants individually explain a 
-small proportion of the variation in a trait, and, thus, sufficient statistical 
-power would require sample sizes in the tens of thousands 
-[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
-resolved this issue by proposing use of multiple SNP instruments in a single MR 
-analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
-a collection of causal SNPs would have greater precision than an estimate derived 
-from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
-summary statistics was encouraged as a best practice [@nature2012asking]. 
-@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
-the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
-a strategy for determining a single causal effect from summary statistics for a 
-collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
-weighting" because each causal SNP's effect is weighted by the inverse of the variance 
-of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
-causal effects.
-
-Specifically, @burgess_mendelian_2013 combined causal effect (ratio) estimates from $M$ independent SNPs by using Equation \@ref(eq:ivw).
-
-\begin{equation}
-  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
-  (\#eq:ivw)
-\end{equation}
-
-@burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
-analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core assumptions.
-
-@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted profile score to accomplish statistical inference in two-sample MR analyses with GWAS data.
 

### 2022-02-24 mr-review.Rmd
added placeholder for inSIDE assumption

@@
 
+InSIDE assumption
+
 ## Statistical models and methods for MR with one instrument
@@
 
-The causal effect parameter is $\beta_1$.
-@burgess_review_2017 derives the estimating equations for $\beta_1$, after noting that the conditional expectation, $\mathbb{E}(Y(0)| X = x, Z = z)$ is independent of $Z$ and reasoning that the causal effect is that value of $\beta_1$ that yields zero covariance between $Z$ and $\mathbb{E}(Y(0)| X = x, Z = z)$ (Equations \@ref(eq:smm1)).
+The causal effect parameter is $\beta_1$. @burgess_review_2017 derives the estimating equations for 
+$\beta_1$, after noting that the conditional expectation, $\mathbb{E}(Y(0)| X = x, Z = z)$ is 
+independent of $Z$ and reasoning that the causal effect is that value of $\beta_1$ that yields zero 
+covariance between $Z$ and $\mathbb{E}(Y(0)| X = x, Z = z)$ (Equations \@ref(eq:smm1)).
 
@@
 
-Note that $k$ indexes SNP instruments and ranges from 1 to $K$.
+Note that $k$ indexes SNP instruments and ranges from 1 to $K$ in Equation \@ref(eq:smm1).
 
@@
 
+@bowden_consistent_2016, noting that consistency with IVW MR is not robust to 
+invalid instrument use, developed a method for consistent MR analysis with  
+weighted median-based estimators. Building on research from @han2008detecting, 
+@bowden_consistent_2016 devised a weighted median-based estimator. Its breakdown point is 50\%, 
+meaning that up to 50\% of SNP instruments can be invalid while maintaining consistency in estimation 
+[@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate 
+statistic, from least to greatest. If an odd number, say $2K +1$, of SNP instruments is used, then 
+the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say 
+$2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio 
+estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.
+
+Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted 
+median estimator. To construct this estimator, order the ratio estimates from least to greatest, as 
+with the simple median estimator. The weighted median estimator, then is the median of a distribution
+defined by having $\tilde \beta_j$ as the $p_j = 100(\sum_{k = 1}^j - \frac{w_j}{2})$ percentile, 
+where $w_j$ is the weight assigned to the $j^{th}$ ordered ratio estimate. In this setting, the 
+simple median estimator is seen to be the weighted median estimator when all weights are equal.
+
+@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
+penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
+the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
+a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized 
+weighted median estimator
+
+weighted mode estimator
+
 @burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
@@
 @zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted profile score to accomplish statistical inference in two-sample MR analyses with GWAS data.
+
+### MR with allele scores
+
+Another approach to integrating multiple SNP instruments into a MR framework is
+through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
+calculate an unweighted score as the number of exposure-increasing alleles in 
+the subject's genotypes. They also calculate a weighted allele score by using 
+the exposure effects as weights. For example, a subject with $g_k$ copies of 
+exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{ik}$ 
+and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.
+
+@burgess_use_2013 found that when weights are obtained from external data or from 
+cross-validation or jackknife approaches applied to the analysis data, the allele score 
+functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
+two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
 

### 2022-02-24 mr-review.Rmd
found citation for mode-based MR
added text about hartwig 2017 - mode based estimation

@@
 collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
-weighting" because each causal SNP's effect is weighted by the inverse of the variance 
+weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
 of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
@@
 
+@greco_m_detecting_2015
+
+@bowden_mendelian_2015
+
 @bowden_consistent_2016, noting that consistency with IVW MR is not robust to 
 invalid instrument use, developed a method for consistent MR analysis with  
-weighted median-based estimators. Building on research from @han2008detecting, 
-@bowden_consistent_2016 devised a weighted median-based estimator. Its breakdown point is 50\%, 
+weighted median-based estimators, building on research from @han2008detecting. 
+Its breakdown point is 50\%, 
 meaning that up to 50\% of SNP instruments can be invalid while maintaining consistency in estimation 
@@
 
-weighted mode estimator
+@kang_instrumental_2016
+
+@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for MR analysis with GWAS summary data. 
+They first calculate Wald ratio estimates for every instrumental SNP. They then apply 
+smoothing to the empirical distribution of ratio estimates. The mode of this smoothed 
+distribution is the simple mode-based estimate of the causal effect 
+[@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by 
+weighting the empirical distribution of ratio estimates by the inverse variance of each 
+estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based 
+estimators with weighted and unweighted median-based estimators, MR Egger regression, and 
+IVW.
+They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings.
+The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017].
+The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017].
+In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.
+@hartwig_robust_2017
 

### 2022-02-24 mr-review.Rmd
more writing!

@@
 @burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
-analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core assumptions.
+analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core MR assumptions.
 
-@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted profile score to accomplish statistical inference in two-sample MR analyses with GWAS data.
+@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted 
+profile score to accomplish statistical inference in two-sample MR analyses with GWAS 
+summary data. Dividing pleiotropic effects into systematic and idiosyncratic, they model the 
+systematic pleiotropy with random effects. In so doing, no SNP satisfies the exclusion 
+restriction assumption. Adjusting their profile likelihood estimator from the setting 
+without pleiotropy, @zhao_statistical_2020 obtain an estimator with consistency and 
+asymptotic normality [@mccullagh_simple_1990]. Idiosyncratic pleiotropy is addressed through robustification of 
+the adjusted profile score [@huber_robust_1964]. They then demonstrate the properties of their estimator by 
+analyzing simulated and real data [@zhao_statistical_2020].
 
@@
 
+@hemani_orienting_2017
+
+@windmeijer_use_2019
+
 # Correlated horizontal pleiotropy
@@
 These observations motivated @yuan_likelihood_2021 to develop a new method, MRAID. MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
-the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref(mraid1) and \@ref(mraid2)).
+the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref(eq:mraid1) and \@ref(eq:mraid2)).
 
@@
 By encoding these assumptions in prior distributions, @yuan_likelihood_2021 enables the inferential procedures to accommodate the observed data in the context of the assumptions on SNPs.
+
+@brown_welch-weighted_2021
 
@@
 These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.
+
+## Multivariate exposures in MR analysis
+
+@grant_pleiotropy_2021: we introduce three methods which build on common approaches in the univariable setting: {MVMR}-Robust; {MVMR}-Median; and {MVMR}-Lasso
 

### 2022-02-25 mr-review.Rmd
citation links added

@@
 csl: elsevier-with-titles.csl
+link-citations: yes
+linkcolor: blue
 output: 
@@
 the exposure effects as weights. For example, a subject with $g_k$ copies of 
-exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{ik}$ 
+exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
 and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.
@@
 library(dagitty)
-library(ggdag)
 library(ggplot2)
@@
 
-@brown_welch-weighted_2021
+\begin{eqnarray}
+U = X\beta_U \odot Z_U + \epsilon_U\nonumber \\
+A = U\nu + B\delta + X\beta_A \odot Z_A + \epsilon_A \nonumber \\
+B = U\eta + A\gamma + X\beta_B \odot Z_B + \epsilon_B 
+  (\#eq:wwer)
+\end{eqnarray}
+
+@oconnor_distinguishing_2018
+
+In efforts to develop MR methods that are robust to correlated horizontal pleiotropy, 
+@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
+accommodates both uni-directional and bi-directional effects between exposure and outcome. 
+@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
+the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$.
+$Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to indicate which .
+Note that the model is agnostic with regards to which is the
+outcome and which the exposure, so we follow the authors' notation here.
+
+In Equations \@ref(eq:wwer),
+
+we use A and B to denote the observed traits in
+the study, and U to denote the unobserved genetic confounder.
+SNPs X affect each of A, B, and U with probabilities q, r, and s
+and effect sizes bA ; bB ; and bU sampled from a normal distribution
+with variances s2A ; s2B ; and s2U , respectively. The probability of ef-
+fect and variance of the sampling distribution combine to deter-
+mine the genetic architecture of each trait independently of the
+others. Finally, h and n specify the effect of the hidden confounder
+U on A and B, while g and d model the causal effect of A on B and B
+on A, respectively. Under this model, the phenotype values are
+given by
+U ¼ XbU +ZU þ εU ;(Equation 1)
+A ¼ Un þ Bd þ XbA +ZA þ εA ;(Equation 2)
+B ¼ Uh þ Ag þ XbB +ZB þ εB ;(Equation 3)
+where Z’s represent indicator variables that the SNP affects that
+trait, sampled as indicated above, + indicates vector element-
+wise (Hadamard) multiplication, and bolding represents matrices.
+The recursive nature of this model makes sampling from it non-
+trivial. In the supplemental methods, we describe how to simulate
+from this model and how to parameterize the model in terms of
+the heritability of each phenotype rather than the variance of
+the effect size distribution. We also explicitly describe how to set
+the parameters to mimic the models considered in O’Connor et
+al.5 and Morrison et al.3
 

### 2022-02-25 mr-review.Rmd
added MR egger

@@
 
-@greco_m_detecting_2015
+In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity Q test.
+In a meta-analysis of MR Wald estimates, one per SNP instrument, they calculate Q in Equation \@ref(eq:q).
 
-@bowden_mendelian_2015
+\begin{equation}
+Q = \sum_{k = 1}^K w_k(\hat\beta^{(k)}_{XY} - \mu_F)^2
+(\#eq:q)
+\end{equation}
+
+In Equation \@ref(eq:q), $w_k$ is the inverse variance of the Wald estimator, and $\mu_F = \frac{\sum_{k = 1}^K w_k \hat\beta^{(k)}_{XY}}{\sum_{k = 1}^K w_k}$.
+
+While the Q test tends to be conservative in small sample sizes, its power increases with increasing sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
+
+@bowden_mendelian_2015 adapted Egger regression, originally developed 
+to detect bias in meta-analyses, to detect bias from horizontal 
+pleiotropy in two-sample MR studies with GWAS summary data 
+[@egger1997bias]. The slope coefficient estimates the causal effect. 
+Under the assumption that the SNP-exposure relationship is independent 
+of the SNP's pleiotropic effects, Egger's test offers a valid test of 
+the null causal hypothesis and consistently estimates the causal 
+effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].
 

### 2022-02-25 mr-review.Rmd
deleted obsolete text

@@
 
-we use A and B to denote the observed traits in
-the study, and U to denote the unobserved genetic confounder.
-SNPs X affect each of A, B, and U with probabilities q, r, and s
-and effect sizes bA ; bB ; and bU sampled from a normal distribution
-with variances s2A ; s2B ; and s2U , respectively. The probability of ef-
-fect and variance of the sampling distribution combine to deter-
-mine the genetic architecture of each trait independently of the
-others. Finally, h and n specify the effect of the hidden confounder
-U on A and B, while g and d model the causal effect of A on B and B
-on A, respectively. Under this model, the phenotype values are
-given by
-U ¼ XbU +ZU þ εU ;(Equation 1)
-A ¼ Un þ Bd þ XbA +ZA þ εA ;(Equation 2)
-B ¼ Uh þ Ag þ XbB +ZB þ εB ;(Equation 3)
-where Z’s represent indicator variables that the SNP affects that
-trait, sampled as indicated above, + indicates vector element-
-wise (Hadamard) multiplication, and bolding represents matrices.
-The recursive nature of this model makes sampling from it non-
-trivial. In the supplemental methods, we describe how to simulate
-from this model and how to parameterize the model in terms of
-the heritability of each phenotype rather than the variance of
-the effect size distribution. We also explicitly describe how to set
-the parameters to mimic the models considered in O’Connor et
-al.5 and Morrison et al.3
-
 # Binary, Count, and Time-to-event outcomes and exposures

### 2022-02-25 mr-review.Rmd
added text about JASA papers

@@
 
-While the Q test tends to be conservative in small sample sizes, its power increases with increasing sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
+While the Q test tends to be conservative in small sample sizes, its power increases with increasing 
+sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
 
@@
 
-@kang_instrumental_2016
+@kang_instrumental_2016 developed a LASSO-based method, sisVIVE, to 
+identify invalid SNP instruments [@tibshirani1996regression]. sisVIVE's 
+advantage over earlier methods is that it doesn't require the analyst 
+to know which SNP instruments are valid. Instead, it requires that at 
+least 50\% of the instruments be valid. sisVIVE outperforms two-stage 
+least squares in many ways and performs similar to oracle two-stage 
+least squares. Simulations and data analysis results reveal that 
+sisVIVE is robust to possibly invalid instruments.
+
+@windmeijer_use_2019, building on the research from 
+@kang_instrumental_2016, implemented an adaptive LASSO-based estimator 
+after recognizing that sisVIVE misclassifies valid SNP instruments as 
+invalid when the invalid SNP instruments have strong effects on the 
+exposure. Consistent selection of invalid SNP instruments, they found, 
+depends on SNP instrument correlations. To address this issue, @windmeijer_use_2019 
+proposed a median estimator with consistency that doesn't 
+depend on SNP instrument-exposure association strength or the SNP instrument 
+correlation structure. They then applied methods from @zou2006adaptive to 
+achieve a consistent estimator with the same distribution as the oracle 
+two-stages least squares.
 
@@
 @windmeijer_use_2019
+@kang_instrumental_2016
 
@@
 
-In efforts to develop MR methods that are robust to correlated horizontal pleiotropy, 
+To develop MR methods that are robust to correlated horizontal pleiotropy, 
 @brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 

### 2022-02-26 mr-review.Rmd
rees_robust_2019 text added

@@
 correlation structure. They then applied methods from @zou2006adaptive to 
-achieve a consistent estimator with the same distribution as the oracle 
+achieve a consistent estimator with the same asymptotic distribution as the oracle 
 two-stages least squares.
 
-@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for MR analysis with GWAS summary data. 
+@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data. 
 They first calculate Wald ratio estimates for every instrumental SNP. They then apply 
@@
 In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.
-@hartwig_robust_2017
+
+@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
+developing three new methods for two-sample MR analysis with GWAS summary data: robust 
+regression, penalized weights, and LASSO penalization. The first two can be viewed as 
+modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
+downweighting or excluding variants with hetero-
+geneous causal estimates.
+Recall that MR-Egger offers consistent causal effect estimates when there is no 
+horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
+"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
+InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
+instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
+MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
+formulate an estimator with high breakdown point and high efficiency. The second estimation 
+method, with penalized weights, is much like the penalized median estimator, from above, 
+except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
+above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
+objective function from MR-Egger to get Equation \@ref(eq:lasso) as the needed objective function.
+
+\begin{equation}
+\sum_j [Var(\hat\beta_j)]^{-1}(\hat\beta_{Yj} - \theta_{0j} - \theta_1 \hat\beta_{1j})^2 + \lambda \sum_j |\theta_{0j}|
+(\#eq:lasso)
+\end{equation}
+
+@rees_robust_2019 studied two procedures for choosing $\lambda$: a heterogeneity stopping rule 
+and a cross-validation rule. Taken together, these three methods offer additional contributions to 
+a suite of sensitivity analysis tools that practitioners should consider.
 
 @burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
-analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core MR assumptions.
+analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or 
+more of the three core MR assumptions.
 
@@
 
-@oconnor_distinguishing_2018
-
 To develop MR methods that are robust to correlated horizontal pleiotropy, 

### 2022-02-27 mr-review.Rmd
more writing

@@
 more of the three core MR assumptions.
+Working in the setting of two-sample MR with GWAS data, they begin by considering a collection of candidate SNP instruments, which may contain some invalid SNP instruments.
+For each candidate SNP instrument, they calculate the Wald ratio estimate of the causal effect.
+With the collection of Wald ratio estimates, @burgess_robust_2020 reason that the set of valid SNP instruments have Wald ratio estimates that arise from a normal distribution centered on the true causal parameter with variance equal to that of the Wald ratio estimator.
+The invalid instruments, however, are assumed to follow a normal distribution centered at zero with a variance greater than that of the Wald ratio estimator.
+They then work with profile log likelihood over a grid of causal parameter values to make inferences.
 
@@
 analyzing simulated and real data [@zhao_statistical_2020].
+
+@foley_mr-clust_2021
+
+@tchetgen_tchetgen_genius_2021 MR-GENIUS
+
+@fang2021evaluating
+
+@sadreev2021navigating
 
@@
 @hemani_orienting_2017
-
-@windmeijer_use_2019
-@kang_instrumental_2016
 

### 2022-02-27 mr-review.Rmd
grapple text added

@@
 
+@bowden_improving_2019
+
+@wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong 
+instruments. Using GWAS summary statistics as inputs, GRAPPLE extends MR-RAPS and can detect presence of horizontally 
+pleiotropic pathways, infer the causal direction, and perform multivariable MR. Central to the 
+GRAPPLE framework is the observation that a horizontally pleiotropic pathway often gives rise to an 
+additional mode in the profile likelihood. GRAPPLE uses the presence of multiple profile likelihood 
+modes to diagnose horizontal pleiotropy with effects that are grouped by pathway. 
+GRAPPLE facilitates identification of SNP instruments that contribute to the 
+additional profile likelihood modes. A researcher using GRAPPLE may then recognize 
+a confounding factor for each mode. With additional GWAS summary data with traits 
+that reflect the confounding factors, GRAPPLE can fit multivariable MR models when
+the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].
+
 @foley_mr-clust_2021

### 2022-02-27 mr-review.Rmd
added text about MR GENIUS

@@
 the exposure complex trait, investigators can tease out causal effects from observational data, provided
-that three assumptions are satisfied. We discuss below the growth of genome-wide association studies 
+that necessary assumptions are satisfied. We discuss below the growth of genome-wide association studies 
 in parallel with development of Mendelian randomization methods.
@@
 traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
-enhancing our understanding of the biology of health and disease. Two-sample 
-Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
+enhancing our understanding of the biology of health and disease. Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
 genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
 inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].
+
 They are termed 
@@
 Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
-publication of the early GWAS (Figure \@ref(fig:article-counts)). Inverse variance 
+publication of the early GWAS (Figure \@ref(fig:article-counts)).
+Many studies use the "two-sample" MR design, in which one cohort of subjects has measurements for the exposure, while a second cohort has measurements for a second trait [@pierce_efficient_2013].
+Importantly, the SNP instruments are measured in both sets of subjects.
+
+Inverse variance 
 weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
@@
 independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
-selected through linkage disequilibrium (LD) clumping.
+selected through linkage disequilibrium (LD) clumping [@purcell2007plink].
 
@@
 
-InSIDE assumption
-
 ## Statistical models and methods for MR with one instrument
@@
 two-stage least squares and the ratio method when used with a single SNP 
-instrument. Additionally, the limited information maximum likelihood framework can accommodate $K$ SNP instruments by replacing $\alpha_1 z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.
+instrument. Additionally, the limited information maximum likelihood framework can accommodate $K > 1$ SNP instruments by replacing $\alpha_1 z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.
 
@@
 For each subject $i$, the exposure and outcome values come from a bivariate normal 
-distribution. They assumed that the mean of the exposure distribution is linear in 
-the SNP instrument and that the mean of the outcome distribution is a linear in the 
+distribution. The mean of the exposure distribution is assumed linear in 
+the SNP instrument, and the mean of the outcome distribution is linear in the 
 mean exposure [@burgess_review_2017;@jones_choice_2012]. 
@@
 Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].  
-We follow @burgess_review_2017 by defining the potential outcome $Y(x)$ as the outcome value that we would have observed had we set the exposure value $X$ to $x$.
-For example, $Y(0)$ denotes the observed outcome had we set the exposure to zero instead of its observed value $x$.
-The structural mean model for a continuous outcome is displayed in Equation \@ref(eq:smm).
+We follow @burgess_review_2017 by defining the potential outcome $Y(x)$ as the outcome value that we would 
+have observed had we set the exposure value $X$ to $x$. For example, $Y(0)$ denotes the observed outcome had 
+we set the exposure to zero instead of its observed value $x$. The structural mean model for a continuous 
+outcome is displayed in Equation \@ref(eq:smm).
 
@@
 Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
-in MR studies. The solution to this task differs among published methods. Early methods chose a 
+in these studies. The solution to this task differs among published methods. Early methods chose a 
 set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
@@
 summary statistics was encouraged as a best practice [@nature2012asking]. 
-@harbord_severity_2013 recognized that one could calculate the ratio estimate of 
+@harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of 
 the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
@@
 In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity Q test.
-In a meta-analysis of MR Wald estimates, one per SNP instrument, they calculate Q in Equation \@ref(eq:q).
+In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate Q in Equation \@ref(eq:q).
 
@@
 a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized 
-weighted median estimator
+weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald 
+ratio estimates.
 
@@
 achieve a consistent estimator with the same asymptotic distribution as the oracle 
-two-stages least squares.
+two-stages least squares. One important limitation of this thread of research is that
+both @kang_instrumental_2016 and @windmeijer_use_2019 require individual-level data.
 
@@
 estimators with weighted and unweighted median-based estimators, MR Egger regression, and 
-IVW.
-They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings.
-The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017].
-The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017].
-In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.
+IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than 
+other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to 
+detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of 
+MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect 
+when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In 
+this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the 
+other estimators.
 
@@
 modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
-downweighting or excluding variants with hetero-
-geneous causal estimates.
+downweighting or excluding variants with heterogeneous causal estimates.
 Recall that MR-Egger offers consistent causal effect estimates when there is no 
@@
 analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or 
-more of the three core MR assumptions.
-Working in the setting of two-sample MR with GWAS data, they begin by considering a collection of candidate SNP instruments, which may contain some invalid SNP instruments.
-For each candidate SNP instrument, they calculate the Wald ratio estimate of the causal effect.
-With the collection of Wald ratio estimates, @burgess_robust_2020 reason that the set of valid SNP instruments have Wald ratio estimates that arise from a normal distribution centered on the true causal parameter with variance equal to that of the Wald ratio estimator.
-The invalid instruments, however, are assumed to follow a normal distribution centered at zero with a variance greater than that of the Wald ratio estimator.
-They then work with profile log likelihood over a grid of causal parameter values to make inferences.
+more of the three core MR assumptions. Working in the setting of two-sample MR with GWAS data, they begin by 
+considering a collection of candidate SNP instruments, which may contain some invalid SNP instruments. For 
+each candidate SNP instrument, they calculate the Wald ratio estimate of the causal effect. With the 
+collection of Wald ratio estimates, @burgess_robust_2020 reason that the set of valid SNP instruments have 
+Wald ratio estimates that arise from a normal distribution centered on the true causal parameter with variance 
+equal to that of the Wald ratio estimator. The invalid instruments, however, are assumed to follow a normal 
+distribution centered at zero with a variance greater than that of the Wald ratio estimator. They then work 
+with profile log likelihood over a grid of causal parameter values to make inferences.
 
@@
 
-@bowden_improving_2019
+@bowden_improving_2019, working with a collection of SNP instruments, reasoned that if all three MR assumptions are satisfied, then the 
+collection of Wald ratio estimates should be homogeneous; thus, any departures from 
+homogeneity may signal a violation of one or more assumptions. They reported 
+modified weights, in addition to studying weights derived from first order and 
+second order approximations to the variance. Their modified weights lead 
+to an estimator that is useful for quantifying heterogeneity and detecting outliers.
 
@@
 
-@foley_mr-clust_2021
+@foley_mr-clust_2021 recognized that in some MR studies, the Wald ratio estimates for different
+SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms 
+and may identify the SNPs that are involved in them. They devised a method to detect 
+these clusters in the collection of Wald ratio estimates and found SNP instruments that 
+point to distinct causal pathways of biological importance [@foley_mr-clust_2021].
 
-@tchetgen_tchetgen_genius_2021 MR-GENIUS
+MR-GENIUS is a framework for two-sample MR analysis when individual-level data are available [@tchetgen_tchetgen_genius_2021]. @tchetgen_tchetgen_genius_2021 focused efforts on relaxing the 
+exclusion restriction assumption and building on existing G-estimation 
+methods to create an estimator that is robust to violations 
+of the exclusion restriction and to additive unmeasured confounding [@robins1994correcting].
+They observe that their estimator, in some settings, reduces to that of @lewbel2012using, which is widely used in econometrics, but not in the MR literature.
 

### 2022-02-27 mr-review.Rmd
added new section on overlapping samples

@@
 
-@fang2021evaluating
-
-@sadreev2021navigating
-
 ### MR with allele scores
@@
 
+## Sample overlap
+
+The issue of overlapping samples in the two-sample MR with GWAS summary data has recently 
+received attention in the published literature.
+
+@fang2021evaluating, working with data from the UK Biobank, found that SNP instruments 
+derived from overlapping samples explained a higher proportion of the variance compared to 
+those derived from non-overlapping samples. They surmise that block jackknife resampling MR
+enables causal inference while avoiding bias due to overlapping samples.
+
+@sadreev2021navigating analyzed 2,514 traits from the UK Biobank and evaluated the impact of winner's curse on MR analysis. Winner's curse, they found, amplifies weak instrument bias without 
+inflating the false discovery rate. This finding led the authors to design a pseudoreplication process that reduces bias in MR studies.
+
 # Applications of MR in GWAS

### 2022-02-28 mr-review.Rmd
fixed section title
minor comment
minor edits
tweaks

@@
 ---
-title: "Mendelian randomization for causal inference in Genome-wide association studies"
+title: "Mendelian randomization for causal inference in genome-wide association studies"
 author: "Frederick J. Boehm & Xiang Zhou"
@@
 
-They are termed 
-"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
-two non-overlapping cohorts of subjects [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.
-
 Genome-wide association studies (GWAS) present unique opportunities for 
@@
 Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
-publication of the early GWAS (Figure \@ref(fig:article-counts)).
-Many studies use the "two-sample" MR design, in which one cohort of subjects has measurements for the exposure, while a second cohort has measurements for a second trait [@pierce_efficient_2013].
-Importantly, the SNP instruments are measured in both sets of subjects.
+publication of the early GWAS (Figure \@ref(fig:article-counts)). Many studies use the "two-sample" MR 
+design, in which one cohort of subjects has measurements for the exposure, while a second cohort has 
+measurements for a second trait [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in 
+both sets of subjects.
 
-Inverse variance 
-weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
+Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
 median estimation [@bowden_consistent_2016], Bayesian weighted MR 
@@
 
-Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
+Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that 
+bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
 the standard MR assumptions and can lead to biased 
@@
 Since the Wald method requires only the regression coefficients, it can be used with summary data.
-However, the Wald method doesn't accommodate multiple SNP instruments, which limits its use in the GWAS setting.
+However, the Wald method doesn't accommodate multiple SNP instruments, which limits its direct use in the GWAS setting.
 
@@
 
-Note that $k$ indexes SNP instruments and ranges from 1 to $K$ in Equation \@ref(eq:smm1).
+Note that $k$ indexes SNP instruments and ranges from $1$ to $K$ in Equation \@ref(eq:smm1).
 
@@
 
-In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity Q test.
-In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate Q in Equation \@ref(eq:q).
+In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity $Q$ test.
+In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).
 
@@
 
-While the Q test tends to be conservative in small sample sizes, its power increases with increasing 
+While the $Q$ test tends to be conservative in small sample sizes, its power increases with increasing 
 sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
@@
 pleiotropy in two-sample MR studies with GWAS summary data 
-[@egger1997bias]. The slope coefficient estimates the causal effect. 
+[@egger1997bias]. The slope coefficient estimates the causal effect of the exposure on the outcome. 
 Under the assumption that the SNP-exposure relationship is independent 
@@
 Note that the model is agnostic with regards to which is the
-outcome and which the exposure, so we follow the authors' notation here.
+outcome and which the exposure, so we follow the authors' notation here. The symbol $\odot$ denotes the Hadamard elementwise product.
 
@@
 
-## Sample overlap
+## Sample overlap in biobanks
 

### 2022-02-28 mr-review.Rmd
added refs & section header for bidirectional MR

@@
 derived from overlapping samples explained a higher proportion of the variance compared to 
-those derived from non-overlapping samples. They surmise that block jackknife resampling MR
+those derived from non-overlapping samples. They argue that block jackknife resampling MR
 enables causal inference while avoiding bias due to overlapping samples.
@@
 inflating the false discovery rate. This finding led the authors to design a pseudoreplication process that reduces bias in MR studies.
+
+@mounier_bias_2021
+@hartwig_two-sample_2016
+@lawlor_commentary_2016
+@burgess_bias_2016
+
+## Bi-directional causal inference with MR
+
+@darrous_simultaneous_2021
 

### 2022-03-01 mr-review.Rmd
organized bidirectional papers into one section
added text for MR TRYX

@@
 above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
-objective function from MR-Egger to get Equation \@ref(eq:lasso) as the needed objective function.
+objective function from MR-Egger to get Equation \@ref(eq:lasso) as the objective function.
 
@@
 
-@wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong 
-instruments. Using GWAS summary statistics as inputs, GRAPPLE extends MR-RAPS and can detect presence of horizontally 
-pleiotropic pathways, infer the causal direction, and perform multivariable MR. Central to the 
-GRAPPLE framework is the observation that a horizontally pleiotropic pathway often gives rise to an 
-additional mode in the profile likelihood. GRAPPLE uses the presence of multiple profile likelihood 
-modes to diagnose horizontal pleiotropy with effects that are grouped by pathway. 
-GRAPPLE facilitates identification of SNP instruments that contribute to the 
-additional profile likelihood modes. A researcher using GRAPPLE may then recognize 
-a confounding factor for each mode. With additional GWAS summary data with traits 
-that reflect the confounding factors, GRAPPLE can fit multivariable MR models when
-the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].
+@cho_exploiting_2020 developed the MR-TRYX framework for exploiting horizontal pleiotropy
+to identify alternative causal pathways. They did this by positing that, for a 
+collection of putative SNP instruments, any heterogeneity in the Wald ratio estimates
+may be due to horizontal pleiotropy. Once they identify a SNP with outlying 
+ratio estimates, they search GWAS results by querying the MR-Base database to identify complex traits (in other studies) that 
+associate with the outlying SNP [@hemani_mr-base_2018]. They follow up promising trait associations to 
+account for their observations with multiple causal pathways.
 
@@
 
-\begin{eqnarray}
-U = X\beta_U \odot Z_U + \epsilon_U\nonumber \\
-A = U\nu + B\delta + X\beta_A \odot Z_A + \epsilon_A \nonumber \\
-B = U\eta + A\gamma + X\beta_B \odot Z_B + \epsilon_B 
-  (\#eq:wwer)
-\end{eqnarray}
-
-To develop MR methods that are robust to correlated horizontal pleiotropy, 
-@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
-accommodates both uni-directional and bi-directional effects between exposure and outcome. 
-@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
-the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$.
-$Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to indicate which .
-Note that the model is agnostic with regards to which is the
-outcome and which the exposure, so we follow the authors' notation here. The symbol $\odot$ denotes the Hadamard elementwise product.
-
-In Equations \@ref(eq:wwer),
-
 # Binary, Count, and Time-to-event outcomes and exposures
@@
 @darrous_simultaneous_2021
+@wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong 
+instruments. Using GWAS summary statistics as inputs, GRAPPLE extends MR-RAPS and can detect presence of horizontally 
+pleiotropic pathways, infer the causal direction, and perform multivariable MR. Central to the 
+GRAPPLE framework is the observation that a horizontally pleiotropic pathway often gives rise to an 
+additional mode in the profile likelihood. GRAPPLE uses the presence of multiple profile likelihood 
+modes to diagnose horizontal pleiotropy with effects that are grouped by pathway. 
+GRAPPLE facilitates identification of SNP instruments that contribute to the 
+additional profile likelihood modes. A researcher using GRAPPLE may then recognize 
+a confounding factor for each mode. With additional GWAS summary data with traits 
+that reflect the confounding factors, GRAPPLE can fit multivariable MR models when
+the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].
+
+\begin{eqnarray}
+U = X\beta_U \odot Z_U + \epsilon_U\nonumber \\
+A = U\nu + B\delta + X\beta_A \odot Z_A + \epsilon_A \nonumber \\
+B = U\eta + A\gamma + X\beta_B \odot Z_B + \epsilon_B 
+  (\#eq:wwer)
+\end{eqnarray}
+
+To develop MR methods that are robust to correlated horizontal pleiotropy, 
+@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
+accommodates both uni-directional and bi-directional effects between exposure and outcome. 
+@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
+the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$.
+$Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to indicate which .
+Note that the model is agnostic with regards to which is the
+outcome and which the exposure, so we follow the authors' notation here. The symbol $\odot$ denotes the Hadamard elementwise product.
+
+In Equations \@ref(eq:wwer),
 

### 2022-03-01 mr-review.Rmd
added text for hemani 2017

@@
 
-@hemani_orienting_2017
+@hemani_orienting_2017, working in the setting of two-sample MR with summary GWAS data, recognized that
+measurement error in outcome and exposure can give misleading results with the inferred 
+causal arrow in the wrong direction. To work around this issue, they extended 
+MR with a method that infers the causal direction between two traits. Specifically, @hemani_orienting_2017 
+adapt Steiger's Z test for a difference in correlations [@steiger1980tests] to assess which of the two 
+traits in an analysis has the stronger correlation with the SNP instrument. That 
+with the stronger correlation is treated as the exposure. @hemani_orienting_2017 
+applied their test in a study of DNA methylation and gene expression, where either 
+direction of causality is plausible, and found 
+that many methylation traits cause changes in gene expression.
 

### 2022-03-01 mr-review.Rmd
added text  for berzuini et al (Bayesian Mimr)

@@
 
+@berzuini_bayesian_2020 proposed a Bayesian approach to MR where SNP instruments
+are permitted to exhibit pleiotropic effects on the outcome. Working in the setting of 
+one-sample MR with individual-level data, they construct their 
+model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects. Their 
+method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP 
+instruments) and permits correlated SNP instruments, unlike many frequentist methods. Posterior inference 
+proceeds by Markov chain monte carlo sampling. They also elaborate their model for 
+a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.
+
 # Correlated horizontal pleiotropy

### 2022-03-01 mr-review.Rmd
added text for grant_efficient_2020 regularization method

@@
 
+@cheng_mr-ldp_2020 sought a probabilistic model that accounts for both linkage disequilibrium among SNP 
+instruments and horizontal pleiotropy. They termed their method MR-LDP. Using an approach from @zhu_bayesian_2017, @cheng_mr-ldp_2020 derived 
+an approximate likelihood for their regression models of the exposure and outcome on the SNP instruments. 
+Incorporation of a random effect into their models accounted for the variance in causal effect estimates due 
+to horizontal pleiotropy. They situate inference within an empirical Bayesian framework and present a 
+parameter-expanded variational Bayes expectation-maximization algorithm for estimation [@liu1998parameter]. 
+By jointly modeling the distribution of GWAS summary statistics and causal effects, @cheng_mr-ldp_2020 
+efficiently accommodates multiple SNP instruments in linkage disequilibrium. One limitation of MR-LDP  
+is its reliance on the InSIDE assumption and resulting inability to accommodate correlated horizontal pleiotropy.
+
+@zhu_iterative_2021 present IMRP, a method for causal effect estimation and horizontal pleiotropy detection. 
+They describe a test to distinguish vertical pleiotropy from horizontal pleiotropy. By combining this test 
+with two-sample MR analysis of GWAS summary statistics, they develop an iterative procedure for estimating 
+the causal effect of exposure on outcome and test for presence of horizontal pleiotropy in the SNP 
+instruments. The iterative procedure alternates between 1) updating the estimate of the causal effect and 2) 
+testing for horizontal pleiotropy with the included SNP instruments. At each test for horizontal pleiotropy, 
+some SNP instruments may be discarded if they demonstrate evidence of horizontal pleiotropic effects. The 
+procedure ends when there remain only SNP instruments that don't exhibit horizontal pleiotropy. In the authors' 
+simulations and data analysis sections, their new method performs about as well as GSMR, IVW, and MR-PRESSO in the 
+settings with and without balanced pleiotropy and with and without satisfaction of the InSIDE assumption [@zhu_iterative_2021].
+Finally, the authors note that IMRP is three orders of magnitude faster than simulations-based MR-PRESSO.
+
+@grant_efficient_2020 build on research from @lin2015regularization to present a method that 
+applies a $\ell_1$ penalty to coefficients of covariates in two-sample MR with GWAS summary 
+statistics. Importantly, their method doesn't penalize the SNP-exposure association. Instead, 
+it shrinks towards zero those coefficients for covariates that demonstrate little pleiotropy. In so doing, 
+they extend work from @burgess_multivariable_2015 and @burgess_re_2015 to get a causal effect estimator that needs no valid SNP instruments and robust to measured vertical pleiotropy and more statistically efficient
+than previous MR methods.
+
 # Correlated horizontal pleiotropy

### 2022-03-01 mr-review.Rmd
mrcip text added

@@
 estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
-is the earliest approach for likelihood-based inference in MR [@davidson1993estimation].
+is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite].
 
@@
 
+@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods. Central to their
+work is the proposal to create a new instrumental variable as a linear combination of genotype data from 
+a collection of SNP instruments. They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically  
+pleiotropic traits. For a collection of $K$ SNPs, they choose the $K$-vector that meets these criteria and 
+maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a 
+$\ell_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation 
+of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. 
+@jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are 
+unbiased in their simulation settings.
+
+@xu_mrcip_2021 consider two classes of horizontal pleiotropy with their method, MRCIP. They partition horizontal 
+pleiotropic effects into correlated or idiosyncratic horizontal pleiotropy. Idiosyncratic horizontal pleiotropy 
+is that with a large effect size, while correlated horizontal pleiotropy, like its definition elsewhere in our 
+manuscript, refers to correlation between the effects of the SNP instruments on the exposure and the effects of 
+the SNP instruments on the outcome. MRCIP relaxes both the exclusion restriction assumption and the InSIDE 
+assumption by directly modeling the correlated horizontal pleiotropy with random effects. Furthermore, to 
+accommodate the idiosyncratic pleiotropy, the authors downweight SNP instruments that demonstrate strong direct 
+effects on the outcome. The find that their MRCIP provides valid causal inference even when there are no valid 
+SNP instruments (*i.e.*, when all SNP instruments are invalid) and when the InSIDE assumption is violated 
+[@xu_mrcip_2021].
+Their PRW-EM algorithm adds a reweighting step to the traditional expectation maximization algorithm to estimate the causal effect parameter in MR analysis [@dempster1977maximum].
+Finally, they compare  MRCIP to other MR methods with both simulated and real data sets.
+@xu_mrcip_2021, like the developers of MR-RAPS, assume that the horizontal pleiotropy is balanced, *i.e.*, that the mean horizontal pleiotropic effect is zero.
+One direction for future study is to augment MRCIP to permit unbalanced horizontal pleiotropy.
+
 # Correlated horizontal pleiotropy

### 2022-03-01 mr-review.Rmd
added text about MR-GxE

@@
 
+@iong_latent_2020 build on other research using mixture models in the two-sample MR setting with GWAS summary 
+statistics. Like MR-Clust, MR-PATH, as the method from @iong_latent_2020 is called, works with the Wald ratio 
+estimates from multiple SNP instruments. Coining the term "mechanistic heterogeneity" to refer to the 
+distribution of Wald ratio estimates due to SNP involvement in distinct causal pathways, @iong_latent_2020 
+specify a $J$-component normal mixture model. They use an expectation maximization algorithm to fit the model 
+parameters (the component weights and the parameters for each of the $J$ normal distributions) and suggest a 
+Bayesian information criterion-based approach to choosing $J$ [@schwarz1978estimating].
+
+@wang_weakinstrument_2021 propose causal effect test statistics that are robust under weak instrument asymptotics 
+[@staiger1994instrumental]. Specifically, they extend three econometric methods to two-sample MR with GWAS 
+summary statistics [@anderson1949estimation;@kleibergen_pivotal_2002;@moreira_conditional_2003] and study the 
+theoretical properties and the performance in applications. @wang_weakinstrument_2021 extend their methods to 
+demonstrate conditions under which their estimators are equivalent to those in works from @zhao_statistical_2020, @windmeijer_use_2019, and @bowden_improving_2019.
+
+@spiller_detecting_2018 and @spiller_interaction-based_2020 present a method, MR-GxE, that exploits gene by 
+environment interactions to detect and to adjust for bias due to horizontal pleiotropy. They leverage the insight 
+that a SNP instrument-covariate interaction that induces differences in SNP instrument-exposure associations 
+@spiller_interaction-based_2020 treat a SNP-covariate interaction as an instrumental variable, and, by so doing, they impose assumptions on that interaction.
+Together, these assumptions require that MR-GxE assume that horizontal pleiotropy effects are constant across the study cohort.
+The authors devise a three-step inferential procedure for two-sample MR with GWAS summary statistics:
+
+1. Estimate the SNP-exposure and SNP-outcome associations at a range of covariate values  
+1. Regress the SNP-outcome estimated associations on the SNP-exposure estimated associations  
+1. Treat the resulting slope as the causal effect estimate and the intercept as the mean horizontal pleiotropy effect.
+
+@spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set 
+of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant 
+pleiotropy assumption.
+
 # Correlated horizontal pleiotropy

### 2022-03-02 mr-review.Rmd
BayesMR text added

@@
 Genome-wide association studies (GWAS) present unique opportunities for 
-discovery of causal relationships via MR. A GWAS
+discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS
 interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
@@
 
+@hemani_automating_2017
+
 # Horizontal pleiotropy
@@
 pleiotropy assumption.
+
+@bucur_inferring_2020 developed BayesMR, a Bayesian framework for two-sample MR with individual-level data and discuss an approximation for use with  
+GWAS summary statistics. Unlike many MR methods, BayesMR both aims to control for 
+horizontal pleiotropy and to quantitatively assess the possibility of reverse 
+causation, where the putative exposure is actually caused by the putative outcome. 
+For $J$ independent SNP instruments, @bucur_inferring_2020 permit each SNP instrument to 
+have both a direct effect on the exposure and a horizontally pleiotropic direct effect 
+on the outcome. Instrument effects and horizontally pleiotropic effects are assumed 
+independent. Note that this is the Bayesian analog of the InSIDE assumption. 
+After specifying the model, they present a nested sampling scheme [@skilling2006nested;@handley2015polychord1;@handley2015polychord] that simultaneously 
+computes the model evidence for both causal directions (*i.e.*, exposure causes 
+outcome and outcome causes exposure) and acquires samples from the posterior 
+distribution for Bayesian statistical inferences.
 

### 2022-03-02 mr-review.Rmd
added text about MVMR

@@
 
+@schmidt_mendelian_2018 present BMRE, a Bayesian implementation of 
+MR Egger. They apply weakly informative priors to the horizontal pleiotropy effect 
+estimator in efforts to increase statistical power to detect a nonzero 
+slope, i.e., the coefficient of the SNP instrument. They develop their method for two-sample MR with individual-level 
+data. Simulations demonstrate that BMRE outperforms the MR Egger method while maintaining type I error rates in the examined scenarios. The authors emphasize a potential role for BMRE in 
+MR sensitivity analysis.
+
+@howey_bayesian_2020 use anchored Bayesian networks to
+
+@sanderson_examination_2019 found that multivariable MR, where multiple exposures jointly 
+cause an outcome, consistently estimates the effect of the exposures on the outcome. They 
+examined multivariable MR with both individual-level data and GWAS summary statistics. 
+@sanderson_examination_2019 also develop a generalized version of Cochran's Q test for 
+quantifying SNP instrument strength and validity. However, this test requires knowledge of 
+the covariances between the effects of the SNP instruments on the exposures. While these 
+covariances could be estimated from individual-level data, such data are often unavailable in 
+the GWAS setting. Importantly, the authors note that the estimands for multivariable MR nad MR 
+differ; multivariable MR estimates the direct causal effects of each exposure on the outcome, 
+while MR estimates the total causal effect of an exposure on the outcome.
+
+@van_der_graaf_mendelian_2020
+
 # Correlated horizontal pleiotropy

### 2022-03-02 mr-review.Rmd
added sentences about 3 more methods

@@
 
-@van_der_graaf_mendelian_2020
+MR-link, reported by @van_der_graaf_mendelian_2020, aims to account for unobserved pleiotropy and 
+linkage disequilibrium by using individual-level data on the outcome variable and summary statistics for the exposure. MR-link uses a three-step procedure to estimate causal effects.
+
+1. Use GCTA-COJO to obtain SNP instruments [@genetic_investigation_of_anthropometric_traits_giant_consortium_conditional_2012]  
+1. Prune all SNPs in linkage disequilibrium with the SNP instruments to obtain a set of "tag" SNPs 
+with correlations less than 0.95  
+1. Solve for the causal effect estimate with ridge regression of the outcome on the matrix resulting from concatenation of the SNP instrument genotypes matrix and the tag SNPs genotypes matrix.
+
+Note that in step 3, ridge regression is preferred to ordinary least squares due to the limited 
+statistical power of ordinary least squares in this setting. Presumably, this is due to collinearity in 
+the matrix of tag SNP genotypes.
+
+@jiang_hierarchical_2021, building on research from @newcombe2016jam, present a two-stage hierarchical Bayesian 
+model that aims to unify MR and transcriptome-wide association studies frameworks.
+
+Genetic instrumental variable regression uses GWAS summary statistics to construct polygenic scores for 
+the outcome of interest [@diprete_genetic_2018]. In constructing polygenic scores from subsets of a GWAS 
+cohort, @diprete_genetic_2018 aim to reduce bias due to horizontal pleiotropy.
 

### 2022-03-03 mr-review.Rmd
whitespace added
moved stuff around

@@
 
+# Assumption violations and Sensitivity Analysis
+
+It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations.
+Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
+must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
+a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
+if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
+possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
+affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].
+
+```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X.
+Because W affects Y through a pathway that doesn't involve exposure X, this scenario violates the exclusion restriction assumption."}
+
+library(dagitty)
+library(ggplot2)
+
+d1 <- dagitty('dag {
+bb="0,0,1,1"
+G [pos="0,0"]
+W [pos="0.5,0"]
+X [exposure, pos="1.0,0"]
+Y [outcome, pos="1.5,0"]
+G -> W
+W -> X
+W -> Y [pos="1,1"]
+X -> Y
+}')
+
+plot(d1)
+
+```
+
+```{r, dag2, fig.cap = "Scenario where the genetic variant G affects exposure X, which affects Y. Unobserved confounder U affects both X and Y."}
+d1 <- dagitty('dag {
+bb="0,0,1,1"
+G [pos="0.0,0.0"]
+U [pos="0.75,0.5"]
+X [exposure, pos="0.5,0.0"]
+Y [outcome, pos="1.0,0.0"]
+G -> X
+U -> X
+U -> Y
+X -> Y
+}'
+)
+plot(d1)
+
+```
+
+Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that 
+enable assumption relaxations and has motivated use of sensitivity analysis to quantify the impact of 
+possible assumption violations.
+
+Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
+the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
+into how the results might differ if the assumptions be violated.
+Because not all confounding 
+variables are known or measured, assumptions 2 and 3 are not fully verifiable 
+in MR studies. Recognition of this fact has inspired the development of 
+sensitivity analysis tools. While we can't assess whether a SNP instrument is 
+associated with every confounder of the exposure-outcome association, it is 
+possible to examine the associations of the SNP instrument with the measured 
+covariates. While absence of such associations doesn't guarantee satisfaction 
+of the assumption, presence of SNP-covariate associations must be investigated
+carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].
+
+@burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when 
+working with GWAS summary statistics. @burgess_sensitivity_2017 presented methods for both assessing the
+MR assumptions, to the extent possible, and performing robust analyses. For example, 
+they used measured covariates to assess for possible associations with the SNP 
+instruments. While they can't rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.
+
+To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of 
+C-reactive protein (CRP) levels on coronary artery disease risk with four genetic 
+variants in the CRP gene region and 17 other genetic variants that affect coronary 
+artery disease risk. They used measured covariates to probe for SNP-covariate 
+associations, and followed it up with scatter plots and Cochran's Q test on the causal 
+estimates to inquire about whether the 
+SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003]. 
+Additionally, @burgess_sensitivity_2017 suggested using a funnel plot, like those in the 
+meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011].
+Additionally, Egger regression can be useful in this
+setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].
+
 ## Statistical models and methods for MR with multiple instruments
+
+One approach to integrating multiple SNP instruments into a MR framework is
+through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
+calculate an unweighted score as the number of exposure-increasing alleles in 
+the subject's genotypes. They also calculate a weighted allele score by using 
+the exposure effects as weights. For example, a subject with $g_k$ copies of 
+exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
+and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.
+
+@burgess_use_2013 found that when weights are obtained from external data or from 
+cross-validation or jackknife approaches applied to the analysis data, the allele score 
+functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
+two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
 
@@
 They observe that their estimator, in some settings, reduces to that of @lewbel2012using, which is widely used in econometrics, but not in the MR literature.
-
-### MR with allele scores
-
-Another approach to integrating multiple SNP instruments into a MR framework is
-through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
-calculate an unweighted score as the number of exposure-increasing alleles in 
-the subject's genotypes. They also calculate a weighted allele score by using 
-the exposure effects as weights. For example, a subject with $g_k$ copies of 
-exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
-and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.
-
-@burgess_use_2013 found that when weights are obtained from external data or from 
-cross-validation or jackknife approaches applied to the analysis data, the allele score 
-functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
-two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
-
-# Assumption violations
-
-It is worthwhile to consider ways in which the standard MR assumptions may be violated. 
-Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
-must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
-a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
-if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
-possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
-affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].
-
-```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X.
-Because W affects Y through a pathway that doesn't involve exposure X, this scenario violates the exclusion restriction assumption."}
-
-library(dagitty)
-library(ggplot2)
-
-d1 <- dagitty('dag {
-bb="0,0,1,1"
-G [pos="0,0"]
-W [pos="0.5,0"]
-X [exposure, pos="1.0,0"]
-Y [outcome, pos="1.5,0"]
-G -> W
-W -> X
-W -> Y [pos="1,1"]
-X -> Y
-}')
-
-plot(d1)
-
-```
-
-```{r, dag2, fig.cap = "Scenario where the genetic variant G affects exposure X, which affects Y. Unobserved confounder U affects both X and Y."}
-d1 <- dagitty('dag {
-bb="0,0,1,1"
-G [pos="0.0,0.0"]
-U [pos="0.75,0.5"]
-X [exposure, pos="0.5,0.0"]
-Y [outcome, pos="1.0,0.0"]
-G -> X
-U -> X
-U -> Y
-X -> Y
-}'
-)
-plot(d1)
-
-```
-
-Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that 
-enable assumption relaxations and has motivated use of sensitivity analysis to quantify the impact of 
-possible assumption violations.
-
-# Sensitivity analysis
-
-Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
-the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
-into how the results might differ if the assumptions be violated.
-Because not all confounding 
-variables are known or measured, assumptions 2 and 3 are not fully verifiable 
-in MR studies. Recognition of this fact has inspired the development of 
-sensitivity analysis tools. While we can't assess whether a SNP instrument is 
-associated with every confounder of the exposure-outcome association, it is 
-possible to examine the associations of the SNP instrument with the measured 
-covariates. While absence of such associations doesn't guarantee satisfaction 
-of the assumption, presence of SNP-covariate associations must be investigated
-carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].
-
-@burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when 
-working with GWAS summary statistics. @burgess_sensitivity_2017 presented methods for both assessing the
-MR assumptions, to the extent possible, and performing robust analyses. For example, 
-they used measured covariates to assess for possible associations with the SNP 
-instruments. While they can't rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.
-
-To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of 
-C-reactive protein (CRP) levels on coronary artery disease risk with four genetic 
-variants in the CRP gene region and 17 other genetic variants that affect coronary 
-artery disease risk. They used measured covariates to probe for SNP-covariate 
-associations, and followed it up with scatter plots and Cochran's Q test on the causal 
-estimates to inquire about whether the 
-SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003]. 
-Additionally, @burgess_sensitivity_2017 suggested using a funnel plot, like those in the 
-meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011].
-Additionally, Egger regression can be useful in this
-setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].
 
@@
 @darrous_simultaneous_2021
+
 @wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong 

### 2022-03-04 mr-review.Rmd
updated yaml & started method.xlsx

@@
 linkcolor: blue
+always_allow_html: true
 output: 

### 2022-03-04 mr-review.Rmd
started omitted.Rmd for removed refs

@@
 thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].
-Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and refined our understanding of
+Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of
 human health and disease.
@@
 
+A number of excellent MR reviews have appeared in the literature [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022]. Our review doesn't 
+aim to replace them, but seeks to complement their findings. 
+We present recent methods developments for an audience of statisticians and epidemiologists. We discuss in detail methods advances before briefly summarizing 
+applications and directions for future research. The value of MR analysis with 
+GWAS data will continue to grow in parallel with statistical methods advances.
+
 Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
@@
 
-Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
-median estimation [@bowden_consistent_2016], Bayesian weighted MR 
-[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
-[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
-[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
-their approaches to two challenges:
-
-1. instrument selection (from among the available SNPs)  
-1. modeling and controlling for horizontal pleiotropy
-
-Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
-be statistically independent. This restriction to 
-independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
-selected through linkage disequilibrium (LD) clumping [@purcell2007plink].
-
-However, LD clumping for choosing instrument SNPs can be problematic for several 
-reasons. First, the chosen SNPs may not be causal for the outcome. This can 
-happen, for instance, by using a noncausal SNP that is in high LD with a 
-causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
-causal SNPs may be in high LD together. LD clumping is likely to omit one or more
-of the causal SNPs, which can diminish statistical power in MR analysis.
-
-Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that 
-bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
-the standard MR assumptions and can lead to biased 
-causal effect estimates and diminished statistical power. We discuss approaches to 
-modeling and accounting for horizontal pleiotropy below.
-
 Recent methods developments have enabled execution of two-sample MR studies with only 
@@
 ability to use published data has contributed to a rapid increase in the number of
-two-sample MR methods for GWAS summary statistics.
+two-sample MR methods for GWAS summary statistics (Figure \@ref(fig:article-counts)).
 
@@
 practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
-way to assess robustness of estimates in the presence of possible assumption violations.
+way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014].
 
@@
 
-Approaches to MR can be classified into four categories: ratio of coefficients method, two-stage methods, 
-likelihood-based methods, and semiparametric methods.
+Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, 
+likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods 
+that
 
@@
 estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
-is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite].
+is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. We consider it below before introducing a Bayesian method.
 
@@
 Note that $k$ indexes SNP instruments and ranges from $1$ to $K$ in Equation \@ref(eq:smm1).
-
-# Mendelian Randomization with multiple SNP instruments
-
-Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
-methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
-Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
-in these studies. The solution to this task differs among published methods. Early methods chose a 
-set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
-SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
-proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
-this is the finding that including multiple correlated SNPs in transcriptome-wide association
-studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].
 
@@
 
-## Statistical models and methods for MR with multiple instruments
-
-One approach to integrating multiple SNP instruments into a MR framework is
-through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
-calculate an unweighted score as the number of exposure-increasing alleles in 
-the subject's genotypes. They also calculate a weighted allele score by using 
-the exposure effects as weights. For example, a subject with $g_k$ copies of 
-exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
-and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.
-
-@burgess_use_2013 found that when weights are obtained from external data or from 
-cross-validation or jackknife approaches applied to the analysis data, the allele score 
-functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
-two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
+# Mendelian Randomization with multiple SNP instruments
+
+Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
+methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
+Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
+in these studies. The solution to this task differs among published methods. Early methods chose a 
+set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
+SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
+proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
+this is the finding that including multiple correlated SNPs in transcriptome-wide association
+studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].
+
+Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
+median estimation [@bowden_consistent_2016], Bayesian weighted MR 
+[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
+[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
+[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
+their approaches to three considerations:
+
+1. instrument selection (from among the available SNPs)  
+1. modeling and controlling for horizontal pleiotropy  
+1. statistical inference procedure
+
+Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
+be statistically independent. This restriction to 
+independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
+selected through linkage disequilibrium (LD) clumping [@purcell2007plink].
+
+However, LD clumping for choosing instrument SNPs can be problematic for several 
+reasons. First, the chosen SNPs may not be causal for the outcome. This can 
+happen, for instance, by using a noncausal SNP that is in high LD with a 
+causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
+causal SNPs may be in high LD together. LD clumping is likely to omit one or more
+of the causal SNPs, which can diminish statistical power in MR analysis.
+
+Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that 
+bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
+the standard MR assumptions and can lead to biased 
+causal effect estimates and diminished statistical power. We discuss approaches to 
+modeling and accounting for horizontal pleiotropy below.
 
@@
 a strategy for determining a single causal effect from summary statistics for a 
-collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
+collection of SNPs.
+
+Below, we discuss a number of MR methods that use multiple 
+instruments. The overall trend appears to be in the direction of increasingly sophisticated 
+modeling of horizontal pleiotropy (and correlated horizontal pleiotropy) while
+attempting to maintain scalability and computational efficiency.
+
+## Statistical models and methods for MR with multiple instruments
+
+One approach to integrating multiple SNP instruments into a MR framework is
+through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
+calculate an unweighted score as the number of exposure-increasing alleles in 
+the subject's genotypes. They also calculate a weighted allele score by using 
+the exposure effects as weights. For example, a subject with $g_k$ copies of 
+exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
+and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.
+
+@burgess_use_2013 found that when weights are obtained from external data or from 
+cross-validation or jackknife approaches applied to the analysis data, the allele score 
+functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
+two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
+
+@burgess_mendelian_2013 termed their method "inverse variance 
 weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
@@
 
-In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity $Q$ test.
-In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).
+In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary 
+data, @greco_m_detecting_2015 developed the between-instrument heterogeneity $Q$ test. In a meta-analysis 
+of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).
 
@@
 
-Genetic instrumental variable regression uses GWAS summary statistics to construct polygenic scores for 
-the outcome of interest [@diprete_genetic_2018]. In constructing polygenic scores from subsets of a GWAS 
-cohort, @diprete_genetic_2018 aim to reduce bias due to horizontal pleiotropy.
-
 # Correlated horizontal pleiotropy

### 2022-03-04 omitted.Rmd
started omitted.Rmd for removed refs

---
title: "Omitted lit"
author: "Frederick Boehm"
date: "3/3/2022"
bibliography: exported.bib
output: html_document
---

Genetic instrumental variable regression uses GWAS summary statistics to construct polygenic scores for 
the outcome of interest [@diprete_genetic_2018].

# references

### 2022-03-04 mr-review.Rmd
fixed title & organized text chunks

@@
 ---
-title: "Mendelian randomization for causal inference in genome-wide association studies"
+title: "Mendelian randomization with genome-wide association studies data"
 author: "Frederick J. Boehm & Xiang Zhou"
@@
 Below, we discuss a number of MR methods that use multiple 
-instruments. The overall trend appears to be in the direction of increasingly sophisticated 
+SNP instruments. The overall trend appears to be in the direction of increasingly sophisticated 
 modeling of horizontal pleiotropy (and correlated horizontal pleiotropy) while
@@
 ## Statistical models and methods for MR with multiple instruments
+
+### Methods with Individual-level data
 
@@
 and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$. 
-
 @burgess_use_2013 found that when weights are obtained from external data or from 
@@
 two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
-
-@burgess_mendelian_2013 termed their method "inverse variance 
-weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
-of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
-causal effects.
-
-Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from $M$ independent SNPs by using Equation \@ref(eq:ivw).
-
-\begin{equation}
-  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
-  (\#eq:ivw)
-\end{equation}
-
-In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary 
-data, @greco_m_detecting_2015 developed the between-instrument heterogeneity $Q$ test. In a meta-analysis 
-of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).
-
-\begin{equation}
-Q = \sum_{k = 1}^K w_k(\hat\beta^{(k)}_{XY} - \mu_F)^2
-(\#eq:q)
-\end{equation}
-
-In Equation \@ref(eq:q), $w_k$ is the inverse variance of the Wald estimator, and $\mu_F = \frac{\sum_{k = 1}^K w_k \hat\beta^{(k)}_{XY}}{\sum_{k = 1}^K w_k}$.
-
-While the $Q$ test tends to be conservative in small sample sizes, its power increases with increasing 
-sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
-
-@bowden_mendelian_2015 adapted Egger regression, originally developed 
-to detect bias in meta-analyses, to detect bias from horizontal 
-pleiotropy in two-sample MR studies with GWAS summary data 
-[@egger1997bias]. The slope coefficient estimates the causal effect of the exposure on the outcome. 
-Under the assumption that the SNP-exposure relationship is independent 
-of the SNP's pleiotropic effects, Egger's test offers a valid test of 
-the null causal hypothesis and consistently estimates the causal 
-effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].
-
-@bowden_consistent_2016, noting that consistency with IVW MR is not robust to 
-invalid instrument use, developed a method for consistent MR analysis with  
-weighted median-based estimators, building on research from @han2008detecting. 
-Its breakdown point is 50\%, 
-meaning that up to 50\% of SNP instruments can be invalid while maintaining consistency in estimation 
-[@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate 
-statistic, from least to greatest. If an odd number, say $2K +1$, of SNP instruments is used, then 
-the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say 
-$2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio 
-estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.
-
-Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted 
-median estimator. To construct this estimator, order the ratio estimates from least to greatest, as 
-with the simple median estimator. The weighted median estimator, then is the median of a distribution
-defined by having $\tilde \beta_j$ as the $p_j = 100(\sum_{k = 1}^j - \frac{w_j}{2})$ percentile, 
-where $w_j$ is the weight assigned to the $j^{th}$ ordered ratio estimate. In this setting, the 
-simple median estimator is seen to be the weighted median estimator when all weights are equal.
-
-@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
-penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
-the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
-a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized 
-weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald 
-ratio estimates.
 
@@
 
-@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data. 
-They first calculate Wald ratio estimates for every instrumental SNP. They then apply 
-smoothing to the empirical distribution of ratio estimates. The mode of this smoothed 
-distribution is the simple mode-based estimate of the causal effect 
-[@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by 
-weighting the empirical distribution of ratio estimates by the inverse variance of each 
-estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based 
-estimators with weighted and unweighted median-based estimators, MR Egger regression, and 
-IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than 
-other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to 
-detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of 
-MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect 
-when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In 
-this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the 
-other estimators.
-
-@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
-developing three new methods for two-sample MR analysis with GWAS summary data: robust 
-regression, penalized weights, and LASSO penalization. The first two can be viewed as 
-modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
-downweighting or excluding variants with heterogeneous causal estimates.
-Recall that MR-Egger offers consistent causal effect estimates when there is no 
-horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
-"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
-InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
-instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
-MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
-formulate an estimator with high breakdown point and high efficiency. The second estimation 
-method, with penalized weights, is much like the penalized median estimator, from above, 
-except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
-above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
-objective function from MR-Egger to get Equation \@ref(eq:lasso) as the objective function.
-
-\begin{equation}
-\sum_j [Var(\hat\beta_j)]^{-1}(\hat\beta_{Yj} - \theta_{0j} - \theta_1 \hat\beta_{1j})^2 + \lambda \sum_j |\theta_{0j}|
-(\#eq:lasso)
-\end{equation}
-
-@rees_robust_2019 studied two procedures for choosing $\lambda$: a heterogeneity stopping rule 
-and a cross-validation rule. Taken together, these three methods offer additional contributions to 
-a suite of sensitivity analysis tools that practitioners should consider.
-
-@burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
-analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or 
-more of the three core MR assumptions. Working in the setting of two-sample MR with GWAS data, they begin by 
-considering a collection of candidate SNP instruments, which may contain some invalid SNP instruments. For 
-each candidate SNP instrument, they calculate the Wald ratio estimate of the causal effect. With the 
-collection of Wald ratio estimates, @burgess_robust_2020 reason that the set of valid SNP instruments have 
-Wald ratio estimates that arise from a normal distribution centered on the true causal parameter with variance 
-equal to that of the Wald ratio estimator. The invalid instruments, however, are assumed to follow a normal 
-distribution centered at zero with a variance greater than that of the Wald ratio estimator. They then work 
-with profile log likelihood over a grid of causal parameter values to make inferences.
-
-@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted 
-profile score to accomplish statistical inference in two-sample MR analyses with GWAS 
-summary data. Dividing pleiotropic effects into systematic and idiosyncratic, they model the 
-systematic pleiotropy with random effects. In so doing, no SNP satisfies the exclusion 
-restriction assumption. Adjusting their profile likelihood estimator from the setting 
-without pleiotropy, @zhao_statistical_2020 obtain an estimator with consistency and 
-asymptotic normality [@mccullagh_simple_1990]. Idiosyncratic pleiotropy is addressed through robustification of 
-the adjusted profile score [@huber_robust_1964]. They then demonstrate the properties of their estimator by 
-analyzing simulated and real data [@zhao_statistical_2020].
-
-@bowden_improving_2019, working with a collection of SNP instruments, reasoned that if all three MR assumptions are satisfied, then the 
-collection of Wald ratio estimates should be homogeneous; thus, any departures from 
-homogeneity may signal a violation of one or more assumptions. They reported 
-modified weights, in addition to studying weights derived from first order and 
-second order approximations to the variance. Their modified weights lead 
-to an estimator that is useful for quantifying heterogeneity and detecting outliers.
-
-@cho_exploiting_2020 developed the MR-TRYX framework for exploiting horizontal pleiotropy
-to identify alternative causal pathways. They did this by positing that, for a 
-collection of putative SNP instruments, any heterogeneity in the Wald ratio estimates
-may be due to horizontal pleiotropy. Once they identify a SNP with outlying 
-ratio estimates, they search GWAS results by querying the MR-Base database to identify complex traits (in other studies) that 
-associate with the outlying SNP [@hemani_mr-base_2018]. They follow up promising trait associations to 
-account for their observations with multiple causal pathways.
-
-@foley_mr-clust_2021 recognized that in some MR studies, the Wald ratio estimates for different
-SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms 
-and may identify the SNPs that are involved in them. They devised a method to detect 
-these clusters in the collection of Wald ratio estimates and found SNP instruments that 
-point to distinct causal pathways of biological importance [@foley_mr-clust_2021].
+@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods. Central to their
+work is the proposal to create a new instrumental variable as a linear combination of genotype data from 
+a collection of SNP instruments. They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically  
+pleiotropic traits. For a collection of $K$ SNPs, they choose the $K$-vector that meets these criteria and 
+maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a 
+$\ell_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation 
+of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. 
+@jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are 
+unbiased in their simulation settings.
+
+@spiller_detecting_2018 and @spiller_interaction-based_2020 present a method, MR-GxE, that exploits gene by 
+environment interactions to detect and to adjust for bias due to horizontal pleiotropy. They leverage the insight 
+that a SNP instrument-covariate interaction that induces differences in SNP instrument-exposure associations 
+@spiller_interaction-based_2020 treat a SNP-covariate interaction as an instrumental variable, and, by so doing, they impose assumptions on that interaction.
+Together, these assumptions require that MR-GxE assume that horizontal pleiotropy effects are constant across the study cohort.
+The authors devise a three-step inferential procedure for two-sample MR with GWAS summary statistics:
+
+1. Estimate the SNP-exposure and SNP-outcome associations at a range of covariate values  
+1. Regress the SNP-outcome estimated associations on the SNP-exposure estimated associations  
+1. Treat the resulting slope as the causal effect estimate and the intercept as the mean horizontal pleiotropy effect.
+
+@spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set 
+of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant 
+pleiotropy assumption.
 
@@
 
-@hemani_automating_2017
+Finally, before considering methods for MR with GWAS summary statistics, we consider 
+a new method for one-sample MR. @berzuini_bayesian_2020 proposed a Bayesian approach to MR where SNP instruments
+are permitted to exhibit pleiotropic effects on the outcome. Working in the setting of 
+one-sample MR with individual-level data, they construct their 
+model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects. Their 
+method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP 
+instruments) and permits correlated SNP instruments, unlike many frequentist methods. Posterior inference 
+proceeds by Markov chain monte carlo sampling. They also elaborate their model for 
+a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.
+
+### Methods with Summary data
+
+The proliferation of publicly available GWAS data coincides with the acceleration
+of methods for two-sample MR with GWAS summary statistics. Our timeline highlights some
+of the many two-sample MR methods with GWAS summary statistics, with an increasing 
+density of methods over time in the last three years (Figure \@ref(fig:timeline)).
+
+```{r, timeline, fig.cap = "Two-sample MR with GWAS summary statistics methods publications density increases over time."}
+knitr::include_graphics("figures/timeline.pdf")
+```
+
+@burgess_mendelian_2013 termed their method "inverse variance 
+weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
+of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
+causal effects.
+
+Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from $M$ independent SNPs by using Equation \@ref(eq:ivw).
+
+\begin{equation}
+  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
+  (\#eq:ivw)
+\end{equation}
+
+@bowden_mendelian_2015 adapted Egger regression, originally developed 
+to detect bias in meta-analyses, to detect bias from horizontal 
+pleiotropy in two-sample MR studies with GWAS summary statistics or individual-level data
+[@egger1997bias]. Their approach is termed MR-Egger. They formulate their model as
+
+\begin{eqnarray}
+X_i = \sum_{j=1}^J\gamma_jG_{ij} + U_i + \epsilon_{Xi}\nonumber \\
+Y_i = \sum_{j=1}^J\alpha_jG_{ij} + \beta X_i + U_i + \epsilon_{Yi}
+(\#eq:egger)
+\end{eqnarray}
+
+where $U$ designates confounders, $X$ and $Y$ denote the exposure and outcome, respectively, $G$ is a SNP 
+instrument, and $\epsilon_X$ and $\epsilon_Y$ are the zero-mean, normally distributed random errors.
+
+@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument $j$ and 
+the effects defined in Equation \@ref(eq:egger) and define the new parameter $\Gamma_j = \alpha_j + \beta\gamma_j$ (Equation \@ref(eq:egger1)).
+
+\begin{eqnarray}
+Y_i = (\alpha_j + \beta\gamma_j)G_{ij} +  \epsilon'_{Yij} \nonumber\\
+= \Gamma_jG_{ij} + \epsilon'_{Yij}
+(\#eq:egger1)
+\end{eqnarray}
+
+In Equation \@ref(eq:egger1), $\alpha_j = 0$ when SNP $j$ is a valid instrument.
+
+MR-Egger fits the regression model in Equation \@ref(eq:egger2).
+
+\begin{equation}
+\hat\Gamma_j = \beta_{0E} + \beta_E \hat\gamma_j
+(\#eq:egger2)
+\end{equation}
+
+The notation in Equation \@ref(eq:egger2) could be a little misleading. The independent
+and dependent variables are $\hat\gamma_j$ and $\hat\Gamma_j$, respectively, while 
+the regression parameters are $\beta_{0E}$ and $\beta_E$.
+
+The slope coefficient $\hat\beta_E$ estimates the causal effect of the exposure on the outcome. 
+Under the assumption that the SNP-exposure relationship is independent 
+of the SNP's pleiotropic effects, Egger's test offers a valid test of 
+the null causal hypothesis and consistently estimates the causal 
+effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].
+
+In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary 
+data, @greco_m_detecting_2015 developed the between-instrument heterogeneity $Q$ test. In a meta-analysis 
+of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).
+
+\begin{equation}
+Q = \sum_{k = 1}^K w_k(\hat\beta^{(k)}_{XY} - \mu_F)^2
+(\#eq:q)
+\end{equation}
+
+In Equation \@ref(eq:q), $w_k$ is the inverse variance of the Wald estimator, and $\mu_F = \frac{\sum_{k = 1}^K w_k \hat\beta^{(k)}_{XY}}{\sum_{k = 1}^K w_k}$.
+
+While the $Q$ test tends to be conservative in small sample sizes, its power increases with increasing 
+sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
+
+@bowden_consistent_2016, noting that consistency with IVW MR is not robust to 
+invalid instrument use, developed a method for consistent MR analysis with  
+weighted median-based estimators, building on research from @han2008detecting. 
+Its breakdown point is 50\%, 
+meaning that up to 50\% of SNP instruments can be invalid while maintaining consistency in estimation 
+[@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate 
+statistic, from least to greatest. If an odd number, say $2K +1$, of SNP instruments is used, then 
+the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say 
+$2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio 
+estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.
+
+Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted 
+median estimator. To construct this estimator, order the ratio estimates from least to greatest, as 
+with the simple median estimator. The weighted median estimator, then is the median of a distribution
+defined by having $\tilde \beta_j$ as the $p_j = 100(\sum_{k = 1}^j - \frac{w_j}{2})$ percentile, 
+where $w_j$ is the weight assigned to the $j^{th}$ ordered ratio estimate. In this setting, the 
+simple median estimator is seen to be the weighted median estimator when all weights are equal.
+
+@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
+penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
+the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
+a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized 
+weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald 
+ratio estimates.
 
@@
 
-@berzuini_bayesian_2020 proposed a Bayesian approach to MR where SNP instruments
-are permitted to exhibit pleiotropic effects on the outcome. Working in the setting of 
-one-sample MR with individual-level data, they construct their 
-model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects. Their 
-method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP 
-instruments) and permits correlated SNP instruments, unlike many frequentist methods. Posterior inference 
-proceeds by Markov chain monte carlo sampling. They also elaborate their model for 
-a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.
+@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data. 
+They first calculate Wald ratio estimates for every instrumental SNP. They then apply 
+smoothing to the empirical distribution of ratio estimates. The mode of this smoothed 
+distribution is the simple mode-based estimate of the causal effect 
+[@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by 
+weighting the empirical distribution of ratio estimates by the inverse variance of each 
+estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based 
+estimators with weighted and unweighted median-based estimators, MR Egger regression, and 
+IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than 
+other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to 
+detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of 
+MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect 
+when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In 
+this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the 
+other estimators.
+
+@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
+developing three new methods for two-sample MR analysis with GWAS summary data: robust 
+regression, penalized weights, and LASSO penalization. The first two can be viewed as 
+modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
+downweighting or excluding variants with heterogeneous causal estimates.
+Recall that MR-Egger offers consistent causal effect estimates when there is no 
+horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
+"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
+InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
+instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
+MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
+formulate an estimator with high breakdown point and high efficiency. The second estimation 
+method, with penalized weights, is much like the penalized median estimator, from above, 
+except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
+above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
+objective function from MR-Egger to get Equation \@ref(eq:lasso) as the objective function.
+
+\begin{equation}
+\sum_j [Var(\hat\beta_j)]^{-1}(\hat\beta_{Yj} - \theta_{0j} - \theta_1 \hat\beta_{1j})^2 + \lambda \sum_j |\theta_{0j}|
+(\#eq:lasso)
+\end{equation}
+
+@rees_robust_2019 studied two procedures for choosing $\lambda$: a heterogeneity stopping rule 
+and a cross-validation rule. Taken together, these three methods offer additional contributions to 
+a suite of sensitivity analysis tools that practitioners should consider.
+
+@burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
+analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or 
+more of the three core MR assumptions. Working in the setting of two-sample MR with GWAS data, they begin by 
+considering a collection of candidate SNP instruments, which may contain some invalid SNP instruments. For 
+each candidate SNP instrument, they calculate the Wald ratio estimate of the causal effect. With the 
+collection of Wald ratio estimates, @burgess_robust_2020 reason that the set of valid SNP instruments have 
+Wald ratio estimates that arise from a normal distribution centered on the true causal parameter with variance 
+equal to that of the Wald ratio estimator. The invalid instruments, however, are assumed to follow a normal 
+distribution centered at zero with a variance greater than that of the Wald ratio estimator. They then work 
+with profile log likelihood over a grid of causal parameter values to make inferences.
+
+@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted 
+profile score to accomplish statistical inference in two-sample MR analyses with GWAS 
+summary data. Dividing pleiotropic effects into systematic and idiosyncratic, they model the 
+systematic pleiotropy with random effects. In so doing, no SNP satisfies the exclusion 
+restriction assumption. Adjusting their profile likelihood estimator from the setting 
+without pleiotropy, @zhao_statistical_2020 obtain an estimator with consistency and 
+asymptotic normality [@mccullagh_simple_1990]. Idiosyncratic pleiotropy is addressed through robustification of 
+the adjusted profile score [@huber_robust_1964]. They then demonstrate the properties of their estimator by 
+analyzing simulated and real data [@zhao_statistical_2020].
+
+@bowden_improving_2019, working with a collection of SNP instruments, reasoned that if all three MR assumptions are satisfied, then the 
+collection of Wald ratio estimates should be homogeneous; thus, any departures from 
+homogeneity may signal a violation of one or more assumptions. They reported 
+modified weights, in addition to studying weights derived from first order and 
+second order approximations to the variance. Their modified weights lead 
+to an estimator that is useful for quantifying heterogeneity and detecting outliers.
+
+@cho_exploiting_2020 developed the MR-TRYX framework for exploiting horizontal pleiotropy
+to identify alternative causal pathways. They did this by positing that, for a 
+collection of putative SNP instruments, any heterogeneity in the Wald ratio estimates
+may be due to horizontal pleiotropy. Once they identify a SNP with outlying 
+ratio estimates, they search GWAS results by querying the MR-Base database to identify complex traits (in other studies) that 
+associate with the outlying SNP [@hemani_mr-base_2018]. They follow up promising trait associations to 
+account for their observations with multiple causal pathways.
+
+@foley_mr-clust_2021 recognized that in some MR studies, the Wald ratio estimates for different
+SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms 
+and may identify the SNPs that are involved in them. They devised a method to detect 
+these clusters in the collection of Wald ratio estimates and found SNP instruments that 
+point to distinct causal pathways of biological importance [@foley_mr-clust_2021].
 
@@
 
-@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods. Central to their
-work is the proposal to create a new instrumental variable as a linear combination of genotype data from 
-a collection of SNP instruments. They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically  
-pleiotropic traits. For a collection of $K$ SNPs, they choose the $K$-vector that meets these criteria and 
-maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a 
-$\ell_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation 
-of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. 
-@jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are 
-unbiased in their simulation settings.
-
 @xu_mrcip_2021 consider two classes of horizontal pleiotropy with their method, MRCIP. They partition horizontal 
@@
 SNP instruments (*i.e.*, when all SNP instruments are invalid) and when the InSIDE assumption is violated 
-[@xu_mrcip_2021].
-Their PRW-EM algorithm adds a reweighting step to the traditional expectation maximization algorithm to estimate the causal effect parameter in MR analysis [@dempster1977maximum].
-Finally, they compare  MRCIP to other MR methods with both simulated and real data sets.
-@xu_mrcip_2021, like the developers of MR-RAPS, assume that the horizontal pleiotropy is balanced, *i.e.*, that the mean horizontal pleiotropic effect is zero.
-One direction for future study is to augment MRCIP to permit unbalanced horizontal pleiotropy.
+[@xu_mrcip_2021]. Their PRW-EM algorithm adds a reweighting step to the traditional expectation 
+maximization algorithm to estimate the causal effect parameter in MR analysis [@dempster1977maximum]. 
+Finally, they compare  MRCIP to other MR methods with both simulated and real data sets. @xu_mrcip_2021, 
+like the developers of MR-RAPS, assume that the horizontal pleiotropy is balanced, *i.e.*, that the mean 
+horizontal pleiotropic effect is zero. One direction for future study is to augment MRCIP to permit 
+unbalanced horizontal pleiotropy.
 
@@
 demonstrate conditions under which their estimators are equivalent to those in works from @zhao_statistical_2020, @windmeijer_use_2019, and @bowden_improving_2019.
-
-@spiller_detecting_2018 and @spiller_interaction-based_2020 present a method, MR-GxE, that exploits gene by 
-environment interactions to detect and to adjust for bias due to horizontal pleiotropy. They leverage the insight 
-that a SNP instrument-covariate interaction that induces differences in SNP instrument-exposure associations 
-@spiller_interaction-based_2020 treat a SNP-covariate interaction as an instrumental variable, and, by so doing, they impose assumptions on that interaction.
-Together, these assumptions require that MR-GxE assume that horizontal pleiotropy effects are constant across the study cohort.
-The authors devise a three-step inferential procedure for two-sample MR with GWAS summary statistics:
-
-1. Estimate the SNP-exposure and SNP-outcome associations at a range of covariate values  
-1. Regress the SNP-outcome estimated associations on the SNP-exposure estimated associations  
-1. Treat the resulting slope as the causal effect estimate and the intercept as the mean horizontal pleiotropy effect.
-
-@spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set 
-of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant 
-pleiotropy assumption.
 
@@
 
-@howey_bayesian_2020 use anchored Bayesian networks to
-
 @sanderson_examination_2019 found that multivariable MR, where multiple exposures jointly 
@@
 the matrix of tag SNP genotypes.
-
-@jiang_hierarchical_2021, building on research from @newcombe2016jam, present a two-stage hierarchical Bayesian 
-model that aims to unify MR and transcriptome-wide association studies frameworks.
 
@@
 
-## Multivariate exposures in MR analysis
-
-@grant_pleiotropy_2021: we introduce three methods which build on common approaches in the univariable setting: {MVMR}-Robust; {MVMR}-Median; and {MVMR}-Lasso
-
 ## Sample overlap in biobanks
@@
 
-@mounier_bias_2021
-@hartwig_two-sample_2016
-@lawlor_commentary_2016
-@burgess_bias_2016
-
 ## Bi-directional causal inference with MR
-
-@darrous_simultaneous_2021
 
@@
 the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].
-
-\begin{eqnarray}
-U = X\beta_U \odot Z_U + \epsilon_U\nonumber \\
-A = U\nu + B\delta + X\beta_A \odot Z_A + \epsilon_A \nonumber \\
-B = U\eta + A\gamma + X\beta_B \odot Z_B + \epsilon_B 
-  (\#eq:wwer)
-\end{eqnarray}
-
-To develop MR methods that are robust to correlated horizontal pleiotropy, 
-@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
-accommodates both uni-directional and bi-directional effects between exposure and outcome. 
-@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
-the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$.
-$Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to indicate which .
-Note that the model is agnostic with regards to which is the
-outcome and which the exposure, so we follow the authors' notation here. The symbol $\odot$ denotes the Hadamard elementwise product.
-
-In Equations \@ref(eq:wwer),
 

### 2022-03-04 omitted.Rmd
fixed title & organized text chunks

@@
 
+@hemani_automating_2017
+
+@howey_bayesian_2020 use anchored Bayesian networks to
+
+@jiang_hierarchical_2021, building on research from @newcombe2016jam, present a two-stage hierarchical Bayesian 
+model that aims to unify MR and transcriptome-wide association studies frameworks.
+
+@mounier_bias_2021
+@hartwig_two-sample_2016
+@lawlor_commentary_2016
+@burgess_bias_2016
+
+@darrous_simultaneous_2021
+
+\begin{eqnarray}
+U = X\beta_U \odot Z_U + \epsilon_U\nonumber \\
+A = U\nu + B\delta + X\beta_A \odot Z_A + \epsilon_A \nonumber \\
+B = U\eta + A\gamma + X\beta_B \odot Z_B + \epsilon_B 
+  (\#eq:wwer)
+\end{eqnarray}
+
+To develop MR methods that are robust to correlated horizontal pleiotropy, 
+@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
+accommodates both uni-directional and bi-directional effects between exposure and outcome. 
+@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
+the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$.
+$Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to designate which .
+Note that the model is agnostic with regards to which is the
+outcome and which the exposure, so we follow the authors' notation here. The symbol $\odot$ denotes the Hadamard elementwise product.
+
+In Equations \@ref(eq:wwer),
+
+## Multivariate exposures in MR analysis
+
+@grant_pleiotropy_2021: we introduce three methods which build on common approaches in the univariable setting: {MVMR}-Robust; {MVMR}-Median; and {MVMR}-Lasso
+
 # references

### 2022-03-10 article-counts-per-year-figure.Rmd
fixed figure

--- 2022-02-02 article-counts-per-year-figure.Rmd
@@
   tidyr::pivot_longer(!year) %>%
-  ggplot2::ggplot(ggplot2::aes(x = year, y = value, colour = name)) + ggplot2::geom_point()
-ggplot2::ggsave("figures/mr-gwas-counts-scatter.png")
+  dplyr::filter(year != 2022) %>%
+  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
+  ggplot2::geom_bar(stat = "identity", 
+                    position=ggplot2::position_dodge()) + 
+  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
+  ggplot2::ylab("Annual publication count") + 
+  ggplot2::xlab("Year") + 
+  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) +
+  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 270, vjust = 0.5))
+
+ggplot2::ggsave("figures/mr-gwas-counts.png")
 ```

### 2022-04-04 reply-to-reviewers.Rmd
started reply to reviewers

---
title: "Reply to reviewers"
author: "Fred Boehm"
date: "4/4/2022"
output: html_document
---

Editor and Reviewer comments:

With pleasure I read this manuscript that compares methods for Mendelian randomization (MR). I think the authors did a lot of work in making such an extensive review. However, I do have some comments:

-I think the current manuscript is somewhere in between guiding  people who are new to MR and giving a complete overview of all methods. I think adding a standard MR dag would be beneficial to the first type of reader. This could be part of Fig 3 perhaps.

-On page 13, the authors write 'This can happen, for instance, by using a noncausal SNP that is in high LD with a causal SNP. Second, because many traits are polygenic or omnigenic, multiple causal SNPs may be
in high LD together. LD clumping is likely to omit one or more of the causal SNPs, which candiminish statistical power in MR analysis.' (p13). I think this is not too much of a concern, as long as there is strong correlation between these SNPs, because the GWAS estimate of a single variant also takes into account it's LD structure with other SNPs. Hence, it is not really only the effect of that specific SNP that is measured and hence not much information should be lost due to clumping.

-The current section 'Applications of MR in GWAS' seems quite random why these specific studies were chosen. To me as a reader I wonder why are these specific studies chosen.

-I think the manuscript could give more guidance to readers on what methods should be used or not. This could easily extend Figure 4. Currently it is unclear how different these methods really are and when they will give different results.

-I think the manuscript would improve if it would say something on future prospects (this is somewhat done in the section summary and outlook, but is currently a bit underwhelming). E.g. what are currently open challenges?

Minor remarks:

-Please beware that Mendelian randomization should be written with capital M (e.g. caption Figure 1).

-mathematical symbols sometimes difficult to read (\hat{} command often overlap)

-I find the graphical abstract unclear.

### 2022-04-26 reply-to-review1/mr-review.Rmd
copied chunks of mr-review.txt into mr-review.Rmd

--- 2022-03-04 mr-review.Rmd
@@
 
-Complex trait genetics studies in humans have yielded thousands of new associations for  
-diverse traits. The next step in understanding biology of health and disease is to
-identify causal relationships among complex traits. Two-sample Mendelian randomization studies 
-in genome-wide association studies can uncover causal relationships 
-between complex traits. By leveraging shared SNP genotypes as instrumental variables, or proxies, for 
-the exposure complex trait, investigators can tease out causal effects from observational data, provided
-that necessary assumptions are satisfied. We discuss below the growth of genome-wide association studies 
-in parallel with development of Mendelian randomization methods.
-We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.
+Genome-wide association studies have yielded thousands of associations for many common diseases and disease-related complex traits.
+The identified associations made it possible to identify the causal risk factors underlying diseases and investigate the causal relationships among complex traits through Mendelian randomization.
+Mendelian randomization is a form of instrumental variable analysis that uses SNP associations from genome-wide association studies as instruments to study and uncover causal relationships between complex traits.
+By leveraging SNP genotypes as instrumental variables, or proxies, for the exposure complex trait, investigators can tease out causal effects from observational data, provided that necessary assumptions are satisfied.
+We discuss below the development of Mendelian randomization methods in parallel with the growth of genome-wide association studies.
+We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods with relaxed causality assumptions and that this area continues to offer opportunities for robust biological discoveries.
 
@@
 
-Great interest has developed in inferring causal relationships between complex 
-traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
-enhancing our understanding of the biology of health and disease. Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
-genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
-inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].
-
-Genome-wide association studies (GWAS) present unique opportunities for 
-discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS
-interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
-In the nearly 17 years since publication of early studies, researchers have reported 
-thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].
-Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of
-human health and disease.
+Great interest has developed in inferring causal relationships between complex traits, i.e., traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies.
+Discovery of such relationships is crucial to enhancing our understanding of the biology of health and disease.
+Inferring causal relationship in genetics is often carried out through Mendelian randomization (MR) analysis.
+MR analysis is a form of instrumental variable analysis in which genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].
+MR analysis is facilitated by the development of genome-wide association studies (GWAS), which present unique opportunities for discovery of causal relationships via MR [@swerdlow_selecting_2016].
+A GWAS interrogates millions of single nucleotide polymorphisms (SNPs) to infer which are associated with the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
+In the nearly 17 years since publication of early studies, researchers have reported thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].
+Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of human health and disease.
+The association results from many large-scale GWAS are nowadays readily available, often in the form of summary statistics that include the marginal SNP p-values and/or their effect size estimates and standard errors[@buniello2019nhgri].
+The identified SNP associations are used as the main input for MR analysis and thus the wide availability of GWAS summary statistics clear the way for effective MR analysis in complex genetics studies.
 
@@
 
-A number of excellent MR reviews have appeared in the literature [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022]. Our review doesn't 
-aim to replace them, but seeks to complement their findings. 
-We present recent methods developments for an audience of statisticians and epidemiologists. We discuss in detail methods advances before briefly summarizing 
-applications and directions for future research. The value of MR analysis with 
-GWAS data will continue to grow in parallel with statistical methods advances.
-
-Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
-publication of the early GWAS (Figure \@ref(fig:article-counts)). Many studies use the "two-sample" MR 
-design, in which one cohort of subjects has measurements for the exposure, while a second cohort has 
-measurements for a second trait [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in 
-both sets of subjects.
-
-Recent methods developments have enabled execution of two-sample MR studies with only 
-GWAS summary statistics [@burgess2015using]. These strategies are especially pertinent now that there is 
-an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
-ability to use published data has contributed to a rapid increase in the number of
-two-sample MR methods for GWAS summary statistics (Figure \@ref(fig:article-counts)).
+Effective MR analyses are enabled by many MR methods developed in recent years.
+Methods for and uses of MR have appeared at a rapid and accelerating pace since the publication of the early GWAS (Figure 1).
+The overall trend of the methodology development is in the direction of increasingly sophisticated modeling of horizontal pleiotropy including both independent and correlated horizontal pleiotropy while attempting to maintain scalability and computational efficiency in the presence of multiple correlated SNPs.
+Here, we present a comprehensive review on 47 MR methods, primarily in the context of GWAS summary statistics, to help practitioners to choose which MR methods to use in applied data analysis.
+We discuss the basics form of MR analysis, the causality assumptions in MR, and how recent MR methods are developed to ensure robust results in the presence of assumption violations.
+We discuss in detail methods advances before briefly summarizing applications and directions for future research.
+Different from existing MR reviews [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022], we discuss in detail recent methods developments that enable modeling of horizontal pleiotropy and correlated horizontal pleiotropy and place these developments in the larger context of MR analysis with GWAS data.
+We present these recent methods developments for an audience of both statisticians and epidemiologists.
+We hope our review will facilitate the further advance of MR methods and their wide application on GWAS data.
+
+## Assumptions of Mendelian randomization
+
+Mendelian randomization uses genetic markers in the form of SNP genotypes as the proxy (“instrument” or “instrumental variable”) for the “exposure,” a complex trait, and asks whether the “exposure” variable is causal for the “outcome” variable, which is typically a second complex trait.
+The exposure and outcome variables can be binary, a count, a time to event, or a continuous variable.
+For brevity and simplicity, we focus on continuous exposures and outcomes before considering other classes of outcome variables in a later section.
+Most MR studies use the “two-sample” MR design, in which one cohort of subjects has measurements for the exposure, while a second cohort has measurements for the outcome, with both cohorts sharing the same set of SNP instruments [@pierce_efficient_2013].
+We focus on the two-sample MR design before considering the one-sample design and the partial two-sample design where samples are partially overlapped between the two cohorts.
+MR aims to infer the causal effect of the exposure variable on the outcome variable in observational studies.
+The proper causal interpretation in an MR study requires the SNP instruments to satisfy three causality assumptions.
+The first assumption states that SNP instruments are associated with the exposure.
+If the first assumption, about the association between the SNP instruments and the exposure, holds, but this association is weak, then an amplification of biases, such as those due to violations of assumptions 2 and 3, may result [@vanderweele_methodological_2014].
+Indeed, bias in causal estimates can increase with decreases in the strength of association between the SNP instrument and exposure [@burgess_avoiding_2011].
+The second assumption requires independence, conditional on the exposure and all measured and unmeasured confounders, between the SNP instruments and the outcome.
+The second MR assumption, often termed the exclusion restriction assumption, can also be stated as the need for the SNP instruments to affect the outcome only through the exposure.
+In a causal diagram, this means that the only path from the SNP instruments to the outcome is that containing the exposure.
+The last of the three assumptions states that the SNP instruments are independent of all (measured and unmeasured) confounders of the relationship between exposure and outcome.
+The third assumption ensures that the SNP instruments are independent of all confounders of the exposure-outcome relationship.
+When the three MR assumptions hold, one may use instrumental variable statistical methods, with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
+In practice, careful evaluation of the assumptions is needed.
+Because assumptions 2 and 3, which involve absence of certain types of confounding, can’t be verified, practitioners must be cautious when planning MR analyses.
+Sensitivity analysis is highly recommended as a way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014].
+We will discuss these sensitivity analyses in more detail in a late section and consider their evaluation in MR studies [@vanderweele_methodological_2014].
+The causal assumptions of MR also help one to understand the origin of the term “Mendelian randomization”.
+Recall Mendel’s inheritance laws [mendel1865versuche], under the assumption that alleles segregate randomly from parent to offspring, the offspring genotypes are unlikely to be associated with confounders of the exposure-outcome relationship.
+Additionally, reverse causation from the outcome or exposure to the genotypes is unanticipated since germ-line genotypes are fixed at conception and, thus, precede realization of other observed variables.
+Therefore, using SNP instruments in the instrumental variable analysis is often referred to as Mendelian randomization.
+
+## Statistical models and methods for MR with one instrument
+We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where there is a single SNP instrument and a single outcome variable.
+Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, likelihood-based methods, and semiparametric methods [@burgess_review_2017].
+We discuss each of these before turning to methods that leverage multiple SNP instruments.
+
+### Ratio of coefficients method
+The ratio of coefficients method, also known as the Wald method, estimates the causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940].
+For a continuous outcome, the causal effect estimator,  is defined in Equation (3.1).
+
+In Equation (3.1),  and  are the slope estimates from the regressions of the outcome and exposure, respectively, on the SNP instrument.
+Since the Wald method requires only the regression coefficients, it can be used with summary data.
+However, the Wald method doesn’t accommodate multiple SNP instruments, which limits its direct use in the GWAS setting as will be discussed later.
+
+### Two-stage methods
+A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969].
+For a continuous outcome, one may perform “two-stage least squares,” which involves two linear regressions.
+First, the exposure variable is regressed on the instrument (i.e., the SNP genotypes) (Equation (3.2)).
+We denote each subject’s exposure variable value as  in Equation (3.2), while  denotes an intercept term and  is the slope.
+denotes the SNP instrument genotype for subject , while , the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.
+The resulting fitted values for the exposure variable from Equation (3.2) are the independent variable in the second linear regression, where the outcome is the dependent variable.
+The causal effect estimate, then, is the coefficient  obtained from the second regression analysis (Equation (3.3)).
+Note also that the random errors in Equation (3.3) are assumed independent of those in Equation (3.2).
+Like the  in Equation (3.2), the  are assumed independent and identically distributed normal random errors.
+
+Note that the uncertainty in the fitted values from the first regression is not considered when performing the second regression.
+For this reason, the variance of the coefficient estimator,  is incorrect in two-stage calculations [@burgess_review_2017;@angrist2008mostly].
+This and other observations led researchers to develop likelihood-based MR methods.
+
+### Likelihood-based methods
+Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood estimates with their many desirable properties [@fisher1925theory].
+Limited information maximum likelihood from econometrics is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite].
+Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations (3.4) and (3.5)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].
+
+Limited information maximum likelihood is sometimes called the maximum likelihood counterpart of two-stage least squares, and it yields the same causal estimate as two-stage least squares and the ratio method when used with a single SNP instrument.
+Additionally, the limited information maximum likelihood framework can accommodate  SNP instruments by replacing  with the sum .
+One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003].
+@burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation (3.6)).
+The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter,  represents the effect between the true means for the exposure and the outcome.
+In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.
+
+For each subject , the exposure and outcome values come from a bivariate normal distribution.
+The mean of the exposure distribution is assumed linear in the SNP instrument, and the mean of the outcome distribution is linear in the mean exposure [@burgess_review_2017;@jones_choice_2012].
+@burgess_improving_2012 demonstrated that the Bayesian model, with weak instruments, outperforms the limited information maximum likelihood model in terms of frequentist coverage levels.
+### Semiparametric methods
+Semiparametric instrumental variable methods, which feature parametric and nonparametric components, typically assume a parametric model connecting the exposure and outcome, but don’t impose distributional assumptions on the errors [@burgess_review_2017].
+Compared to fully parametric models, semiparametric models often are more robust to model misspecification [@burgess_review_2017;@clarke_instrumental_2012].
+We follow [@burgess_review_2017] by discussing three semiparametric strategies, generalized method of moments, continuous updating estimator, and G-estimation of structural mean models.
+Generalized method of moments can be viewed as a more flexible form of two-stage least squares that handles heteroscedastic errors and nonlinearity in the two regressions [@burgess_review_2017].
+Before specifying the model, we introduce notation.
+is the conditional expectation of  if we forced  to take value  for every subject [@pearl2009causality].
+The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation (3.7).
+
+The GMM estimate is the value of the vector  that satisfies Equation (3.7), where . @palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations (3.7).
+Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].
+We follow [@burgess_review_2017] by defining the potential outcome  as the outcome value that we would have observed had we set the exposure value  to .
+For example,  denotes the observed outcome had we set the exposure to zero instead of its observed value .
+The structural mean model for a continuous outcome is displayed in Equation (3.8).
+
+The causal effect parameter is .
+[@burgess_review_2017] derives the estimating equations for , after noting that the conditional expectation,  is independent of  and reasoning that the causal effect is that value of  that yields zero covariance between  and  (Equations (3.9)).
+
+Note that  indexes SNP instruments and ranges from  to  in Equation (3.9).
+
+## Mendelian Randomization with multiple independent SNP instruments
+While using a single SNP instrument for MR analysis can be effective, @smith_mendelian_2004 recognized that many genetic variants individually explain a small proportion of the variation in a trait, and, thus, sufficient statistical power of MR analysis with a single SNP instrument would require sample sizes in the tens of thousands [@burgess_mendelian_2013;@schatzkin2009mendelian].
+@palmer_using_2012 resolved this issue by proposing use of multiple SNP instruments in MR analysis.
+In particular, @pierce_power_2011 reasoned that a single causal estimate derived from a collection of causal SNPs would have greater precision than an estimate derived from only one SNP [@burgess_mendelian_2013].
+Recognition of complex traits’ diverse genetic architectures has thus fueled development of two-sample MR methods with multiple SNP instruments [@bowden_mendelian_2015;@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011].
+There are two important considerations for MR analysis with multiple SNP instruments.
+The first consideration is how to choose a subset of the available SNPs to serve as instruments.
+The solution to this task differs among published methods and majority of MR methods chose a set of independent SNPs as instruments [@verbanck_detection_2018;@zhu_causal_2018].
+The independent SNPs may be selected through linkage disequilibrium (LD) clumping [@purcell2007plink].
+The second consideration is how to make use of GWAS summary statistics for MR analysis, as sharing of GWAS summary statistics was encouraged and became a common practice [@nature2012asking].
+@harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised a strategy for determining a single causal effect from summary statistics for a collection of independent SNPs.
+Below, we discuss a number of MR methods that make use of multiple independent SNP instruments.
+
+## Statistical models and methods for MR with multiple instruments
+### Methods with Individual-level data
+One approach to integrating multiple SNP instruments into a MR framework is through calculation of allele scores [@burgess_use_2013].
+@burgess_use_2013 calculate an unweighted score as the number of exposure-increasing alleles in the subject’s genotypes.
+They also calculate a weighted allele score by using the exposure effects as weights.
+For example, a subject with  copies of exposure-increasing alleles for SNP  has an unweighted allele score  and a weighted allele score .
+@burgess_use_2013 found that when weights are obtained from external data or from cross-validation or jackknife approaches applied to the analysis data, the allele score functions as a single instrumental variable and greatly diminishes the bias compared to that of the two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
+@kang_instrumental_2016 developed a LASSO-based method, sisVIVE, to identify invalid SNP instruments [@tibshirani_regression_1996].
+sisVIVE’s advantage over earlier methods is that it doesn’t require the analyst to know which SNP instruments are valid.
+Instead, it requires that at least 50% of the instruments be valid.
+sisVIVE outperforms two-stage least squares in many ways and performs similar to oracle two-stage least squares.
+Simulations and data analysis results reveal that sisVIVE is robust to possibly invalid instruments.
+@windmeijer_use_2019, building on the research from @kang_instrumental_2016, implemented an adaptive LASSO-based estimator after recognizing that sisVIVE misclassifies valid SNP instruments as invalid when the invalid SNP instruments have strong effects on the exposure.
+Consistent selection of invalid SNP instruments, they found, depends on SNP instrument correlations.
+To address this issue, @windmeijer_use_2019 proposed a median estimator with consistency that doesn’t depend on SNP instrument-exposure association strength or the SNP instrument correlation structure.
+They then applied methods from @zou_adaptive_2006 to achieve a consistent estimator with the same asymptotic distribution as the oracle two-stages least squares.
+One important limitation of this thread of research is that both @kang_instrumental_2016 and @windmeijer_use_2019 require individual-level data.
+@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods.
+Central to their work is the proposal to create a new instrumental variable as a linear combination of genotype data from a collection of SNP instruments.
+They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically
+pleiotropic traits.
+For a collection of  SNPs, they choose the -vector that meets these criteria and maximizes the correlation with the exposure.
+To obtain their second estimator, @jiang_constrained_2019 apply a  penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation of the new instrumental variable.
+They then compare their method with sisVIVE and allele score methods.
+@jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are unbiased in their simulation settings.
+@spiller_detecting_2018 and @spiller_interaction-based_2020 present a method, MR-GxE, that exploits gene by environment interactions to detect and to adjust for bias due to horizontal pleiotropy.
+@spiller_interaction-based_2020 treat a SNP-covariate interaction as an instrumental variable, and, by so doing, they impose assumptions on that interaction.
+Together, these assumptions require that MR-GxE assume that horizontal pleiotropy effects are constant across the study cohort.
+The authors devise a three-step inferential procedure for two-sample MR with GWAS summary statistics.
+First, they estimate the SNP-exposure and SNP-outcome associations at a range of covariate values.
+Second, they regress the SNP-outcome estimated associations on the SNP-exposure estimated associations.
+Third, they treat the resulting slope as the causal effect estimate and the intercept as the mean horizontal pleiotropy effect.
+
+## Methods with Summary data
+The proliferation of publicly available GWAS data accelerates the development of MR methods with multiple SNP instruments that make use of GWAS summary statistics for model fitting.
+Our timeline highlights some of the many multi-instrument MR methods with GWAS summary statistics, with an increasing density of methods over time in the last three years (Figure 2).
+These methods include the inverse variance weighted MR [61], MR-Egger [51], weighted median estimation [68], Bayesian weighted MR [69], robust adjusted profile score [70], MRMix [71], CAUSE [72], and MRAID [73].
+These methods differ in their approaches to three considerations that include instrument selection (from among the available SNPs), modeling and controlling for horizontal pleiotropy, and statistical inference procedure.
+We summarize some of these MR methods that use independent or correlated SNP instruments below.
+We will cover the remaining MR methods that model horizontal pleiotropy in a later section.
+
+Figure 2: Two-sample MR with GWAS summary statistics methods publications density increases over time.
+[61] termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
+Specifically, [61] combined causal effect ratio estimates from  independent SNPs by using Equation (5.1).
+
+[68], noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
+weighted median-based estimators, building on research from [94].
+Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [95].
+Specifically, let  denote the  ordered ratio estimate statistic, from least to greatest.
+If an odd number, say , of SNP instruments is used, then the simple median estimator is defined as the  ordered ratio estimate.
+For an even number, say , number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, .
+Due to the inefficiency of the simple median estimator, [68] define the weighted median estimator.
+To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator.
+The weighted median estimator, then is the median of a distribution defined by having  as the  percentile, where  is the weight assigned to the  ordered ratio estimate.
+In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
+[68] also studied the penalized weighted median estimator.
+They defined the penalized weights, , where  is the p-value resulting from the comparison of  to a  distribution [68,96].
+The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
+While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other.
+In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [14,61–63].
+In addition, LD clumping for choosing instrument SNPs can be problematic for two reasons.
+First, the chosen SNPs may not be causal for the outcome.
+This can happen, for instance, by using a noncausal SNP that is in high LD with a causal SNP.
+Second, because many traits are polygenic or omnigenic, multiple causal SNPs may be in high LD together.
+LD clumping is likely to omit one or more of the causal SNPs, which can diminish statistical power in MR analysis.
+In support of this is the finding that including multiple correlated SNPs in transcriptome-wide association studies improves power compared to using independent SNPs only [64–67].
+Consequently, many MR methods have been recently developed to model multiple correlated instruments.
+[62] developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation.
+However, many MR methods use a set of pre-selected SNPs as instruments.
+Typically, these SNPs are selected to be statistically independent.
+This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [61].
 
@@
 likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods 
-that
+that leverage multiple SNP instruments.
 
@@
 estimates from the regressions of the outcome and exposure, respectively, on the 
-SNP instrument.
-Since the Wald method requires only the regression coefficients, it can be used with summary data.
-However, the Wald method doesn't accommodate multiple SNP instruments, which limits its direct use in the GWAS setting.
+SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with 
+summary data. However, the Wald method doesn't accommodate multiple SNP instruments, which limits its 
+direct use in the GWAS setting.
 

### 2022-04-27 reply-to-review1/mr-review.Rmd
added citations for a few more sections

@@
 Our timeline highlights some of the many multi-instrument MR methods with GWAS summary statistics, with an increasing density of methods over time in the last three years (Figure 2).
-These methods include the inverse variance weighted MR [61], MR-Egger [51], weighted median estimation [68], Bayesian weighted MR [69], robust adjusted profile score [70], MRMix [71], CAUSE [72], and MRAID [73].
+These methods include the inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted median estimation [@bowden_consistent_2016], Bayesian weighted MR [@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix [@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID [@yuan_likelihood_2021].
 These methods differ in their approaches to three considerations that include instrument selection (from among the available SNPs), modeling and controlling for horizontal pleiotropy, and statistical inference procedure.
@@
 Figure 2: Two-sample MR with GWAS summary statistics methods publications density increases over time.
-[61] termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
-Specifically, [61] combined causal effect ratio estimates from  independent SNPs by using Equation (5.1).
-
-[68], noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
-weighted median-based estimators, building on research from [94].
-Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [95].
+@burgess_mendelian_2013 termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
+Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from  independent SNPs by using Equation (5.1).
+
+@bowden_consistent_2016, noting that consistency with IVW MR is not robust to invalid instrument use, developed a method for consistent MR analysis with
+weighted median-based estimators, building on research from @han2008detecting.
+Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [@hampel1971general].
 Specifically, let  denote the  ordered ratio estimate statistic, from least to greatest.
@@
 For an even number, say , number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, .
-Due to the inefficiency of the simple median estimator, [68] define the weighted median estimator.
+Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted median estimator.
 To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator.
@@
 In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
-[68] also studied the penalized weighted median estimator.
-They defined the penalized weights, , where  is the p-value resulting from the comparison of  to a  distribution [68,96].
+@bowden_consistent_2016 also studied the penalized weighted median estimator.
+They defined the penalized weights, , where  is the p-value resulting from the comparison of  to a  distribution [@bowden_consistent_2016;@cochran1950comparison].
 The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
 While the above MR methods all use independent SNP instruments, using independent instruments may not be ideal when there are multiple causal SNPs in linkage disequilibrium with each other.
-In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [14,61–63].
-In addition, LD clumping for choosing instrument SNPs can be problematic for two reasons.
-First, the chosen SNPs may not be causal for the outcome.
-This can happen, for instance, by using a noncausal SNP that is in high LD with a causal SNP.
-Second, because many traits are polygenic or omnigenic, multiple causal SNPs may be in high LD together.
-LD clumping is likely to omit one or more of the causal SNPs, which can diminish statistical power in MR analysis.
-In support of this is the finding that including multiple correlated SNPs in transcriptome-wide association studies improves power compared to using independent SNPs only [64–67].
+In this setting, discarding some of the causal SNPs may only capture a small proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011].
 Consequently, many MR methods have been recently developed to model multiple correlated instruments.
-[62] developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation.
+@burgess_combining_2016 developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation.
 However, many MR methods use a set of pre-selected SNPs as instruments.
 Typically, these SNPs are selected to be statistically independent.
-This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [61].
+This restriction to independent SNPs is needed for valid inference in methods like standard inverse variance weighting MR [@burgess_mendelian_2013].
+
+## Assumption violations and Sensitivity Analysis
+It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations.
+Let’s first consider the exclusion restriction assumption (assumption 2 above).
+It states that the SNP instrument must affect the outcome only through the exposure.
+The causal diagram in Figure  illustrates a scenario that violates this assumption [@vanderweele_methodological_2014].
+Bias in causal effect estimation would result.
+However, if the intermediate is measured and treated as the exposure, then an unbiased estimate of causal effect is possible, since the intermediate captures the two pathways, intermediate -> outcome and intermediate -> exposure -> outcome, through which the SNP instrument affects the outcome Y [@vanderweele_methodological_2014;@textor_graphical_2011;@textor_robust_2016].
+
+Figure 3: Causal diagrams for MR.
+A.
+Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure.
+Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption.
+B.
+Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently.
+
+Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that enable assumption relaxations and has motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
+Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify the three MR assumptions with observational data.
+The goal of a sensitivity analysis is to gain insight into how the results might differ if the assumptions be violated.
+Because not all confounding variables are known or measured, assumptions 2 and 3 are not fully verifiable in MR studies.
+Recognition of this fact has inspired the development of sensitivity analysis tools.
+While we can’t assess whether a SNP instrument is associated with every confounder of the exposure-outcome association, it is possible to examine the associations of the SNP instrument with the measured covariates.
+While absence of such associations doesn’t guarantee satisfaction of the assumption, presence of SNP-covariate associations must be investigated carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].
+@burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when working with GWAS summary statistics.
+@burgess_sensitivity_2017 presented methods for both assessing the MR assumptions, to the extent possible, and performing robust analyses.
+For example, they used measured covariates to assess for possible associations with the SNP instruments.
+While they can’t rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.
+To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of C-reactive protein (CRP) levels on coronary artery disease risk with four genetic variants in the CRP gene region and 17 other genetic variants that affect coronary artery disease risk.
+They used measured covariates to probe for SNP-covariate associations, and followed it up with scatter plots and Cochran’s Q test on the causal estimates to inquire about whether the SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003].
+Additionally, @burgess_sensitivity_2017 suggested using a funnel plot, like those in the meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011].
+Additionally, Egger regression can be useful in this setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].
+@spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant pleiotropy assumption.
+MR-GENIUS is a framework for two-sample MR analysis when individual-level data are available [@tchetgen_tchetgen_genius_2021].
+@tchetgen_tchetgen_genius_2021 focused efforts on relaxing the exclusion restriction assumption and building on existing G-estimation methods to create an estimator that is robust to violations of the exclusion restriction and to additive unmeasured confounding [@robins1994correcting].
+They observe that their estimator, in some settings, reduces to that of @lewbel2012using, which is widely used in econometrics, but not in the MR literature.
+
+## Explicit modeling of horizontal pleiotropy
+Besides sensitivity analysis, several methods have been developed to directly model horizontal pleiotropy to ensure the validity of MR assumptions.
+Pleiotropy, where a single genetic variant affects multiple traits, has a long history of study in genetics and complex traits [@solovieff_pleiotropy_2013;@visscher_plethora_2016;@stearns_one_2010].
+In MR setting, horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that bypasses the exposure variable [@verbanck_detection_2018].
+The presence of horizontal pleiotropy constitutes a violation of the standard MR assumptions and can lead to biased causal effect estimates and diminished statistical power.
+@verbanck_detection_2018 documented widespread horizontal pleiotropy in GWAS.
+We discuss approaches to modeling and accounting for horizontal pleiotropy below.
+In the short time since publication of @verbanck_detection_2018, researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021].
+The first occurs via exposure-independent paths.
+The resulting horizontal pleiotropic effects are independent of the SNP-exposure relationships.
+The second type of horizontal pleiotropy manifests in the presence of unobserved exposure-outcome confounding.
+It induces correlation between horizontal pleiotropic effects and SNP-exposure effects.
+Both types of horizontal pleiotropy violate standard MR modeling assumptions and can bias causal effect estimates and can increase false discoveries [@yuan_likelihood_2021].
+Early MR analyses avoided confounding from horizontal pleiotropy by discarding instrumental SNPs that might be associated with the outcome [@verbanck_detection_2018;@zhu_causal_2018].
+More recent methods have attempted to model horizontal pleiotropy [@yuan_testing_2020;@morrison_mendelian_2020;@yuan_likelihood_2021].
+CAUSE [@morrison_mendelian_2020] and MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both types of horizontal pleiotropy.
+Modeling both types of horizontal pleiotropy is particularly challenging because the MR model likelihood often involves an integral that can’t be solved analytically.
+Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
+@bowden_mendelian_2015 adapted Egger regression, originally developed to detect bias in meta-analyses, to detect bias from horizontal pleiotropy in two-sample MR studies with GWAS summary statistics or individual-level data [@egger_bias_1997].
+Their approach is termed MR-Egger.
+They formulate their model as
+
+where  designates confounders,  and  denote the exposure and outcome, respectively,  is a SNP instrument, and  and  are the zero-mean, normally distributed random errors.
+@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument  and the effects defined in Equation (5.2) and define the new parameter  (Equation (5.3)).
+
+In Equation (5.3),  when SNP  is a valid instrument.
+MR-Egger fits the regression model in Equation (5.4).
+
+The notation in Equation (5.4) could be a little misleading. The independent and dependent variables are  and , respectively, while the regression parameters are  and .
+The slope coefficient  estimates the causal effect of the exposure on the outcome.
+Under the assumption that the SNP-exposure relationship is independent of the SNP’s pleiotropic effects, Egger’s test offers a valid test of the null causal hypothesis and consistently estimates the causal effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].
+In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity  test.
+In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate  in Equation (5.5).
+
+In Equation (5.5),  is the inverse variance of the Wald estimator, and .
+While the  test tends to be conservative in small sample sizes, its power increases with increasing sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
+
+@hemani_orienting_2017, working in the setting of two-sample MR with summary GWAS data, recognized that measurement error in outcome and exposure can give misleading results with the inferred causal arrow in the wrong direction.
+To work around this issue, they extended MR with a method that infers the causal direction between two traits.
+Specifically, @hemani_orienting_2017 adapt Steiger’s Z test for a difference in correlations [@steiger1980tests] to assess which of the two traits in an analysis has the stronger correlation with the SNP instrument.
+That with the stronger correlation is treated as the exposure.
+@hemani_orienting_2017 applied their test in a study of DNA methylation and gene expression, where either direction of causality is plausible, and found that many methylation traits cause changes in gene expression.
+@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data.
+They first calculate Wald ratio estimates for every instrumental SNP.
+They then apply smoothing to the empirical distribution of ratio estimates.
+The mode of this smoothed distribution is the simple mode-based estimate of the causal effect [@hartwig_robust_2017].
+The inverse variance-weighted mode-based estimate is obtained by weighting the empirical distribution of ratio estimates by the inverse variance of each estimate.
+@hartwig_robust_2017 compared their weighted and unweighted mode-based estimators with weighted and unweighted median-based estimators, MR Egger regression, and IVW.
+They found that their mode-based estimators demonstrated less bias and lower type I error rates than other estimators in some simulation settings.
+The mode-based estimators possessed lower statistical power to detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of MR Egger regression [@hartwig_robust_2017].
+The mode-based estimators consistently estimated the causal effect when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017].
+In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.
+@rees_robust_2019 approached the challenge of horizontal pleiotropy by developing three new methods for two-sample MR analysis with GWAS summary data: robust regression, penalized weights, and LASSO penalization.
+The first two can be viewed as modifications of MR-Egger and IVW methods.
+Together, they offer three strategies for downweighting or excluding variants with heterogeneous causal estimates.
+Recall that MR-Egger offers consistent causal effect estimates when there is no horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the “Instrument Strength Independent of Direct Effect” (InSIDE) assumption [@bowden_mendelian_2015].
+The InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP instrument-outcome effects and the SNP instrument-exposure effects.
+@rees_robust_2019 uses MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to formulate an estimator with high breakdown point and high efficiency.
+The second estimation method, with penalized weights, is much like the penalized median estimator, from above, except that the second argument in the minimum function is , instead of the  above.
+Finally the LASSO-based method fashions an estimator by applying a penalty to the objective function from MR-Egger to get Equation (6.1) as the objective function.
+
+@rees_robust_2019 studied two procedures for choosing : a heterogeneity stopping rule and a cross-validation rule.
+Taken together, these three methods offer additional contributions to a suite of sensitivity analysis tools that practitioners should consider.
+@burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core MR assumptions.
+Working in the setting of two-sample MR with GWAS data, they begin by considering a collection of candidate SNP instruments, which may contain some invalid SNP instruments.
+For each candidate SNP instrument, they calculate the Wald ratio estimate of the causal effect.
+With the collection of Wald ratio estimates, @burgess_robust_2020 reason that the set of valid SNP instruments have Wald ratio estimates that arise from a normal distribution centered on the true causal parameter with variance equal to that of the Wald ratio estimator.
+The invalid instruments, however, are assumed to follow a normal distribution centered at zero with a variance greater than that of the Wald ratio estimator.
+They then work with profile log likelihood over a grid of causal parameter values to make inferences.
+@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted profile score to accomplish statistical inference in two-sample MR analyses with GWAS summary data.
+Dividing pleiotropic effects into systematic and idiosyncratic, they model the systematic pleiotropy with random effects.
+In so doing, no SNP satisfies the exclusion restriction assumption.
+Adjusting their profile likelihood estimator from the setting without pleiotropy, @zhao_statistical_2020 obtain an estimator with consistency and asymptotic normality [@mccullagh_simple_1990].
+Idiosyncratic pleiotropy is addressed through robustification of the adjusted profile score [@huber_robust_1964].
+They then demonstrate the properties of their estimator by analyzing simulated and real data [@zhao_statistical_2020].
+@bowden_improving_2019, working with a collection of SNP instruments, reasoned that if all three MR assumptions are satisfied, then the collection of Wald ratio estimates should be homogeneous; thus, any departures from homogeneity may signal a violation of one or more assumptions.
+They reported modified weights, in addition to studying weights derived from first order and second order approximations to the variance.
+Their modified weights lead to an estimator that is useful for quantifying heterogeneity and detecting outliers.
+@cho_exploiting_2020 developed the MR-TRYX framework for exploiting horizontal pleiotropy to identify alternative causal pathways.
+They did this by positing that, for a collection of putative SNP instruments, any heterogeneity in the Wald ratio estimates may be due to horizontal pleiotropy.
+Once they identify a SNP with outlying ratio estimates, they search GWAS results by querying the MR-Base database to identify complex traits (in other studies) that associate with the outlying SNP [@hemani_mr-base_2018].
+They follow up promising trait associations to account for their observations with multiple causal pathways.
+@foley_clustering_2021 recognized that in some MR studies, the Wald ratio estimates for different SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms and may identify the SNPs that are involved in them.
+They devised a method to detect these clusters in the collection of Wald ratio estimates and found SNP instruments that point to distinct causal pathways of biological importance [@foley_clustering_2021].
+@cheng_mr-ldp_2020 sought a probabilistic model that accounts for both linkage disequilibrium among SNP instruments and horizontal pleiotropy.
+They termed their method MR-LDP.
+Using an approach from @zhu_bayesian_2017, @cheng_mr-ldp_2020 derived an approximate likelihood for their regression models of the exposure and outcome on the SNP instruments.
+Incorporation of a random effect into their models accounted for the variance in causal effect estimates due to horizontal pleiotropy.
+They situate inference within an empirical Bayesian framework and present a parameter-expanded variational Bayes expectation-maximization algorithm for estimation [@liu_parameter_1998].
+By jointly modeling the distribution of GWAS summary statistics and causal effects, @cheng_mr-ldp_2020 efficiently accommodates multiple SNP instruments in linkage disequilibrium.
+One limitation of MR-LDP
+is its reliance on the InSIDE assumption and resulting inability to accommodate correlated horizontal pleiotropy.
+@zhu_iterative_2021 present IMRP, a method for causal effect estimation and horizontal pleiotropy detection.
+They describe a test to distinguish vertical pleiotropy from horizontal pleiotropy.
+By combining this test with two-sample MR analysis of GWAS summary statistics, they develop an iterative procedure for estimating the causal effect of exposure on outcome and test for presence of horizontal pleiotropy in the SNP instruments.
+The iterative procedure alternates between 1) updating the estimate of the causal effect and 2) testing for horizontal pleiotropy with the included SNP instruments.
+At each test for horizontal pleiotropy, some SNP instruments may be discarded if they demonstrate evidence of horizontal pleiotropic effects.
+The procedure ends when there remain only SNP instruments that don’t exhibit horizontal pleiotropy.
+In the authors’ simulations and data analysis sections, their new method performs about as well as GSMR, IVW, and MR-PRESSO in the settings with and without balanced pleiotropy and with and without satisfaction of the InSIDE assumption [@zhu_iterative_2021].
+Finally, the authors note that IMRP is three orders of magnitude faster than simulations-based MR-PRESSO.
+@grant_efficient_2020 build on research from @lin2015regularization to present a method that applies a  penalty to coefficients of covariates in two-sample MR with GWAS summary statistics.
+Importantly, their method doesn’t penalize the SNP-exposure association.
+Instead, it shrinks towards zero those coefficients for covariates that demonstrate little pleiotropy.
+In so doing, they extend work from @burgess_multivariable_2015 and @burgess_re_2015 to get a causal effect estimator that needs no valid SNP instruments and is robust to measured vertical pleiotropy and more statistically efficient than previous MR methods.
+@xu_robust_2021 consider two classes of horizontal pleiotropy with their method, MRCIP.
+They partition horizontal pleiotropic effects into correlated or idiosyncratic horizontal pleiotropy.
+Idiosyncratic horizontal pleiotropy is that with a large effect size, while correlated horizontal pleiotropy, like its definition elsewhere in our manuscript, refers to correlation between the effects of the SNP instruments on the exposure and the effects of the SNP instruments on the outcome.
+MRCIP relaxes both the exclusion restriction assumption and the InSIDE assumption by directly modeling the correlated horizontal pleiotropy with random effects.
+Furthermore, to accommodate the idiosyncratic pleiotropy, the authors downweight SNP instruments that demonstrate strong direct effects on the outcome.
+They find that their MRCIP provides valid causal inference even when there are no valid SNP instruments (i.e., when all SNP instruments are invalid) and when the InSIDE assumption is violated [@xu_robust_2021].
+Their PRW-EM algorithm adds a reweighting step to the traditional expectation maximization algorithm to estimate the causal effect parameter in MR analysis [@dempster1977maximum].
+Finally, they compare MRCIP to other MR methods with both simulated and real data sets.
+@xu_robust_2021, like the developers of MR-RAPS, assume that the horizontal pleiotropy is balanced, i.e., that the mean horizontal pleiotropic effect is zero.
+One direction for future study is to augment MRCIP to permit unbalanced horizontal pleiotropy.
+@iong_latent_2020 build on other research using mixture models in the two-sample MR setting with GWAS summary statistics.
+Like MR-Clust, MR-PATH, as the method from @iong_latent_2020 is called, works with the Wald ratio estimates from multiple SNP instruments.
+Coining the term “mechanistic heterogeneity” to refer to the distribution of Wald ratio estimates due to SNP involvement in distinct causal pathways, @iong_latent_2020 specify a -component normal mixture model.
+They use an expectation maximization algorithm to fit the model parameters (the component weights and the parameters for each of the  normal distributions) and suggest a Bayesian information criterion-based approach to choosing  [@schwarz1978estimating].
+@wang_weakinstrument_2021 propose causal effect test statistics that are robust under weak instrument asymptotics [@staiger1994instrumental].
+Specifically, they extend three econometric methods to two-sample MR with GWAS summary statistics [@anderson1949estimation;@kleibergen_pivotal_2002;@moreira_conditional_2003] and study the theoretical properties and the performance in applications.
+@wang_weakinstrument_2021 extend their methods to demonstrate conditions under which their estimators are equivalent to those in works from @zhao_statistical_2020, @windmeijer_use_2019, and @bowden_improving_2019.
+@bucur_inferring_2020 developed BayesMR, a Bayesian framework for two-sample MR with individual-level data and discuss an approximation for use with
+GWAS summary statistics.
+Unlike many MR methods, BayesMR both aims to control for horizontal pleiotropy and to quantitatively assess the possibility of reverse causation, where the putative exposure is actually caused by the putative outcome.
+For  independent SNP instruments, @bucur_inferring_2020 permit each SNP instrument to have both a direct effect on the exposure and a horizontally pleiotropic direct effect on the outcome.
+Instrument effects and horizontally pleiotropic effects are assumed independent.
+Note that this is the Bayesian analog of the InSIDE assumption.
+After specifying the model, they present a nested sampling scheme [@skilling2006nested;@handley2015polychord1;@handley2015polychord] that simultaneously computes the model evidence for both causal directions (i.e., exposure causes outcome and outcome causes exposure) and acquires samples from the posterior distribution for Bayesian statistical inferences.
+@schmidt_mendelian_2018 present BMRE, a Bayesian implementation of MR Egger.
+They apply weakly informative priors to the horizontal pleiotropy effect estimator in efforts to increase statistical power to detect a nonzero slope, i.e., the coefficient of the SNP instrument.
+They develop their method for two-sample MR with individual-level data.
+Simulations demonstrate that BMRE outperforms the MR Egger method while maintaining type I error rates in the examined scenarios.
+The authors emphasize a potential role for BMRE in MR sensitivity analysis.
+@sanderson_examination_2019 found that multivariable MR, where multiple exposures jointly cause an outcome, consistently estimates the effect of the exposures on the outcome.
+They examined multivariable MR with both individual-level data and GWAS summary statistics.
+@sanderson_examination_2019 also develop a generalized version of Cochran’s Q test for quantifying SNP instrument strength and validity.
+However, this test requires knowledge of the covariances between the effects of the SNP instruments on the exposures.
+While these covariances could be estimated from individual-level data, such data are often unavailable in the GWAS setting.
+Importantly, the authors note that the estimands for multivariable MR nad MR differ; multivariable MR estimates the direct causal effects of each exposure on the outcome, while MR estimates the total causal effect of an exposure on the outcome.
+MR-link, reported by @van_der_graaf_mendelian_2020, aims to account for unobserved pleiotropy and linkage disequilibrium by using individual-level data on the outcome variable and summary statistics for the exposure.
+MR-link uses a three-step procedure to estimate causal effects.
+    1. Use GCTA-COJO to obtain SNP instruments [@genetic_investigation_of_anthropometric_traits_giant_consortium_conditional_2012]
+
+    2. Prune all SNPs in linkage disequilibrium with the SNP instruments to obtain a set of “tag” SNPs with correlations less than 0.95
+
+    3. Solve for the causal effect estimate with ridge regression of the outcome on the matrix resulting from concatenation of the SNP instrument genotypes matrix and the tag SNPs genotypes matrix.
+Note that in step 3, ridge regression is preferred to ordinary least squares due to the limited statistical power of ordinary least squares in this setting.
+Presumably, this is due to collinearity in the matrix of tag SNP genotypes.
+
+## Correlated horizontal pleiotropy
+
+The second type of horizontal pleiotropy, from above, is sometimes called “correlated horizontal pleiotropy.” Correlated horizontal pleiotropy occurs when a SNP affects both the exposure and a variable that confounds the relationship between exposure and outcome (Figure 3B).
+CAUSE and MRAID are two methods that aim to model correlated horizontal pleiotropy [@morrison_mendelian_2020;@yuan_likelihood_2021].
+Their approaches to modeling correlated horizontal pleiotropy differ in important ways that ultimately affect performance.
+CAUSE allows for a proportion of SNPs to exhibit correlated pleiotropy, which @morrison_mendelian_2020 models as an effect on a shared, unobserved factor.
+The remaining SNPs are independent of this unobserved factor.
+Every SNP can have a nonzero pleiotropic effect on the outcome, and these pleiotropic effects are uncorrelated with the variant effects on the exposure.
+The model, then, is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on adaptive shrinkage from @stephens_false_2016.
+Finally, @morrison_mendelian_2020 compares two models to determine whether the GWAS summary statistics are consistent with a causal effect of the
+exposure on the outcome.
+Specifically, they estimate the difference in the expected log pointwise posterior density for the model with causal effect fixed at zero and the model that permits a nonzero causal effect [@vehtari_practical_2017].
+@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is biased.
+CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].
+These observations motivated @yuan_likelihood_2021 to develop a new method, MRAID.
+MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021].
+We focus on the approach that uses GWAS summary statistics.
+Two equations are central to MRAID modeling (Equations (7.1) and (7.2)).
+
+The estimated marginal effects for the  SNP instruments on the exposure are denoted by , while  represents the estimated effects of the same SNPs on the outcome.
+The  by  SNP instrument correlation matrices for the exposure and outcome are written as  and , respectively.
+and  can be estimated with 1000 Genomes Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry [@the_1000_genomes_project_consortium_1000_2012].
+The -vector error terms,  and , follow multivariate normal distributions with mean zero and variances  and , respectively.
+@yuan_likelihood_2021 then constructs a Gibbs sampler to make likelihood-based inferences [@geman1984stochastic].
+To facilitate computations, the investigators make several assumptions about the collection of SNPs:
+    1. Relatively small proportion of SNPs have nonzero effects on the exposure
+
+    2. A relatively small proportion of SNPs demonstrate horizontal pleiotropy
+
+    3. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs
+
+    4. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.
+By encoding these assumptions in prior distributions, @yuan_likelihood_2021 enables the inferential procedures to accommodate the observed data in the context of the assumptions on SNPs.
+
+## Practice recommendations
+
+## Binary, Count, and Time-to-event outcomes and exposures
+
+While we’ve focused above on continuous outcomes and continuous exposures, some data are more naturally treated as binary, count, or time-to-event variables.
+MR methods for non-continuous outcomes and exposures is an active area of research.
+@burgess_improving_2012 and @allman_novel_2021 developed MR strategies for binary exposures and binary outcomes.
+@burgess_improving_2012 extended methods from @wald_fitting_1940 by fitting two logistic regressions, one for the outcome and one for the exposure.
+They then calculated the causal effect estimator as the ratio of the two logistic regression coefficient estimators.
+@allman_novel_2021 devised a new MR method for binary outcome and binary exposure by drawing on @gauvreau_analysis_1997 and treating the exposure and outcome as correlated binary random variables.
+They implemented an iterative optimization algorithm to simultaneously infer the causal effect parameter and other parameters.
+Researchers have also leveraged other generalized linear models for analysis in MR studies [@nelder1972generalized].
+However, this area is relatively unexplored.
+@hazewinkel_mendelian_2022 treated hospitalizations as a count variable and modeled it with quasi-Poisson methods.
+@allman_mendelian_2022 characterized statistical properties, including bias and power, with simulations of count random variables and binary random variables.
+With growing interest in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].
+Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017].
+@guo_body_2017, for example, studied post-diagnosis survival time in women with breast cancer.
+They treated survival as a time-to-event outcome, which enabled them to model it with a proportional hazards model [@cox_regression_1972] in two-stage MR with individual-level data.
+Similarly, @he_causal_2018, in a study of cardiovascular disease, applied proportional hazards regression when examining time to stroke as an outcome variable.
+In very recent work, @tchetgen_tchetgen_genius_2021 reported methods, MR GENIUS, that are robust to violations of the exclusion restriction assumption (under many assumed MR data generating processes).
+They considered binary, continuous, and time-to-event outcomes within their framework.
+We are unaware of MR studies that treat the exposure variable as a time-to-event, but it’s conceivable that future research will consider time-to-event traits as exposures in MR analyses.
 

### 2022-04-27 reply-to-review1/mr-review.Rmd
updated bib file to match Rmd citations

@@
 date: "`r lubridate::now()`"
-bibliography: exported.bib
-csl: elsevier-with-titles.csl
+bibliography: ../exported.bib
+csl: ../elsevier-with-titles.csl
 link-citations: yes
@@
 library(magrittr)
-dat <- readxl::read_xlsx(path = "data/publication-counts.xlsx", sheet = "counts")
+dat <- readxl::read_xlsx(path = "../data/publication-counts.xlsx", sheet = "counts")
 dat %>%
@@
 
-ggplot2::ggsave("figures/mr-gwas-counts.png")
+ggplot2::ggsave("../figures/mr-gwas-counts.png")
 
@@
 They follow up promising trait associations to account for their observations with multiple causal pathways.
-@foley_clustering_2021 recognized that in some MR studies, the Wald ratio estimates for different SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms and may identify the SNPs that are involved in them.
-They devised a method to detect these clusters in the collection of Wald ratio estimates and found SNP instruments that point to distinct causal pathways of biological importance [@foley_clustering_2021].
+@fang2021evaluating recognized that in some MR studies, the Wald ratio estimates for different SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms and may identify the SNPs that are involved in them.
+They devised a method to detect these clusters in the collection of Wald ratio estimates and found SNP instruments that point to distinct causal pathways of biological importance [@fang2021evaluating].
 @cheng_mr-ldp_2020 sought a probabilistic model that accounts for both linkage disequilibrium among SNP instruments and horizontal pleiotropy.
@@
 
-# Mendelian Randomization with one Instrument
+## Recent Findings
+### Omnigenic MR
+Recent research on the omnigenic hypothesis, which posits that every SNP’s effect on a trait is nonzero, has informed MR methods development [@boyle2017].
+@wang_mendelian_nodate recognized the limitations of previous MR methods in the context of the omnigenic hypothesis.
+In response, @wang_mendelian_nodate developed a MR method that uses all genome-wide SNPs as instruments.
+Using GWAS summary statistics as inputs, their method relies on a composite likelihood framework for scalable computation and allows for horizontal pleiotropy.
+@wang_mendelian_nodate used extensive simulations, including those with model misspecifications, to characterize their method’s statistical power and robustness.
+Finally, they applied the new method to identify multiple complex traits that affect coronary artery disease and asthma.
+These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.
+### Sample overlap in biobanks
+The issue of overlapping samples in the two-sample MR with GWAS summary data has recently received attention in the published literature.
+@fang2021evaluating, working with data from the UK Biobank, found that SNP instruments derived from overlapping samples explained a higher proportion of the variance compared to those derived from non-overlapping samples.
+They argue that block jackknife resampling MR enables causal inference while avoiding bias due to overlapping samples.
+@sadreev2021navigating analyzed 2,514 traits from the UK Biobank and evaluated the impact of winner’s curse on MR analysis.
+Winner’s curse, they found, amplifies weak instrument bias without inflating the false discovery rate.
+This finding led the authors to design a pseudoreplication process that reduces bias in MR studies.
+@berzuini_bayesian_2020 proposed a Bayesian approach for one-sample MR analysis where SNP instruments are permitted to exhibit pleiotropic effects on the outcome.
+Working in the setting of one-sample MR with individual-level data, they construct their model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects.
+Their method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP instruments) and permits correlated SNP instruments, unlike many frequentist methods.
+Posterior inference proceeds by Markov chain Monte Carlo sampling.
+They also elaborate their model for a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.
 
-We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where 
-there is a single genetic marker and a single outcome variable. Mendelian randomization with one 
-instrument uses a single genetic marker 
-as the proxy ("instrument" or "instrumental variable") for the "exposure", a complex trait, and asks 
-whether variation in the genotypes at the genetic marker causes variation in the 
-"outcome" variable, which is typically a second complex trait. When the genetic marker 
-is a SNP on an autosome, the cohort is partitioned into three 
-classes, two homozygote classes and a single heterozygote class, based on genotypes 
-for that marker. When the three assumptions of mendelian randomization are satisfied, 
-estimation of the causal effect of the exposure on the outcome is possible.
+### Bi-directional causal inference with MR to account for reverse causation
+Reverse causation occurs when the outcome at an early time point has a causal effect on the exposure variable.
+While reverse causation is often assumed to not occur in MR study designs, @burgess_are_2021 points out several scenarios where this assumption may not hold [@davey_smith_mendelian_2003].
+In recognition of this observation, @wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong instruments.
+Using GWAS summary statistics as inputs, GRAPPLE extends MR-RAPS and can detect presence of horizontally pleiotropic pathways, infer the causal direction, and perform multivariable MR.
+Central to the GRAPPLE framework is the observation that a horizontally pleiotropic pathway often gives rise to an additional mode in the profile likelihood.
+GRAPPLE uses the presence of multiple profile likelihood modes to diagnose horizontal pleiotropy with effects that are grouped by pathway.
+GRAPPLE facilitates identification of SNP instruments that contribute to the additional profile likelihood modes.
+A researcher using GRAPPLE may then recognize a confounding factor for each mode.
+With additional GWAS summary data with traits that reflect the confounding factors, GRAPPLE can fit multivariable MR models when the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].
 
-The outcome variable can be binary, a count, a time to event, or a continuous variable. For brevity and simplicity, 
-we focus on continuous outcomes before considering other classes of outcome variables in a later section.
+## Summary and Outlook
 
-## Assumptions of Mendelian Randomization
-
-Three assumptions are central to MR. Satisfying them enables inference of 
-causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
-Because assumptions 2 and 3, which involve absence of certain types of confounding, can't be verified, 
-practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
-way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014].
-
-The first assumption states that the exposure is associated with the outcome. The second requires 
-independence, conditional on the exposure and all measured and unmeasured confounders, between the 
-genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
-independent of all (measured and unmeasured) confounders of the relationship between exposure and 
-outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].
-
-If the first assumption, about the association between the genetic marker and the exposure, holds, 
-but this association is weak, then an amplification of biases, such as those due to violations of 
-assumptions 2 and 3, may result [@vanderweele_methodological_2014]. @burgess_avoiding_2011, in analyses 
-of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
-the strength of association between the SNP and exposure. We later discuss, in the context of 
-methods for multiple SNP instruments, how this finding, coupled with the recognition of 
-the polygenicity of many complex traits, has motivated methods to work with large numbers of weak SNP instruments [@ye_debiased_2021;@chao_consistent_2005].
-
-The second MR assumption, often termed the exclusion restriction assumption, can also be stated as 
-the need for the genetic marker to affect the outcome only through the exposure. In a 
-causal diagram, this means that the only path from the marker to the outcome is that 
-containing the exposure.
-
-The third assumption ensures that the genetic marker is
-independent of all confounders of the exposure-outcome relationship. Like the exclusion restriction
-assumption, we can't verify this assumption with observational data, so analysts need to 
-use tools like sensitivity analysis to ensure validity of findings.
-
-When the three MR assumptions hold, one may use instrumental variable statistical methods, 
-with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
-causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
-
-To understand the origin of the term "Mendelian randomization", it is helpful to 
-recall Mendel's inheritance laws [@mendel1865versuche]. Under the assumption that alleles segregate 
-randomly from parent to offspring, the offspring genotypes are unlikely to be 
-associated with confounders of the exposure-outcome relationship. Additionally,
-reverse causation is unanticipated since germ-line genotypes are 
-fixed at conception and, thus, precede realization of other observed variables. Reverse causation 
-occurs when the outcome at an early time point has a causal effect on the exposure variable. While 
-reverse causation is often assumed to not occur in MR study designs, @burgess_are_2021 points out 
-several scenarios where this assumption may not hold [@davey_smith_mendelian_2003].
-
-## Statistical models and methods for MR with one instrument
-
-Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, 
-likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods 
-that leverage multiple SNP instruments.
-
-### Ratio of coefficients method
-
-The ratio of coefficients method, also known as the Wald method, estimates the 
-causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940]. 
-For a continuous outcome, the causal effect estimator, $\hat\gamma_{Wald}$ is defined in Equation \@ref(eq:wald).
-
-\begin{equation}
-  \hat\gamma_{Wald} = \frac{\hat\beta_{YZ}}{\hat\beta_{XZ}}
-  (\#eq:wald)
-\end{equation}
-
-In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope 
-estimates from the regressions of the outcome and exposure, respectively, on the 
-SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with 
-summary data. However, the Wald method doesn't accommodate multiple SNP instruments, which limits its 
-direct use in the GWAS setting.
-
-### Two-stage methods
-
-A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
-outcome, one may perform "two-stage least squares", which involves two linear regressions. 
-First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)).
-We denote each subject's exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope.
-$z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.
-
-The resulting 
-fitted values for the exposure variable from Equation \@ref(eq:lm1) are the independent variable in the second linear regression, 
-where the outcome is the dependent variable. The causal effect estimate, then, is the 
-coefficient $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).
-Note also that the random errors in Equation \@ref(eq:lm2) are assumed independent of those in Equation \@ref(eq:lm1).
-Like the $\epsilon_i$ in Equation \@ref(eq:lm1), the $\tau_i$ are assumed independent and identically distributed normal random errors.
-
-\begin{equation}
-  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
-  (\#eq:lm1)
-\end{equation}
-
-\begin{equation}
-  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
-  (\#eq:lm2)
-\end{equation}
-
-Note that the uncertainty in the fitted values from 
-the first regression is not considered when performing the second regression.
-For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is 
-incorrect in two-stage calculations. This and other observations led researchers to develop 
-likelihood-based MR methods.
-
-### Likelihood-based methods
-
-Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
-estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
-is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. We consider it below before introducing a Bayesian method.
-
-Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations \@ref(eq:liml1) and \@ref(eq:liml2)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].
-
-\begin{equation}
-  x_i = \alpha_0 + \alpha_1 z_{i1} + \epsilon_{Xi}
-  (\#eq:liml1)
-\end{equation}
-
-\begin{equation}
-  y_i = \beta_0 + \beta_1 x_i + \epsilon_{Yi}
-  (\#eq:liml2)
-\end{equation}
-
-Limited information maximum likelihood is sometimes called the maximum likelihood 
-counterpart of two-stage least squares, and it yields the same causal estimate as 
-two-stage least squares and the ratio method when used with a single SNP 
-instrument. Additionally, the limited information maximum likelihood framework can accommodate $K > 1$ SNP instruments by replacing $\alpha_1 z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.
-
-One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003].
-@burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation \@ref(eq:b1)).
-The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter, $\beta_1$ represents the effect between the true means for the exposure and the outcome.
-In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.
-
-\begin{eqnarray}
-\begin{pmatrix}
-X_i \\
-Y_i
-\end{pmatrix} \sim N_2\left(\begin{pmatrix} \xi_i\\
-\eta_i
-\end{pmatrix}, \Sigma
-\right)\nonumber \\
-\xi_i = \alpha_0 + \alpha_1 z_{i1}\nonumber \\
-\eta_i = \beta_0 + \beta_1 \xi_i
-  (\#eq:b1)
-\end{eqnarray}
-
-For each subject $i$, the exposure and outcome values come from a bivariate normal 
-distribution. The mean of the exposure distribution is assumed linear in 
-the SNP instrument, and the mean of the outcome distribution is linear in the 
-mean exposure [@burgess_review_2017;@jones_choice_2012]. 
-@burgess_improving_2012 demonstrated that the Bayesian model, with weak 
-instruments, outperforms the limited information maximum likelihood model in terms 
-of frequentist coverage levels.
-
-### Semiparametric methods
-
-Semiparametric instrumental variable methods, which feature parametric and nonparametric components, 
-typically assume a parametric model connecting the exposure and outcome, but don't 
-impose distributional assumptions on the errors [@burgess_review_2017]. Compared to 
-fully parametric models, semiparametric models often are more robust to model 
-misspecification [@burgess_review_2017;@clarke_instrumental_2012]. We follow 
-@burgess_review_2017 by discussing three semiparametric strategies, generalized 
-method of moments, continuous updating estimator, and G-estimation of structural 
-mean models.
-
-Generalized method of moments can be viewed as a more flexible form of two-stage 
-least squares that handles heteroscedastic errors and nonlinearity in the two 
-regressions [@burgess_review_2017]. Before specifying the model, we introduce 
-notation. $\mathbb{E}(Y|\text{do}(X = x))$ is the conditional expectation of $Y$ if 
-we forced $X$ to take value $x$ for every subject [@pearl2009causality].
-
-The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation \@ref(eq:gmm).
-
-\begin{eqnarray}
-  \sum_i (y_i - f(x_i;\beta)) = 0\nonumber \\
-  \sum_i z_i(y_i - f(x_i;\beta)) = 0
-  (\#eq:gmm)
-\end{eqnarray}
-
-The GMM estimate is the value of the vector $\beta$ that satisfies Equation \@ref(eq:gmm), where $f(x_i; \beta) = \mathbb{E}(Y|\text{do}(X = x))$.
-@palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations \@ref(eq:gmm).
-
-Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].  
-We follow @burgess_review_2017 by defining the potential outcome $Y(x)$ as the outcome value that we would 
-have observed had we set the exposure value $X$ to $x$. For example, $Y(0)$ denotes the observed outcome had 
-we set the exposure to zero instead of its observed value $x$. The structural mean model for a continuous 
-outcome is displayed in Equation \@ref(eq:smm).
-
-\begin{equation}
-\mathbb{E}(Y(x) - Y(0) | X = x, Z = z) = \beta_1 x
-  (\#eq:smm)
-\end{equation}
-
-The causal effect parameter is $\beta_1$. @burgess_review_2017 derives the estimating equations for 
-$\beta_1$, after noting that the conditional expectation, $\mathbb{E}(Y(0)| X = x, Z = z)$ is 
-independent of $Z$ and reasoning that the causal effect is that value of $\beta_1$ that yields zero 
-covariance between $Z$ and $\mathbb{E}(Y(0)| X = x, Z = z)$ (Equations \@ref(eq:smm1)).
-
-\begin{equation}
-\sum_i (z_{ik} - \bar z_{\cdot k})(y_i - \beta_1 x_i) = 0
-  (\#eq:smm1)
-\end{equation}
-
-Note that $k$ indexes SNP instruments and ranges from $1$ to $K$ in Equation \@ref(eq:smm1).
-
-# Assumption violations and Sensitivity Analysis
-
-It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations.
-Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
-must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
-a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
-if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
-possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
-affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].
-
-```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X.
-Because W affects Y through a pathway that doesn't involve exposure X, this scenario violates the exclusion restriction assumption."}
-
-library(dagitty)
-library(ggplot2)
-
-d1 <- dagitty('dag {
-bb="0,0,1,1"
-G [pos="0,0"]
-W [pos="0.5,0"]
-X [exposure, pos="1.0,0"]
-Y [outcome, pos="1.5,0"]
-G -> W
-W -> X
-W -> Y [pos="1,1"]
-X -> Y
-}')
-
-plot(d1)
-
-```
-
-```{r, dag2, fig.cap = "Scenario where the genetic variant G affects exposure X, which affects Y. Unobserved confounder U affects both X and Y."}
-d1 <- dagitty('dag {
-bb="0,0,1,1"
-G [pos="0.0,0.0"]
-U [pos="0.75,0.5"]
-X [exposure, pos="0.5,0.0"]
-Y [outcome, pos="1.0,0.0"]
-G -> X
-U -> X
-U -> Y
-X -> Y
-}'
-)
-plot(d1)
-
-```
-
-Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that 
-enable assumption relaxations and has motivated use of sensitivity analysis to quantify the impact of 
-possible assumption violations.
-
-Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
-the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
-into how the results might differ if the assumptions be violated.
-Because not all confounding 
-variables are known or measured, assumptions 2 and 3 are not fully verifiable 
-in MR studies. Recognition of this fact has inspired the development of 
-sensitivity analysis tools. While we can't assess whether a SNP instrument is 
-associated with every confounder of the exposure-outcome association, it is 
-possible to examine the associations of the SNP instrument with the measured 
-covariates. While absence of such associations doesn't guarantee satisfaction 
-of the assumption, presence of SNP-covariate associations must be investigated
-carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].
-
-@burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when 
-working with GWAS summary statistics. @burgess_sensitivity_2017 presented methods for both assessing the
-MR assumptions, to the extent possible, and performing robust analyses. For example, 
-they used measured covariates to assess for possible associations with the SNP 
-instruments. While they can't rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.
-
-To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of 
-C-reactive protein (CRP) levels on coronary artery disease risk with four genetic 
-variants in the CRP gene region and 17 other genetic variants that affect coronary 
-artery disease risk. They used measured covariates to probe for SNP-covariate 
-associations, and followed it up with scatter plots and Cochran's Q test on the causal 
-estimates to inquire about whether the 
-SNP instruments all identify the same causal parameter [@greco_m_detecting_2015;@small_sensitivity_2007;@higgins_measuring_2003]. 
-Additionally, @burgess_sensitivity_2017 suggested using a funnel plot, like those in the 
-meta-analysis literature, to visualize possible evidence of directional pleiotropy, where the average pleiotropic effects of the SNP instruments is nonzero [@sterne_recommendations_2011].
-Additionally, Egger regression can be useful in this
-setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].
-
-# Mendelian Randomization with multiple SNP instruments
-
-Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
-methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011;@bowden_mendelian_2015]. 
-Choosing a subset of the available SNPs to serve as instruments is one issue that arises 
-in these studies. The solution to this task differs among published methods. Early methods chose a 
-set of independent SNPs [@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-2;@zeng_causal_2019-1]. However, this approach can be problematic when there are multiple causal 
-SNPs in linkage disequilibrium with each other. In this setting, discarding some of the causal SNPs may only capture a small
-proportion of trait variance explained by the exposure and lead to a power loss in MR [@burgess_review_2017;@burgess_mendelian_2013;@burgess_combining_2016;@burgess_bias_2011]. In support of
-this is the finding that including multiple correlated SNPs in transcriptome-wide association
-studies improves power compared to using independent SNPs only [@liu2021multi;@gusev_integrative_2016;@yuan_testing_2020;@zeng_non-parametric_2017].
-
-Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
-median estimation [@bowden_consistent_2016], Bayesian weighted MR 
-[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
-[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
-[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
-their approaches to three considerations:
-
-1. instrument selection (from among the available SNPs)  
-1. modeling and controlling for horizontal pleiotropy  
-1. statistical inference procedure
-
-Many MR methods use a set of pre-selected SNPs as instruments. Typically, these SNPs are selected to 
-be statistically independent. This restriction to 
-independent SNPs is needed for valid inference in methods like inverse variance weighting MR [@burgess_mendelian_2013]. The independent SNPs may be 
-selected through linkage disequilibrium (LD) clumping [@purcell2007plink].
-
-However, LD clumping for choosing instrument SNPs can be problematic for several 
-reasons. First, the chosen SNPs may not be causal for the outcome. This can 
-happen, for instance, by using a noncausal SNP that is in high LD with a 
-causal SNP. Second, because many traits are polygenic or omnigenic, multiple 
-causal SNPs may be in high LD together. LD clumping is likely to omit one or more
-of the causal SNPs, which can diminish statistical power in MR analysis.
-
-Horizontal pleiotropy occurs when a SNP instrument affects the outcome through at least one pathway that 
-bypasses the exposure variable [@verbanck_detection_2018]. It constitutes a violation of 
-the standard MR assumptions and can lead to biased 
-causal effect estimates and diminished statistical power. We discuss approaches to 
-modeling and accounting for horizontal pleiotropy below.
-
-@smith_mendelian_2004 recognized that many genetic variants individually explain a 
-small proportion of the variation in a trait, and, thus, sufficient statistical 
-power would require sample sizes in the tens of thousands 
-[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
-resolved this issue by proposing use of multiple SNP instruments in a single MR 
-analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
-a collection of causal SNPs would have greater precision than an estimate derived 
-from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
-summary statistics was encouraged as a best practice [@nature2012asking]. 
-@harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of 
-the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
-a strategy for determining a single causal effect from summary statistics for a 
-collection of SNPs.
-
-Below, we discuss a number of MR methods that use multiple 
-SNP instruments. The overall trend appears to be in the direction of increasingly sophisticated 
-modeling of horizontal pleiotropy (and correlated horizontal pleiotropy) while
-attempting to maintain scalability and computational efficiency.
-
-## Statistical models and methods for MR with multiple instruments
-
-### Methods with Individual-level data
-
-One approach to integrating multiple SNP instruments into a MR framework is
-through calculation of allele scores [@burgess_use_2013]. @burgess_use_2013 
-calculate an unweighted score as the number of exposure-increasing alleles in 
-the subject's genotypes. They also calculate a weighted allele score by using 
-the exposure effects as weights. For example, a subject with $g_k$ copies of 
-exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_{k}$ 
-and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$. 
-@burgess_use_2013 found that when weights are obtained from external data or from 
-cross-validation or jackknife approaches applied to the analysis data, the allele score 
-functions as a single instrumental variable and greatly diminishes the bias compared to that of the 
-two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
-
-@kang_instrumental_2016 developed a LASSO-based method, sisVIVE, to 
-identify invalid SNP instruments [@tibshirani1996regression]. sisVIVE's 
-advantage over earlier methods is that it doesn't require the analyst 
-to know which SNP instruments are valid. Instead, it requires that at 
-least 50\% of the instruments be valid. sisVIVE outperforms two-stage 
-least squares in many ways and performs similar to oracle two-stage 
-least squares. Simulations and data analysis results reveal that 
-sisVIVE is robust to possibly invalid instruments.
-
-@windmeijer_use_2019, building on the research from 
-@kang_instrumental_2016, implemented an adaptive LASSO-based estimator 
-after recognizing that sisVIVE misclassifies valid SNP instruments as 
-invalid when the invalid SNP instruments have strong effects on the 
-exposure. Consistent selection of invalid SNP instruments, they found, 
-depends on SNP instrument correlations. To address this issue, @windmeijer_use_2019 
-proposed a median estimator with consistency that doesn't 
-depend on SNP instrument-exposure association strength or the SNP instrument 
-correlation structure. They then applied methods from @zou2006adaptive to 
-achieve a consistent estimator with the same asymptotic distribution as the oracle 
-two-stages least squares. One important limitation of this thread of research is that
-both @kang_instrumental_2016 and @windmeijer_use_2019 require individual-level data.
-
-@jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods. Central to their
-work is the proposal to create a new instrumental variable as a linear combination of genotype data from 
-a collection of SNP instruments. They require that the new instrumental variable be standardized to have norm 1 and be orthogonal to potentially vertically  
-pleiotropic traits. For a collection of $K$ SNPs, they choose the $K$-vector that meets these criteria and 
-maximizes the correlation with the exposure. To obtain their second estimator, @jiang_constrained_2019 apply a 
-$\ell_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation 
-of the new instrumental variable. They then compare their method with sisVIVE and allele score methods. 
-@jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are 
-unbiased in their simulation settings.
-
-@spiller_detecting_2018 and @spiller_interaction-based_2020 present a method, MR-GxE, that exploits gene by 
-environment interactions to detect and to adjust for bias due to horizontal pleiotropy. They leverage the insight 
-that a SNP instrument-covariate interaction that induces differences in SNP instrument-exposure associations 
-@spiller_interaction-based_2020 treat a SNP-covariate interaction as an instrumental variable, and, by so doing, they impose assumptions on that interaction.
-Together, these assumptions require that MR-GxE assume that horizontal pleiotropy effects are constant across the study cohort.
-The authors devise a three-step inferential procedure for two-sample MR with GWAS summary statistics:
-
-1. Estimate the SNP-exposure and SNP-outcome associations at a range of covariate values  
-1. Regress the SNP-outcome estimated associations on the SNP-exposure estimated associations  
-1. Treat the resulting slope as the causal effect estimate and the intercept as the mean horizontal pleiotropy effect.
-
-@spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set 
-of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant 
-pleiotropy assumption.
-
-MR-GENIUS is a framework for two-sample MR analysis when individual-level data are available [@tchetgen_tchetgen_genius_2021]. @tchetgen_tchetgen_genius_2021 focused efforts on relaxing the 
-exclusion restriction assumption and building on existing G-estimation 
-methods to create an estimator that is robust to violations 
-of the exclusion restriction and to additive unmeasured confounding [@robins1994correcting].
-They observe that their estimator, in some settings, reduces to that of @lewbel2012using, which is widely used in econometrics, but not in the MR literature.
-
-Finally, before considering methods for MR with GWAS summary statistics, we consider 
-a new method for one-sample MR. @berzuini_bayesian_2020 proposed a Bayesian approach to MR where SNP instruments
-are permitted to exhibit pleiotropic effects on the outcome. Working in the setting of 
-one-sample MR with individual-level data, they construct their 
-model with a shrinkage prior, so that many SNP instruments have no horizontal pleiotropic effects. Their 
-method relaxes the exclusion restriction assumption to permit horizontal pleiotropy (for some SNP 
-instruments) and permits correlated SNP instruments, unlike many frequentist methods. Posterior inference 
-proceeds by Markov chain monte carlo sampling. They also elaborate their model for 
-a univariate exposure into that for a bivariate exposure to analyze the causal effects of body mass index and serum phenylalanine levels on blood pressure.
-
-### Methods with Summary data
-
-The proliferation of publicly available GWAS data coincides with the acceleration
-of methods for two-sample MR with GWAS summary statistics. Our timeline highlights some
-of the many two-sample MR methods with GWAS summary statistics, with an increasing 
-density of methods over time in the last three years (Figure \@ref(fig:timeline)).
-
-```{r, timeline, fig.cap = "Two-sample MR with GWAS summary statistics methods publications density increases over time."}
-knitr::include_graphics("figures/timeline.pdf")
-```
-
-@burgess_mendelian_2013 termed their method "inverse variance 
-weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
-of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
-causal effects.
-
-Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from $M$ independent SNPs by using Equation \@ref(eq:ivw).
-
-\begin{equation}
-  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
-  (\#eq:ivw)
-\end{equation}
-
-@bowden_mendelian_2015 adapted Egger regression, originally developed 
-to detect bias in meta-analyses, to detect bias from horizontal 
-pleiotropy in two-sample MR studies with GWAS summary statistics or individual-level data
-[@egger1997bias]. Their approach is termed MR-Egger. They formulate their model as
-
-\begin{eqnarray}
-X_i = \sum_{j=1}^J\gamma_jG_{ij} + U_i + \epsilon_{Xi}\nonumber \\
-Y_i = \sum_{j=1}^J\alpha_jG_{ij} + \beta X_i + U_i + \epsilon_{Yi}
-(\#eq:egger)
-\end{eqnarray}
-
-where $U$ designates confounders, $X$ and $Y$ denote the exposure and outcome, respectively, $G$ is a SNP 
-instrument, and $\epsilon_X$ and $\epsilon_Y$ are the zero-mean, normally distributed random errors.
-
-@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument $j$ and 
-the effects defined in Equation \@ref(eq:egger) and define the new parameter $\Gamma_j = \alpha_j + \beta\gamma_j$ (Equation \@ref(eq:egger1)).
-
-\begin{eqnarray}
-Y_i = (\alpha_j + \beta\gamma_j)G_{ij} +  \epsilon'_{Yij} \nonumber\\
-= \Gamma_jG_{ij} + \epsilon'_{Yij}
-(\#eq:egger1)
-\end{eqnarray}
-
-In Equation \@ref(eq:egger1), $\alpha_j = 0$ when SNP $j$ is a valid instrument.
-
-MR-Egger fits the regression model in Equation \@ref(eq:egger2).
-
-\begin{equation}
-\hat\Gamma_j = \beta_{0E} + \beta_E \hat\gamma_j
-(\#eq:egger2)
-\end{equation}
-
-The notation in Equation \@ref(eq:egger2) could be a little misleading. The independent
-and dependent variables are $\hat\gamma_j$ and $\hat\Gamma_j$, respectively, while 
-the regression parameters are $\beta_{0E}$ and $\beta_E$.
-
-The slope coefficient $\hat\beta_E$ estimates the causal effect of the exposure on the outcome. 
-Under the assumption that the SNP-exposure relationship is independent 
-of the SNP's pleiotropic effects, Egger's test offers a valid test of 
-the null causal hypothesis and consistently estimates the causal 
-effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].
-
-In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary 
-data, @greco_m_detecting_2015 developed the between-instrument heterogeneity $Q$ test. In a meta-analysis 
-of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).
-
-\begin{equation}
-Q = \sum_{k = 1}^K w_k(\hat\beta^{(k)}_{XY} - \mu_F)^2
-(\#eq:q)
-\end{equation}
-
-In Equation \@ref(eq:q), $w_k$ is the inverse variance of the Wald estimator, and $\mu_F = \frac{\sum_{k = 1}^K w_k \hat\beta^{(k)}_{XY}}{\sum_{k = 1}^K w_k}$.
-
-While the $Q$ test tends to be conservative in small sample sizes, its power increases with increasing 
-sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
-
-@bowden_consistent_2016, noting that consistency with IVW MR is not robust to 
-invalid instrument use, developed a method for consistent MR analysis with  
-weighted median-based estimators, building on research from @han2008detecting. 
-Its breakdown point is 50\%, 
-meaning that up to 50\% of SNP instruments can be invalid while maintaining consistency in estimation 
-[@hampel1971general]. Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate 
-statistic, from least to greatest. If an odd number, say $2K +1$, of SNP instruments is used, then 
-the simple median estimator is defined as the $K + 1$ ordered ratio estimate. For an even number, say 
-$2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio 
-estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.
-
-Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted 
-median estimator. To construct this estimator, order the ratio estimates from least to greatest, as 
-with the simple median estimator. The weighted median estimator, then is the median of a distribution
-defined by having $\tilde \beta_j$ as the $p_j = 100(\sum_{k = 1}^j - \frac{w_j}{2})$ percentile, 
-where $w_j$ is the weight assigned to the $j^{th}$ ordered ratio estimate. In this setting, the 
-simple median estimator is seen to be the weighted median estimator when all weights are equal.
-
-@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
-penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
-the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
-a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison]. The penalized 
-weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald 
-ratio estimates.
-
-# Horizontal pleiotropy
-
-Pleiotropy, where a single genetic variant affects multiple traits, has a long history of study in genetics and complex traits [@solovieff_pleiotropy_2013;@visscher_plethora_2016;@stearns_one_2010]. 
-Horizontal pleiotropy refers to the presence of multiple causal pathways from a genetic 
-variant to the outcome.
-The fact that there are pathways that exclude the exposure and connect the SNP to the outcome violates
-standard MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
-in GWAS. This finding motivated the development of MR-PRESSO and inspired 
-further research into methods to model horizontal pleiotropy [@verbanck_detection_2018].
-
-In the short time since publication of @verbanck_detection_2018, researchers have recognized two 
-types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
-exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
-SNP-exposure relationships. The second type of horizontal pleiotropy manifests
-in the presence of unobserved exposure-outcome confounding. It induces correlation between 
-horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
-violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
-discoveries [@yuan_likelihood_2021].
-Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
-instrumental SNPs that might be associated with the outcome 
-[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
-More recent methods have attempted to model horizontal pleiotropy 
-[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
-MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
-types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
-challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
-Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
-
-@hemani_orienting_2017, working in the setting of two-sample MR with summary GWAS data, recognized that
-measurement error in outcome and exposure can give misleading results with the inferred 
-causal arrow in the wrong direction. To work around this issue, they extended 
-MR with a method that infers the causal direction between two traits. Specifically, @hemani_orienting_2017 
-adapt Steiger's Z test for a difference in correlations [@steiger1980tests] to assess which of the two 
-traits in an analysis has the stronger correlation with the SNP instrument. That 
-with the stronger correlation is treated as the exposure. @hemani_orienting_2017 
-applied their test in a study of DNA methylation and gene expression, where either 
-direction of causality is plausible, and found 
-that many methylation traits cause changes in gene expression.
-
-@hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data. 
-They first calculate Wald ratio estimates for every instrumental SNP. They then apply 
-smoothing to the empirical distribution of ratio estimates. The mode of this smoothed 
-distribution is the simple mode-based estimate of the causal effect 
-[@hartwig_robust_2017]. The inverse variance-weighted mode-based estimate is obtained by 
-weighting the empirical distribution of ratio estimates by the inverse variance of each 
-estimate. @hartwig_robust_2017 compared their weighted and unweighted mode-based 
-estimators with weighted and unweighted median-based estimators, MR Egger regression, and 
-IVW. They found that their mode-based estimators demonstrated less bias and lower type I error rates than 
-other estimators in some simulation settings. The mode-based estimators possessed lower statistical power to 
-detect a causal effect compared to the IVW and weighted median-based methods, but their power exceeded that of 
-MR Egger regression [@hartwig_robust_2017]. The mode-based estimators consistently estimated the causal effect 
-when the mode across SNP instruments of the horizontal pleiotropy effects was zero [@hartwig_robust_2017]. In 
-this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the 
-other estimators.
-
-@rees_robust_2019 approached the challenge of horizontal pleiotropy by 
-developing three new methods for two-sample MR analysis with GWAS summary data: robust 
-regression, penalized weights, and LASSO penalization. The first two can be viewed as 
-modifications of MR-Egger and IVW methods. Together, they offer three strategies for 
-downweighting or excluding variants with heterogeneous causal estimates.
-Recall that MR-Egger offers consistent causal effect estimates when there is no 
-horizontal pleiotropy or when the horizontal pleiotropy effects adhere to the 
-"Instrument Strength Independent of Direct Effect" (InSIDE) assumption [@bowden_mendelian_2015]. The 
-InSIDE assumption is satisfied when there is no correlation between the pleiotropic SNP 
-instrument-outcome effects and the SNP instrument-exposure effects. @rees_robust_2019 uses 
-MM estimation, which is one technique for robust linear regression [@koller2011sharpening], to 
-formulate an estimator with high breakdown point and high efficiency. The second estimation 
-method, with penalized weights, is much like the penalized median estimator, from above, 
-except that the second argument in the minimum function is $100q_j$, instead of the $20q_j$ 
-above. Finally the LASSO-based method fashions an estimator by applying a penalty to the 
-objective function from MR-Egger to get Equation \@ref(eq:lasso) as the objective function.
-
-\begin{equation}
-\sum_j [Var(\hat\beta_j)]^{-1}(\hat\beta_{Yj} - \theta_{0j} - \theta_1 \hat\beta_{1j})^2 + \lambda \sum_j |\theta_{0j}|
-(\#eq:lasso)
-\end{equation}
-
-@rees_robust_2019 studied two procedures for choosing $\lambda$: a heterogeneity stopping rule 
-and a cross-validation rule. Taken together, these three methods offer additional contributions to 
-a suite of sensitivity analysis tools that practitioners should consider.
-
-@burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR
-analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or 
-more of the three core MR assumptions. Working in the setting of two-sample MR with GWAS data, they begin by 
-considering a collection of candidate SNP instruments, which may contain some invalid SNP instruments. For 
-each candidate SNP instrument, they calculate the Wald ratio estimate of the causal effect. With the 
-collection of Wald ratio estimates, @burgess_robust_2020 reason that the set of valid SNP instruments have 
-Wald ratio estimates that arise from a normal distribution centered on the true causal parameter with variance 
-equal to that of the Wald ratio estimator. The invalid instruments, however, are assumed to follow a normal 
-distribution centered at zero with a variance greater than that of the Wald ratio estimator. They then work 
-with profile log likelihood over a grid of causal parameter values to make inferences.
-
-@zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted 
-profile score to accomplish statistical inference in two-sample MR analyses with GWAS 
-summary data. Dividing pleiotropic effects into systematic and idiosyncratic, they model the 
-systematic pleiotropy with random effects. In so doing, no SNP satisfies the exclusion 
-restriction assumption. Adjusting their profile likelihood estimator from the setting 
-without pleiotropy, @zhao_statistical_2020 obtain an estimator with consistency and 
-asymptotic normality [@mccullagh_simple_1990]. Idiosyncratic pleiotropy is addressed through robustification of 
-the adjusted profile score [@huber_robust_1964]. They then demonstrate the properties of their estimator by 
-analyzing simulated and real data [@zhao_statistical_2020].
-
-@bowden_improving_2019, working with a collection of SNP instruments, reasoned that if all three MR assumptions are satisfied, then the 
-collection of Wald ratio estimates should be homogeneous; thus, any departures from 
-homogeneity may signal a violation of one or more assumptions. They reported 
-modified weights, in addition to studying weights derived from first order and 
-second order approximations to the variance. Their modified weights lead 
-to an estimator that is useful for quantifying heterogeneity and detecting outliers.
-
-@cho_exploiting_2020 developed the MR-TRYX framework for exploiting horizontal pleiotropy
-to identify alternative causal pathways. They did this by positing that, for a 
-collection of putative SNP instruments, any heterogeneity in the Wald ratio estimates
-may be due to horizontal pleiotropy. Once they identify a SNP with outlying 
-ratio estimates, they search GWAS results by querying the MR-Base database to identify complex traits (in other studies) that 
-associate with the outlying SNP [@hemani_mr-base_2018]. They follow up promising trait associations to 
-account for their observations with multiple causal pathways.
-
-@foley_mr-clust_2021 recognized that in some MR studies, the Wald ratio estimates for different
-SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms 
-and may identify the SNPs that are involved in them. They devised a method to detect 
-these clusters in the collection of Wald ratio estimates and found SNP instruments that 
-point to distinct causal pathways of biological importance [@foley_mr-clust_2021].
-
-@cheng_mr-ldp_2020 sought a probabilistic model that accounts for both linkage disequilibrium among SNP 
-instruments and horizontal pleiotropy. They termed their method MR-LDP. Using an approach from @zhu_bayesian_2017, @cheng_mr-ldp_2020 derived 
-an approximate likelihood for their regression models of the exposure and outcome on the SNP instruments. 
-Incorporation of a random effect into their models accounted for the variance in causal effect estimates due 
-to horizontal pleiotropy. They situate inference within an empirical Bayesian framework and present a 
-parameter-expanded variational Bayes expectation-maximization algorithm for estimation [@liu1998parameter]. 
-By jointly modeling the distribution of GWAS summary statistics and causal effects, @cheng_mr-ldp_2020 
-efficiently accommodates multiple SNP instruments in linkage disequilibrium. One limitation of MR-LDP  
-is its reliance on the InSIDE assumption and resulting inability to accommodate correlated horizontal pleiotropy.
-
-@zhu_iterative_2021 present IMRP, a method for causal effect estimation and horizontal pleiotropy detection. 
-They describe a test to distinguish vertical pleiotropy from horizontal pleiotropy. By combining this test 
-with two-sample MR analysis of GWAS summary statistics, they develop an iterative procedure for estimating 
-the causal effect of exposure on outcome and test for presence of horizontal pleiotropy in the SNP 
-instruments. The iterative procedure alternates between 1) updating the estimate of the causal effect and 2) 
-testing for horizontal pleiotropy with the included SNP instruments. At each test for horizontal pleiotropy, 
-some SNP instruments may be discarded if they demonstrate evidence of horizontal pleiotropic effects. The 
-procedure ends when there remain only SNP instruments that don't exhibit horizontal pleiotropy. In the authors' 
-simulations and data analysis sections, their new method performs about as well as GSMR, IVW, and MR-PRESSO in the 
-settings with and without balanced pleiotropy and with and without satisfaction of the InSIDE assumption [@zhu_iterative_2021].
-Finally, the authors note that IMRP is three orders of magnitude faster than simulations-based MR-PRESSO.
-
-@grant_efficient_2020 build on research from @lin2015regularization to present a method that 
-applies a $\ell_1$ penalty to coefficients of covariates in two-sample MR with GWAS summary 
-statistics. Importantly, their method doesn't penalize the SNP-exposure association. Instead, 
-it shrinks towards zero those coefficients for covariates that demonstrate little pleiotropy. In so doing, 
-they extend work from @burgess_multivariable_2015 and @burgess_re_2015 to get a causal effect estimator that needs no valid SNP instruments and robust to measured vertical pleiotropy and more statistically efficient
-than previous MR methods.
-
-@xu_mrcip_2021 consider two classes of horizontal pleiotropy with their method, MRCIP. They partition horizontal 
-pleiotropic effects into correlated or idiosyncratic horizontal pleiotropy. Idiosyncratic horizontal pleiotropy 
-is that with a large effect size, while correlated horizontal pleiotropy, like its definition elsewhere in our 
-manuscript, refers to correlation between the effects of the SNP instruments on the exposure and the effects of 
-the SNP instruments on the outcome. MRCIP relaxes both the exclusion restriction assumption and the InSIDE 
-assumption by directly modeling the correlated horizontal pleiotropy with random effects. Furthermore, to 
-accommodate the idiosyncratic pleiotropy, the authors downweight SNP instruments that demonstrate strong direct 
-effects on the outcome. The find that their MRCIP provides valid causal inference even when there are no valid 
-SNP instruments (*i.e.*, when all SNP instruments are invalid) and when the InSIDE assumption is violated 
-[@xu_mrcip_2021]. Their PRW-EM algorithm adds a reweighting step to the traditional expectation 
-maximization algorithm to estimate the causal effect parameter in MR analysis [@dempster1977maximum]. 
-Finally, they compare  MRCIP to other MR methods with both simulated and real data sets. @xu_mrcip_2021, 
-like the developers of MR-RAPS, assume that the horizontal pleiotropy is balanced, *i.e.*, that the mean 
-horizontal pleiotropic effect is zero. One direction for future study is to augment MRCIP to permit 
-unbalanced horizontal pleiotropy.
-
-@iong_latent_2020 build on other research using mixture models in the two-sample MR setting with GWAS summary 
-statistics. Like MR-Clust, MR-PATH, as the method from @iong_latent_2020 is called, works with the Wald ratio 
-estimates from multiple SNP instruments. Coining the term "mechanistic heterogeneity" to refer to the 
-distribution of Wald ratio estimates due to SNP involvement in distinct causal pathways, @iong_latent_2020 
-specify a $J$-component normal mixture model. They use an expectation maximization algorithm to fit the model 
-parameters (the component weights and the parameters for each of the $J$ normal distributions) and suggest a 
-Bayesian information criterion-based approach to choosing $J$ [@schwarz1978estimating].
-
-@wang_weakinstrument_2021 propose causal effect test statistics that are robust under weak instrument asymptotics 
-[@staiger1994instrumental]. Specifically, they extend three econometric methods to two-sample MR with GWAS 
-summary statistics [@anderson1949estimation;@kleibergen_pivotal_2002;@moreira_conditional_2003] and study the 
-theoretical properties and the performance in applications. @wang_weakinstrument_2021 extend their methods to 
-demonstrate conditions under which their estimators are equivalent to those in works from @zhao_statistical_2020, @windmeijer_use_2019, and @bowden_improving_2019.
-
-@bucur_inferring_2020 developed BayesMR, a Bayesian framework for two-sample MR with individual-level data and discuss an approximation for use with  
-GWAS summary statistics. Unlike many MR methods, BayesMR both aims to control for 
-horizontal pleiotropy and to quantitatively assess the possibility of reverse 
-causation, where the putative exposure is actually caused by the putative outcome. 
-For $J$ independent SNP instruments, @bucur_inferring_2020 permit each SNP instrument to 
-have both a direct effect on the exposure and a horizontally pleiotropic direct effect 
-on the outcome. Instrument effects and horizontally pleiotropic effects are assumed 
-independent. Note that this is the Bayesian analog of the InSIDE assumption. 
-After specifying the model, they present a nested sampling scheme [@skilling2006nested;@handley2015polychord1;@handley2015polychord] that simultaneously 
-computes the model evidence for both causal directions (*i.e.*, exposure causes 
-outcome and outcome causes exposure) and acquires samples from the posterior 
-distribution for Bayesian statistical inferences.
-
-@schmidt_mendelian_2018 present BMRE, a Bayesian implementation of 
-MR Egger. They apply weakly informative priors to the horizontal pleiotropy effect 
-estimator in efforts to increase statistical power to detect a nonzero 
-slope, i.e., the coefficient of the SNP instrument. They develop their method for two-sample MR with individual-level 
-data. Simulations demonstrate that BMRE outperforms the MR Egger method while maintaining type I error rates in the examined scenarios. The authors emphasize a potential role for BMRE in 
-MR sensitivity analysis.
-
-@sanderson_examination_2019 found that multivariable MR, where multiple exposures jointly 
-cause an outcome, consistently estimates the effect of the exposures on the outcome. They 
-examined multivariable MR with both individual-level data and GWAS summary statistics. 
-@sanderson_examination_2019 also develop a generalized version of Cochran's Q test for 
-quantifying SNP instrument strength and validity. However, this test requires knowledge of 
-the covariances between the effects of the SNP instruments on the exposures. While these 
-covariances could be estimated from individual-level data, such data are often unavailable in 
-the GWAS setting. Importantly, the authors note that the estimands for multivariable MR nad MR 
-differ; multivariable MR estimates the direct causal effects of each exposure on the outcome, 
-while MR estimates the total causal effect of an exposure on the outcome.
-
-MR-link, reported by @van_der_graaf_mendelian_2020, aims to account for unobserved pleiotropy and 
-linkage disequilibrium by using individual-level data on the outcome variable and summary statistics for the exposure. MR-link uses a three-step procedure to estimate causal effects.
-
-1. Use GCTA-COJO to obtain SNP instruments [@genetic_investigation_of_anthropometric_traits_giant_consortium_conditional_2012]  
-1. Prune all SNPs in linkage disequilibrium with the SNP instruments to obtain a set of "tag" SNPs 
-with correlations less than 0.95  
-1. Solve for the causal effect estimate with ridge regression of the outcome on the matrix resulting from concatenation of the SNP instrument genotypes matrix and the tag SNPs genotypes matrix.
-
-Note that in step 3, ridge regression is preferred to ordinary least squares due to the limited 
-statistical power of ordinary least squares in this setting. Presumably, this is due to collinearity in 
-the matrix of tag SNP genotypes.
-
-# Correlated horizontal pleiotropy
-
-The second type of horizontal pleiotropy, from above, is sometimes called 
-"correlated horizontal pleiotropy". Correlated horizontal pleiotropy occurs 
-when a SNP affects both the exposure and a variable that confounds the 
-relationship between exposure and outcome (Figure \@ref(fig:dag3)).
-
-```{r, dag3, fig.cap = "Correlated horizontal pleiotropy occurs when the genetic variant G affects exposure X, which in turn affects Y, and G affects U, which in turn affects X and Y independently."}
-d1 <- dagitty('dag {
-bb="0,0,1,1"
-G [pos="0.0,0.0"]
-U [pos="0.75,0.5"]
-X [exposure, pos="0.5,0.0"]
-Y [outcome, pos="1.0,0.0"]
-G -> X
-G -> U
-U -> X
-U -> Y
-X -> Y
-}'
-)
-plot(d1)
-
-```
-
-CAUSE and MRAID are two methods that aim to model correlated horizontal pleiotropy [@morrison_mendelian_2020;@yuan_likelihood_2021]. 
-Their approaches to modeling correlated horizontal pleiotropy differ in important ways that ultimately 
-affect performance. 
-CAUSE allows for a proportion of SNPs to exhibit correlated pleiotropy, which @morrison_mendelian_2020 
-models as an effect on a shared, unobserved factor. The remaining SNPs are independent of this 
-unobserved factor. Every SNP can have a nonzero pleiotropic effect on the outcome, and these 
-pleiotropic effects are uncorrelated with the variant effects on the exposure. The model, then, 
-is written as a mixture of biavariate normal distributions, and inference proceeds by borrowing ideas on 
-adaptive shrinkage from @stephens_false_2016. Finally, @morrison_mendelian_2020 compares two models 
-to determine whether the GWAS summary statistics are consistent with a causal effect of the        
-exposure on the outcome. Specifically, they estimate the difference in the expected log pointwise 
-posterior density for the model with causal effect fixed at zero and the model that permits a nonzero 
-causal effect [@vehtari_practical_2017].
-
-@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
-biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].
-These observations motivated @yuan_likelihood_2021 to develop a new method, MRAID. MRAID accommodates both individual-level data and GWAS summary statistics [@yuan_likelihood_2021]. We focus on 
-the approach that uses GWAS summary statistics. Two equations are central to MRAID modeling (Equations \@ref(eq:mraid1) and \@ref(eq:mraid2)).
-
-\begin{equation}
-  \hat{\beta_X} = \Sigma_1\beta + e_x
-  (\#eq:mraid1)
-\end{equation}
-
-\begin{equation}
-  \hat{\beta_Y} = \alpha \Sigma_2 \beta + \Sigma_2\eta_0 + \Sigma_2 \eta_1 + e_y
-  (\#eq:mraid2)
-\end{equation}
-
-The estimated marginal effects for the $p$ SNP instruments on the exposure are denoted by $\hat \beta_X$, while 
-$\hat\beta_Y$ represents the estimated effects of the same SNPs on the outcome. The $p$ by $p$ SNP instrument correlation matrices for the exposure and outcome 
-are written as $\Sigma_1$ and $\Sigma_2$, respectively. $\Sigma_1$ and $\Sigma_2$ can be estimated with 1000 Genomes 
-Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry 
-[@the_1000_genomes_project_consortium_1000_2012]. The $p$-vector error terms, $e_x$ and $e_y$, follow multivariate normal 
-distributions with mean zero and variances $\Sigma_1\frac{\sigma_X^2}{n_1 - 1}$ 
-and $\Sigma_2\frac{\sigma_Y^2}{n_2 - 1}$, 
-respectively.
-@yuan_likelihood_2021 then constructs a Gibbs sampler to make likelihood-based inferences [@geman1984stochastic].
-To facilitate computations, the investigators make several assumptions about the collection of SNPs:
-
-1. Relatively small proportion of SNPs have nonzero effects on the exposure  
-1. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
-1. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs  
-1. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.
-
-By encoding these assumptions in prior distributions, @yuan_likelihood_2021 enables the inferential procedures to accommodate the observed data in the context of the assumptions on SNPs.
-
-# Binary, Count, and Time-to-event outcomes and exposures
-
-While we've focused above on continuous outcomes and continuous exposures, some
-data are more naturally treated as binary, count, or time-to-event variables. MR methods 
-for non-continuous outcomes and exposures is an active area of research. @burgess_improving_2012 and @allman_novel_2021 
-developed MR strategies for binary exposures and binary outcomes. @burgess_improving_2012 extended 
-methods from @wald_fitting_1940 by fitting two logistic regressions, one for the outcome and one for 
-the exposure. They then calculated the causal effect estimator as the ratio of the two logistic 
-regression coefficient estimators. @allman_novel_2021 devised a new MR method for binary outcome and 
-binary exposure by drawing on @gauvreau_analysis_1997 and treating the exposure and outcome as correlated binary 
-random variables. They implemented an iterative optimization algorithm to 
-simultaneously infer the causal effect parameter and other parameters.
-
-Researchers have also leveraged other generalized linear models for analysis in MR 
-studies [@nelder1972generalized]. However, this area is relatively unexplored. @hazewinkel_mendelian_2022 treated hospitalizations as a count variable and 
-modeled it with quasi-Poisson methods. @allman_mendelian_2022 characterized statistical properties, including bias and 
-power, with simulations of count random variables and binary random variables. With growing interest
-in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, 
-there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].
-
-Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017]. 
-@guo_body_2017, for example, studied post-diagnosis survival time in women with breast cancer. They 
-treated survival as a time-to-event outcome, which enabled them to model it with a proportional 
-hazards model [@cox_regression_1972] in two-stage MR with individual-level data. Similarly, 
-@he_causal_2018, in a study of cardiovascular disease, applied proportional hazards regression when 
-examining time to stroke as an outcome variable. In very recent work, @tchetgen_tchetgen_genius_2021
-reported methods, MR GENIUS, that are robust to violations of the exclusion restriction assumption (under 
-many assumed MR data generating processes). They considered binary, continuous, and time-to-event 
-outcomes within their framework. We are unaware of MR studies that treat 
-the exposure variable as a time-to-event, but it's conceivable that future research will consider 
-time-to-event traits as exposures in MR analyses.
-
-# Recent Findings
-
-## Omnigenic MR
-
-Recent research on the omnigenic hypothesis, which posits 
-that every SNP's effect on a trait is nonzero, has informed
-MR methods development [@Boyle2017]. @wang_mendelian_nodate 
-recognized the limitations of previous MR methods in the 
-context of the omnigenic hypothesis. In response, @wang_mendelian_nodate 
-developed a MR method that uses all genome-wide SNPs as instruments. 
-Using GWAS summary statistics as inputs, their method relies on a composite likelihood framework for scalable computation and allows for horizontal pleiotropy. @wang_mendelian_nodate used extensive 
-simulations, including those with model misspecifications, to characterize their method's statistical power and robustness. Finally, they applied the new method to 
-identify multiple complex traits that affect coronary artery disease and asthma.
-These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.
-
-## Sample overlap in biobanks
-
-The issue of overlapping samples in the two-sample MR with GWAS summary data has recently 
-received attention in the published literature.
-
-@fang2021evaluating, working with data from the UK Biobank, found that SNP instruments 
-derived from overlapping samples explained a higher proportion of the variance compared to 
-those derived from non-overlapping samples. They argue that block jackknife resampling MR
-enables causal inference while avoiding bias due to overlapping samples.
-
-@sadreev2021navigating analyzed 2,514 traits from the UK Biobank and evaluated the impact of winner's curse on MR analysis. Winner's curse, they found, amplifies weak instrument bias without 
-inflating the false discovery rate. This finding led the authors to design a pseudoreplication process that reduces bias in MR studies.
-
-## Bi-directional causal inference with MR
-
-@wang_causal_2021 developed a framework, GRAPPLE, for performing two-sample MR with weak and strong 
-instruments. Using GWAS summary statistics as inputs, GRAPPLE extends MR-RAPS and can detect presence of horizontally 
-pleiotropic pathways, infer the causal direction, and perform multivariable MR. Central to the 
-GRAPPLE framework is the observation that a horizontally pleiotropic pathway often gives rise to an 
-additional mode in the profile likelihood. GRAPPLE uses the presence of multiple profile likelihood 
-modes to diagnose horizontal pleiotropy with effects that are grouped by pathway. 
-GRAPPLE facilitates identification of SNP instruments that contribute to the 
-additional profile likelihood modes. A researcher using GRAPPLE may then recognize 
-a confounding factor for each mode. With additional GWAS summary data with traits 
-that reflect the confounding factors, GRAPPLE can fit multivariable MR models when
-the InSIDE assumption holds for the remaining horizontally pleiotropic effects [@wang_causal_2021].
-
-# Applications of MR in GWAS
-
-MR analysis has revealed mechanistic insights for many 
-common diseases [@mancuso2017integrating;@parisinos2020genome;@zeng_causal_2019-2;@larsson2019smoking;@zheng2020phenome;@robinson2016insight;@luo2019association;@au2018impact;@zhao2019effect;@zeng_causal_2019;@zeng_causal_2019-1].
-In some cases, a randomized clinical trial is not possible for ethical or other reasons.
-Below, we discuss recent applications of two-sample MR to GWAS summary statistics.
-
-Birthweight is thought to impact long-term susceptibility to adulthood diseases. This scientific 
-hypothesis, sometimes called the "fetal origins of adult diseases" motivated work by @zeng_causal_2019 in 
-which the authors used MR to identify causal relationships between lower birthweight and greater risks of 
-coronary artery disease, myocardial infarction, and type 2 diabetes mellitus [@barker_fetal_1990;@lucas_fetal_1999;@barker_fetal_2002]. Specifically, 
-@zeng_causal_2019 used 
-inverse variance-weighted MR to examine a total of 21 adult diseases and 38 other complex traits. Three causal associations met the Bonferroni multiple hypothesis testing threshold 
-[@bonferroni1936teoria]. The authors analyzed data from 37 GWAS with up to 340,000 subjects with European 
-ancestry [@zeng_causal_2019]. They estimated odds ratios of 1.34, 1.30, and 1.41 for coronary artery 
-disease, myocardial infarction, and type 2 diabetes mellitus, respectively. Sensitivity analyses revealed 
-the robustness of these findings to possible MR assumptions violations, including pleiotropy, maternal 
-effects, and reverse causation [@zeng_causal_2019]. Curiously, none of the 38 other complex traits 
-appeared to mediate the causal associations, perhaps due to limited sample sizes for these traits [@zeng_causal_2019]. @zeng_causal_2019 argue that further 
-studies, including consideration of socioeconomic status and incorporation of larger sample sizes, may 
-reveal mediating factors for these causal relationships and, thus, provide biological and public health insights.
-
-Amyotrophic lateral sclerosis is an incurable and largely untreatable neurological condition [@rowland_amyotrophic_2001;@kiernan_amyotrophic_2011].
-Working with GWAS summary statistics, @zeng_causal_2019-1 found evidence of causal relationships between
-type 2 diabetes mellitus and amyotrophic lateral sclerosis in both European and East Asian populations. 
-Interestingly, these effects differ in direction; in the European population, type 2 diabetes mellitus 
-exerted a neuroprotective effect and diminished the risk of amyotrophic lateral sclerosis, while in the 
-East Asian population type 2 diabetes mellitus increased the risk of amyotrophic lateral sclerosis. 
-@zeng_causal_2019-1 used inverse variance-weighted methods to estimate the causal effect from summary 
-statistics derived from approximately 660,000 type 2 diabetes mellitus European subjects and 81,000 
-amyotrophic lateral sclerosis European subjects. Smaller sample sizes (approximately 191,000 type 2 
-diabetes mellitus subjects and 4100 amyotrophic lateral sclerosis subjects) were available in the East 
-Asian cohorts. The authors used multiple SNP instruments with strong associations with type 2 diabetes 
-mellitus [@zeng_causal_2019-1]. The odds ratios of 0.93 and 1.28 in the Europeans and East Asians, 
-respectively, were robust against instrument selection, various model misspecifications, and estimation 
-biases [@zeng_causal_2019-1]. Egger regression and MR-PRESSO analyses were consistent with the absence of horizonal pleiotropy 
-[@zeng_causal_2019-1;@hemani_evaluating_2018;@verbanck_detection_2018;@ong2019implementing]. Future studies may reveal deeper biological insights into these observations and, ultimately, 
-contribute to treatments for amyotrophic lateral sclerosis.
-
-Birth weight is associated with childhood asthma, but whether there is a causal relationship 
-between the two remains unclear [@brooks_impact_2001]. To investigate this possibility, 
-@zeng_birth_2019 performed a MR analysis to assess the causal effect of birth weight on adult 
-asthma. They used data from two GWAS, with 140,000 subjects in the birth weight study and 62,000 
-subjects in the adult asthma study. With both the genetic risk score method and the 
-inverse variance-weighted method, @zeng_birth_2019 found no evidence of a causal effect of birth
-weight on adult asthma. Sensitivity analysis results offered reassurance that the 
-finding of no causal effect is robust. MR-PRESSO and MR-Egger found no evidence for bias from 
-SNP instrument outliers or horizontal pleiotropy. @zeng_birth_2019 concludes that confounding
-in other studies may explain the discrepancy between their findings and others' conclusions.
-
-# Summary and Outlook
-
-Modern GWAS for biobank-scale data offer the possibility
-of discovering even more causal relationships among complex traits. The large
-sample sizes in biobank studies, often greater than 100,000 
-subjects, offer the possibility of uncovering associations of modest strength 
-that might be undetectable in smaller studies. For example, the UK 
-Biobank Study involves more than 500,000 subjects. Each UK Biobank subject's 
-data consists of millions of SNP genotypes and measurements for 
-thousands of traits [@bycroft_uk_2018].
+In summary, we have presented a comprehensive review on 47 methods for MR analysis in GWAS (Figure 4).
+The overall trend of the methodology development for MR analysis is in the direction of increasingly sophisticated modeling of horizontal pleiotropy including both independent and correlated horizontal pleiotropy while attempting to maintain scalability and computational efficiency in the presence of multiple correlated SNPs.
+We hope our detailed review would benefit both methodology developers and applied analysts to further advance the development of MR methods and aid in their applications towards large, biobank-scale datasets, thus offering the possibility of discovering even more causal relationships among complex traits [@bycroft_uk_2018].
 

### 2022-04-27 reply-to-review1/old-eqns.Rmd
updated bib file to match Rmd citations

--- 2022-03-04 mr-review.Rmd
@@
 ---
-title: "Mendelian randomization with genome-wide association studies data"
-author: "Frederick J. Boehm & Xiang Zhou"
-date: "`r lubridate::now()`"
-bibliography: exported.bib
-csl: elsevier-with-titles.csl
-link-citations: yes
-linkcolor: blue
-always_allow_html: true
-output: 
-  - bookdown::pdf_document2
-  - bookdown::html_document2
-  - bookdown::word_document2
+title: "old equations"
+author: "Frederick Boehm"
+date: "4/27/2022"
+output: html_document
 ---
 
-```{r setup, echo = FALSE}
-knitr::opts_chunk$set(echo = FALSE, message = FALSE)
-library(magrittr)
-```
-
-# Abstract
-
-Complex trait genetics studies in humans have yielded thousands of new associations for  
-diverse traits. The next step in understanding biology of health and disease is to
-identify causal relationships among complex traits. Two-sample Mendelian randomization studies 
-in genome-wide association studies can uncover causal relationships 
-between complex traits. By leveraging shared SNP genotypes as instrumental variables, or proxies, for 
-the exposure complex trait, investigators can tease out causal effects from observational data, provided
-that necessary assumptions are satisfied. We discuss below the growth of genome-wide association studies 
-in parallel with development of Mendelian randomization methods.
-We argue that the recent availability of GWAS summary statistics for diverse complex traits has motivated new Mendelian randomization methods and that this area continues to offer opportunities for additional methods developments.
-
-# Introduction
-
-Great interest has developed in inferring causal relationships between complex 
-traits, *i.e.*, traits that seemingly are not inherited in a Mendelian fashion, in observational human genetics studies. Discovery of such relationships is crucial to 
-enhancing our understanding of the biology of health and disease. Mendelian randomization (MR) studies are a form of instrumental variable analysis in which
-genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for 
-inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].
-
-Genome-wide association studies (GWAS) present unique opportunities for 
-discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS
-interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
-In the nearly 17 years since publication of early studies, researchers have reported 
-thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012].
-Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of
-human health and disease.
-
-```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
-library(magrittr)
-dat <- readxl::read_xlsx(path = "data/publication-counts.xlsx", sheet = "counts")
-dat %>%
-  tidyr::pivot_longer(!year) %>%
-  dplyr::filter(year < 2022) %>%
-  ggplot2::ggplot(ggplot2::aes(x = year, y = value, fill = name)) + 
-  ggplot2::geom_bar(stat = "identity", 
-                    position=ggplot2::position_dodge()) + 
-  ggplot2::scale_fill_manual(labels = c("GWAS", "MR"), values=c('#999999','#E69F00')) + 
-  ggplot2::ylab("Annual publication count") + 
-  ggplot2::xlab("Year") + 
-  #ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
-  ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
-  ggplot2::theme(text = ggplot2::element_text(size = 20))
-
-ggplot2::ggsave("figures/mr-gwas-counts.png")
-
-```
-
-A number of excellent MR reviews have appeared in the literature [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022]. Our review doesn't 
-aim to replace them, but seeks to complement their findings. 
-We present recent methods developments for an audience of statisticians and epidemiologists. We discuss in detail methods advances before briefly summarizing 
-applications and directions for future research. The value of MR analysis with 
-GWAS data will continue to grow in parallel with statistical methods advances.
-
-Methods for and uses of MR have appeared at a rapid and accelerating pace since the 
-publication of the early GWAS (Figure \@ref(fig:article-counts)). Many studies use the "two-sample" MR 
-design, in which one cohort of subjects has measurements for the exposure, while a second cohort has 
-measurements for a second trait [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in 
-both sets of subjects.
-
-Recent methods developments have enabled execution of two-sample MR studies with only 
-GWAS summary statistics [@burgess2015using]. These strategies are especially pertinent now that there is 
-an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
-ability to use published data has contributed to a rapid increase in the number of
-two-sample MR methods for GWAS summary statistics (Figure \@ref(fig:article-counts)).
-
-# Mendelian Randomization with one Instrument
-
-We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where 
-there is a single genetic marker and a single outcome variable. Mendelian randomization with one 
-instrument uses a single genetic marker 
-as the proxy ("instrument" or "instrumental variable") for the "exposure", a complex trait, and asks 
-whether variation in the genotypes at the genetic marker causes variation in the 
-"outcome" variable, which is typically a second complex trait. When the genetic marker 
-is a SNP on an autosome, the cohort is partitioned into three 
-classes, two homozygote classes and a single heterozygote class, based on genotypes 
-for that marker. When the three assumptions of mendelian randomization are satisfied, 
-estimation of the causal effect of the exposure on the outcome is possible.
-
-The outcome variable can be binary, a count, a time to event, or a continuous variable. For brevity and simplicity, 
-we focus on continuous outcomes before considering other classes of outcome variables in a later section.
-
-## Assumptions of Mendelian Randomization
-
-Three assumptions are central to MR. Satisfying them enables inference of 
-causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
-Because assumptions 2 and 3, which involve absence of certain types of confounding, can't be verified, 
-practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
-way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014].
-
-The first assumption states that the exposure is associated with the outcome. The second requires 
-independence, conditional on the exposure and all measured and unmeasured confounders, between the 
-genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
-independent of all (measured and unmeasured) confounders of the relationship between exposure and 
-outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].
-
-If the first assumption, about the association between the genetic marker and the exposure, holds, 
-but this association is weak, then an amplification of biases, such as those due to violations of 
-assumptions 2 and 3, may result [@vanderweele_methodological_2014]. @burgess_avoiding_2011, in analyses 
-of simulated and real data, demonstrated that bias in causal estimates increases with decreases in 
-the strength of association between the SNP and exposure. We later discuss, in the context of 
-methods for multiple SNP instruments, how this finding, coupled with the recognition of 
-the polygenicity of many complex traits, has motivated methods to work with large numbers of weak SNP instruments [@ye_debiased_2021;@chao_consistent_2005].
-
-The second MR assumption, often termed the exclusion restriction assumption, can also be stated as 
-the need for the genetic marker to affect the outcome only through the exposure. In a 
-causal diagram, this means that the only path from the marker to the outcome is that 
-containing the exposure.
-
-The third assumption ensures that the genetic marker is
-independent of all confounders of the exposure-outcome relationship. Like the exclusion restriction
-assumption, we can't verify this assumption with observational data, so analysts need to 
-use tools like sensitivity analysis to ensure validity of findings.
-
-When the three MR assumptions hold, one may use instrumental variable statistical methods, 
-with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
-causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
-
-To understand the origin of the term "Mendelian randomization", it is helpful to 
-recall Mendel's inheritance laws [@mendel1865versuche]. Under the assumption that alleles segregate 
-randomly from parent to offspring, the offspring genotypes are unlikely to be 
-associated with confounders of the exposure-outcome relationship. Additionally,
-reverse causation is unanticipated since germ-line genotypes are 
-fixed at conception and, thus, precede realization of other observed variables. Reverse causation 
-occurs when the outcome at an early time point has a causal effect on the exposure variable. While 
-reverse causation is often assumed to not occur in MR study designs, @burgess_are_2021 points out 
-several scenarios where this assumption may not hold [@davey_smith_mendelian_2003].
-
-## Statistical models and methods for MR with one instrument
-
-Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, 
-likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods 
-that
-
-### Ratio of coefficients method
-
-The ratio of coefficients method, also known as the Wald method, estimates the 
-causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940].
 For a continuous outcome, the causal effect estimator, $\hat\gamma_{Wald}$ is defined in Equation \@ref(eq:wald).
@@
 SNP instrument.
-Since the Wald method requires only the regression coefficients, it can be used with summary data.
-However, the Wald method doesn't accommodate multiple SNP instruments, which limits its direct use in the GWAS setting.
 
@@
 
-A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969]. For a continuous 
-outcome, one may perform "two-stage least squares", which involves two linear regressions. 
 First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)).
@@
 thousands of traits [@bycroft_uk_2018].
-
-# References

### 2022-04-27 reply-to-review1/mr-review.Rmd
added math notation in latex

@@
 
-```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. mendelian randomization and 2. genome-wide association study."}
+```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. Mendelian randomization and 2. genome-wide association study."}
 library(magrittr)
@@
   ggplot2::xlab("Year") + 
-  #ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
+  ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
   ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
@@
 The ratio of coefficients method, also known as the Wald method, estimates the causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940].
-For a continuous outcome, the causal effect estimator,  is defined in Equation (3.1).
+For a continuous outcome, the causal effect estimator, \@ref(eq:wald).
 
-In Equation (3.1),  and  are the slope estimates from the regressions of the outcome and exposure, respectively, on the SNP instrument.
+\begin{equation}
+  \hat\gamma_{Wald} = \frac{\hat\beta_{YZ}}{\hat\beta_{XZ}}
+  (\#eq:wald)
+\end{equation}
+
+In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope estimates from the regressions of the outcome and exposure, respectively, on the SNP instrument.
 Since the Wald method requires only the regression coefficients, it can be used with summary data.
@@
 For a continuous outcome, one may perform “two-stage least squares,” which involves two linear regressions.
-First, the exposure variable is regressed on the instrument (i.e., the SNP genotypes) (Equation (3.2)).
-We denote each subject’s exposure variable value as  in Equation (3.2), while  denotes an intercept term and  is the slope.
-denotes the SNP instrument genotype for subject , while , the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.
-The resulting fitted values for the exposure variable from Equation (3.2) are the independent variable in the second linear regression, where the outcome is the dependent variable.
-The causal effect estimate, then, is the coefficient  obtained from the second regression analysis (Equation (3.3)).
-Note also that the random errors in Equation (3.3) are assumed independent of those in Equation (3.2).
-Like the  in Equation (3.2), the  are assumed independent and identically distributed normal random errors.
+First, the exposure variable is regressed on the instrument (i.e., the SNP genotypes) (Equation \@ref(eq:lm1)).
+We denote each subject’s exposure variable value as $x_i$ in Equation \@ref(eq:lm1), while $\alpha_0$ denotes an intercept term and $\alpha_1$ is the slope.
+$z_i$ denotes the SNP instrument genotype for subject $i$, while $\epsilon_i$, the error term, is assumed to be independent among subjects and normally distributed with a shared common variance and mean zero.
+
+The resulting fitted values for the exposure variable from Equation \@ref(eq:lm1) are the independent variable in the second linear regression, where the outcome is the dependent variable.
+The causal effect estimate, then, is the coefficient  $\hat \beta_1$ obtained from the second regression analysis (Equation \@ref(eq:lm2)).
+Note also that the random errors in Equation \@ref(eq:lm2) are assumed independent of those in Equation \@ref(eq:lm1).
+Like the $\epsilon_i$ in Equation \@ref(eq:lm1), the $\tau_i$ are assumed independent and identically distributed normal random errors.
+
+\begin{equation}
+  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
+  (\#eq:lm1)
+\end{equation}
+
+\begin{equation}
+  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
+  (\#eq:lm2)
+\end{equation}
 
@@
 Limited information maximum likelihood from econometrics is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite].
-Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations (3.4) and (3.5)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].
+Limited information maximum likelihood with a single SNP instrument is modeled with two equations (Equations \@ref(eq:liml1) and \@ref(eq:liml2)), where the random errors follow a bivariate normal distribution [@burgess_review_2017].
+
+\begin{equation}
+  x_i = \alpha_0 + \alpha_1 z_{i1} + \epsilon_{Xi}
+  (\#eq:liml1)
+\end{equation}
+
+\begin{equation}
+  y_i = \beta_0 + \beta_1 x_i + \epsilon_{Yi}
+  (\#eq:liml2)
+\end{equation}
 
 Limited information maximum likelihood is sometimes called the maximum likelihood counterpart of two-stage least squares, and it yields the same causal estimate as two-stage least squares and the ratio method when used with a single SNP instrument.
-Additionally, the limited information maximum likelihood framework can accommodate  SNP instruments by replacing  with the sum .
+Additionally, the limited information maximum likelihood framework can accommodate  more than one SNP instrument by replacing $\alpha_1z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.
 One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003].
-@burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation (3.6)).
-The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter,  represents the effect between the true means for the exposure and the outcome.
+@burgess_improving_2012 examined a model that is similar to that from the limited information maximum likelihood framework (Equation \@ref(eq:b1)).
+The Bayesian model differs from the limited information maximum likelihood model in that the causal effect parameter, $\beta_1$ represents the effect between the true means for the exposure and the outcome.
 In the limited information maximum likelihood model, the causal effect is that of the measured effect on the measured outcome.
 
-For each subject , the exposure and outcome values come from a bivariate normal distribution.
+\begin{eqnarray}
+\begin{pmatrix}
+X_i \\
+Y_i
+\end{pmatrix} \sim N_2\left(\begin{pmatrix} \xi_i\\
+\eta_i
+\end{pmatrix}, \Sigma
+\right)\nonumber \\
+\xi_i = \alpha_0 + \alpha_1 z_{i1}\nonumber \\
+\eta_i = \beta_0 + \beta_1 \xi_i
+  (\#eq:b1)
+\end{eqnarray}
+
+For each subject $i$, the exposure and outcome values come from a bivariate normal distribution.
 The mean of the exposure distribution is assumed linear in the SNP instrument, and the mean of the outcome distribution is linear in the mean exposure [@burgess_review_2017;@jones_choice_2012].
 @burgess_improving_2012 demonstrated that the Bayesian model, with weak instruments, outperforms the limited information maximum likelihood model in terms of frequentist coverage levels.
+
 ### Semiparametric methods
@@
 We follow [@burgess_review_2017] by discussing three semiparametric strategies, generalized method of moments, continuous updating estimator, and G-estimation of structural mean models.
+
 Generalized method of moments can be viewed as a more flexible form of two-stage least squares that handles heteroscedastic errors and nonlinearity in the two regressions [@burgess_review_2017].
 Before specifying the model, we introduce notation.
-is the conditional expectation of  if we forced  to take value  for every subject [@pearl2009causality].
-The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation (3.7).
+$\mathbb{E}(Y|\text{do}(X = x))$ is the conditional expectation of $Y$ if
+we forced $X$ to take value $x$ for every subject [@pearl2009causality].
+The generalized method of moments equations, with a single SNP instrument, then, can be written as in Equation \@ref(eq:gmm).
 
-The GMM estimate is the value of the vector  that satisfies Equation (3.7), where . @palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations (3.7).
+\begin{eqnarray}
+  \sum_i (y_i - f(x_i;\beta)) = 0\nonumber \\
+  \sum_i z_i(y_i - f(x_i;\beta)) = 0
+  (\#eq:gmm)
+\end{eqnarray}
+
+The GMM estimate is the value of the vector $\beta$ that satisfies Equation \@ref(eq:gmm), where $f(x_i; \beta) = \mathbb{E}(Y|\text{do}(X = x))$.
+@palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations \@ref(eq:gmm).
+
 Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].
@@
 
-The causal effect parameter is .
-[@burgess_review_2017] derives the estimating equations for , after noting that the conditional expectation,  is independent of  and reasoning that the causal effect is that value of  that yields zero covariance between  and  (Equations (3.9)).
+$Y(x)$ as the outcome value that we would 
+have observed had we set the exposure value $X$ to $x$. For example, $Y(0)$ denotes the observed outcome had 
+we set the exposure to zero instead of its observed value $x$. The structural mean model for a continuous 
+outcome is displayed in Equation \@ref(eq:smm).
 
-Note that  indexes SNP instruments and ranges from  to  in Equation (3.9).
+\begin{equation}
+\mathbb{E}(Y(x) - Y(0) | X = x, Z = z) = \beta_1 x
+  (\#eq:smm)
+\end{equation}
+
+The causal effect parameter is $\beta_1$. @burgess_review_2017 derives the estimating equations for 
+$\beta_1$, after noting that the conditional expectation, $\mathbb{E}(Y(0)| X = x, Z = z)$ is 
+independent of $Z$ and reasoning that the causal effect is that value of $\beta_1$ that yields zero 
+covariance between $Z$ and $\mathbb{E}(Y(0)| X = x, Z = z)$ (Equations \@ref(eq:smm1)).
+
+\begin{equation}
+\sum_i (z_{ik} - \bar z_{\cdot k})(y_i - \beta_1 x_i) = 0
+  (\#eq:smm1)
+\end{equation}
+
+Note that $k$ indexes SNP instruments and ranges from $1$ to $K$ in Equation \@ref(eq:smm1).
 
@@
 They also calculate a weighted allele score by using the exposure effects as weights.
-For example, a subject with  copies of exposure-increasing alleles for SNP  has an unweighted allele score  and a weighted allele score .
+For example, a subject with $g_k$ copies of exposure-increasing alleles for SNP $k$ has an unweighted allele score $z = \sum_{k = 1}^K g_k$ and a weighted allele score $z = \sum_{k = 1}^K w_kg_k$.
 @burgess_use_2013 found that when weights are obtained from external data or from cross-validation or jackknife approaches applied to the analysis data, the allele score functions as a single instrumental variable and greatly diminishes the bias compared to that of the two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
@@
 pleiotropic traits.
-For a collection of  SNPs, they choose the -vector that meets these criteria and maximizes the correlation with the exposure.
-To obtain their second estimator, @jiang_constrained_2019 apply a  penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation of the new instrumental variable.
+For a collection of  SNPs, they choose the $K$-vector that meets these criteria and maximizes the correlation with the exposure.
+To obtain their second estimator, @jiang_constrained_2019 apply a $l_0$ penalty to constrain the number of SNP instruments that are given nonzero weights in the calculation of the new instrumental variable.
 They then compare their method with sisVIVE and allele score methods.
@@
 
-Figure 2: Two-sample MR with GWAS summary statistics methods publications density increases over time.
+```{r, timeline, fig.cap = "Two-sample MR with GWAS summary statistics methods publications density increases over time."}
+knitr::include_graphics("../figures/timeline.pdf")
+```
+
 @burgess_mendelian_2013 termed their method “inverse variance weighting” (IVW) because each causal SNP’s effect is weighted by the inverse of the variance of the ratio estimator, and the overall causal effect is the sum of the weighted SNP causal effects.
-Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from  independent SNPs by using Equation (5.1).
+Specifically, @burgess_mendelian_2013 combined causal effect ratio estimates from  independent SNPs by using Equation \@ref(eq:ivw).
+
+\begin{equation}
+  \hat \beta_{IVW} = \frac{\sum_{j = 1}^M \hat\theta_j [Var(\hat\theta_j)]^{-1}}{\sum_{j = 1}^M [Var(\hat\theta_j)]^{-1}}
+  (\#eq:ivw)
+\end{equation}
 
@@
 Its breakdown point is 50%, meaning that up to 50% of SNP instruments can be invalid while maintaining consistency in estimation [@hampel1971general].
-Specifically, let  denote the  ordered ratio estimate statistic, from least to greatest.
-If an odd number, say , of SNP instruments is used, then the simple median estimator is defined as the  ordered ratio estimate.
-For an even number, say , number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, .
+Specifically, let $\tilde \beta_j$ denote the $j^{th}$ ordered ratio estimate statistic, from least to greatest.
+If an odd number, say $2K = 1$, of SNP instruments is used, then the simple median estimator is defined as the $K + 1$ ordered ratio estimate.
+For an even number, say $2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.
 Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted median estimator.
 To construct this estimator, order the ratio estimates from least to greatest, as with the simple median estimator.
-The weighted median estimator, then is the median of a distribution defined by having  as the  percentile, where  is the weight assigned to the  ordered ratio estimate.
+The weighted median estimator, then is the median of a distribution defined by having $\tilde \beta_j$ as the $p_j = 100$ percentile, where $w_j$ is the weight assigned to the $j^{th}$ ordered ratio estimate.
 In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
-@bowden_consistent_2016 also studied the penalized weighted median estimator.
-They defined the penalized weights, , where  is the p-value resulting from the comparison of  to a  distribution [@bowden_consistent_2016;@cochran1950comparison].
+@bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
+penalized weights, $w_j^* = w_j' \times \min(1, 20q_j)$, where $q_j$ is the p-value resulting from 
+the comparison of $Q_j = w_j'(\tilde\beta_j - \hat\beta_{IVW})^2$ to 
+a $\chi^2_1$ distribution [@bowden_consistent_2016;@cochran1950comparison].
 The penalized weighted median estimator leaves most variants unaffected, but downweights those with outlying Wald ratio estimates.
@@
 ## Assumption violations and Sensitivity Analysis
+
 It is worthwhile to consider ways in which the standard MR assumptions may be violated and strategies for identifying such violations.
@@
 It states that the SNP instrument must affect the outcome only through the exposure.
-The causal diagram in Figure  illustrates a scenario that violates this assumption [@vanderweele_methodological_2014].
+The causal diagram in Figure 3B illustrates a scenario that violates this assumption [@vanderweele_methodological_2014].
 Bias in causal effect estimation would result.
-However, if the intermediate is measured and treated as the exposure, then an unbiased estimate of causal effect is possible, since the intermediate captures the two pathways, intermediate -> outcome and intermediate -> exposure -> outcome, through which the SNP instrument affects the outcome Y [@vanderweele_methodological_2014;@textor_graphical_2011;@textor_robust_2016].
+However, if the intermediate is measured and treated as the exposure, then an unbiased estimate of causal effect is possible, since the intermediate captures the two pathways, intermediate -> outcome and intermediate -> exposure -> outcome, through which the SNP instrument affects the outcome Y [@vanderweele_methodological_2014].
 
-Figure 3: Causal diagrams for MR.
+```{r, dags, fig.cap = "Causal diagrams for MR.
 A.
+Scenario where the genetic variant affects an exposure, which in turn affects an outcome.
+B.
 Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure.
 Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption.
-B.
-Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently.
+C.
+Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently."}
+knitr::include_graphics("../figures/dags.pdf")
+```
 
@@
 
-where  designates confounders,  and  denote the exposure and outcome, respectively,  is a SNP instrument, and  and  are the zero-mean, normally distributed random errors.
-@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument  and the effects defined in Equation (5.2) and define the new parameter  (Equation (5.3)).
+\begin{eqnarray}
+X_i = \sum_{j=1}^J\gamma_jG_{ij} + U_i + \epsilon_{Xi}\nonumber \\
+Y_i = \sum_{j=1}^J\alpha_jG_{ij} + \beta X_i + U_i + \epsilon_{Yi}
+(\#eq:egger)
+\end{eqnarray}
 
-In Equation (5.3),  when SNP  is a valid instrument.
-MR-Egger fits the regression model in Equation (5.4).
+where $U$ designates confounders, $X$ and $Y$ denote the exposure and outcome, respectively, $G$ is a SNP 
+instrument, and $\epsilon_X$ and $\epsilon_Y$ are the zero-mean, normally distributed random errors.
 
-The notation in Equation (5.4) could be a little misleading. The independent and dependent variables are  and , respectively, while the regression parameters are  and .
-The slope coefficient  estimates the causal effect of the exposure on the outcome.
-Under the assumption that the SNP-exposure relationship is independent of the SNP’s pleiotropic effects, Egger’s test offers a valid test of the null causal hypothesis and consistently estimates the causal effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].
+@bowden_mendelian_2015 then write the outcome in terms of the SNP instrument $j$ and 
+the effects defined in Equation \@ref(eq:egger) and define the new parameter $\Gamma_j = \alpha_j + \beta\gamma_j$ (Equation \@ref(eq:egger1)).
+
+\begin{eqnarray}
+Y_i = (\alpha_j + \beta\gamma_j)G_{ij} +  \epsilon'_{Yij} \nonumber\\
+= \Gamma_jG_{ij} + \epsilon'_{Yij}
+(\#eq:egger1)
+\end{eqnarray}
+
+In Equation \@ref(eq:egger1), $\alpha_j = 0$ when SNP $j$ is a valid instrument.
+
+MR-Egger fits the regression model in Equation \@ref(eq:egger2).
+
+\begin{equation}
+\hat\Gamma_j = \beta_{0E} + \beta_E \hat\gamma_j
+(\#eq:egger2)
+\end{equation}
+
+The notation in Equation \@ref(eq:egger2) could be a little misleading. The independent
+and dependent variables are $\hat\gamma_j$ and $\hat\Gamma_j$, respectively, while 
+the regression parameters are $\beta_{0E}$ and $\beta_E$.
+
+The slope coefficient $\hat\beta_E$ estimates the causal effect of the exposure on the outcome. 
+Under the assumption that the SNP-exposure relationship is independent 
+of the SNP's pleiotropic effects, Egger's test offers a valid test of 
+the null causal hypothesis and consistently estimates the causal 
+effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015]. 
 In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity  test.
-In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate  in Equation (5.5).
+In a meta-analysis
+of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).
 
-In Equation (5.5),  is the inverse variance of the Wald estimator, and .
-While the  test tends to be conservative in small sample sizes, its power increases with increasing sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
+\begin{equation}
+Q = \sum_{k = 1}^K w_k(\hat\beta^{(k)}_{XY} - \mu_F)^2
+(\#eq:q)
+\end{equation}
+
+In Equation \@ref(eq:q), $w_k$ is the inverse variance of the Wald estimator, and $\mu_F = \frac{\sum_{k = 1}^K w_k \hat\beta^{(k)}_{XY}}{\sum_{k = 1}^K w_k}$.
+
+While the $Q$ test tends to be conservative in small sample sizes, its power increases with increasing 
+sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].
 
@@
 The second estimation method, with penalized weights, is much like the penalized median estimator, from above, except that the second argument in the minimum function is , instead of the  above.
-Finally the LASSO-based method fashions an estimator by applying a penalty to the objective function from MR-Egger to get Equation (6.1) as the objective function.
+Finally the LASSO-based method fashions an estimator by applying a penalty to the objective function from MR-Egger to get Equation \@ref(eq:lasso) as the objective function.
 
-@rees_robust_2019 studied two procedures for choosing : a heterogeneity stopping rule and a cross-validation rule.
-Taken together, these three methods offer additional contributions to a suite of sensitivity analysis tools that practitioners should consider.
+\begin{equation}
+\sum_j [Var(\hat\beta_j)]^{-1}(\hat\beta_{Yj} - \theta_{0j} - \theta_1 \hat\beta_{1j})^2 + \lambda \sum_j |\theta_{0j}|
+(\#eq:lasso)
+\end{equation}
+
+@rees_robust_2019 studied two procedures for choosing $\lambda$: a heterogeneity stopping rule 
+and a cross-validation rule. Taken together, these three methods offer additional contributions to 
+a suite of sensitivity analysis tools that practitioners should consider.
+
 @burgess_robust_2020 introduced the contamination mixture method for robust and efficient MR analysis with hundreds of SNPs, some of which may be invalid instruments by violating one or more of the three core MR assumptions.
@@
 Finally, the authors note that IMRP is three orders of magnitude faster than simulations-based MR-PRESSO.
-@grant_efficient_2020 build on research from @lin2015regularization to present a method that applies a  penalty to coefficients of covariates in two-sample MR with GWAS summary statistics.
+@grant_efficient_2020 build on research from @lin2015regularization to present a method that applies a $\ell_1$ penalty to coefficients of covariates in two-sample MR with GWAS summary statistics.
 Importantly, their method doesn’t penalize the SNP-exposure association.
@@
 Like MR-Clust, MR-PATH, as the method from @iong_latent_2020 is called, works with the Wald ratio estimates from multiple SNP instruments.
-Coining the term “mechanistic heterogeneity” to refer to the distribution of Wald ratio estimates due to SNP involvement in distinct causal pathways, @iong_latent_2020 specify a -component normal mixture model.
-They use an expectation maximization algorithm to fit the model parameters (the component weights and the parameters for each of the  normal distributions) and suggest a Bayesian information criterion-based approach to choosing  [@schwarz1978estimating].
+Coining the term “mechanistic heterogeneity” to refer to the distribution of Wald ratio estimates due to SNP involvement in distinct causal pathways, @iong_latent_2020 specify a $J$-component normal mixture model.
+They use an expectation maximization algorithm to fit the model parameters (the component weights and the parameters for each of the $J$ normal distributions) and suggest a Bayesian information criterion-based approach to choosing $J$ [@schwarz1978estimating].
 @wang_weakinstrument_2021 propose causal effect test statistics that are robust under weak instrument asymptotics [@staiger1994instrumental].
@@
 While these covariances could be estimated from individual-level data, such data are often unavailable in the GWAS setting.
-Importantly, the authors note that the estimands for multivariable MR nad MR differ; multivariable MR estimates the direct causal effects of each exposure on the outcome, while MR estimates the total causal effect of an exposure on the outcome.
+Importantly, the authors note that the estimands for multivariable MR and MR differ; multivariable MR estimates the direct causal effects of each exposure on the outcome, while MR estimates the total causal effect of an exposure on the outcome.
 MR-link, reported by @van_der_graaf_mendelian_2020, aims to account for unobserved pleiotropy and linkage disequilibrium by using individual-level data on the outcome variable and summary statistics for the exposure.
@@
 
-The second type of horizontal pleiotropy, from above, is sometimes called “correlated horizontal pleiotropy.” Correlated horizontal pleiotropy occurs when a SNP affects both the exposure and a variable that confounds the relationship between exposure and outcome (Figure 3B).
+The second type of horizontal pleiotropy, from above, is sometimes called “correlated horizontal pleiotropy.” Correlated horizontal pleiotropy occurs when a SNP affects both the exposure and a variable that confounds the relationship between exposure and outcome (Figure 3C).
 CAUSE and MRAID are two methods that aim to model correlated horizontal pleiotropy [@morrison_mendelian_2020;@yuan_likelihood_2021].
@@
 We focus on the approach that uses GWAS summary statistics.
-Two equations are central to MRAID modeling (Equations (7.1) and (7.2)).
+Two equations are central to MRAID modeling (Equations \@ref(eq:mraid1) and \@ref(eq:mraid2)).
 
-The estimated marginal effects for the  SNP instruments on the exposure are denoted by , while  represents the estimated effects of the same SNPs on the outcome.
-The  by  SNP instrument correlation matrices for the exposure and outcome are written as  and , respectively.
-and  can be estimated with 1000 Genomes Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry [@the_1000_genomes_project_consortium_1000_2012].
-The -vector error terms,  and , follow multivariate normal distributions with mean zero and variances  and , respectively.
+\begin{equation}
+  \hat{\beta_X} = \Sigma_1\beta + e_x
+  (\#eq:mraid1)
+\end{equation}
+
+\begin{equation}
+  \hat{\beta_Y} = \alpha \Sigma_2 \beta + \Sigma_2\eta_0 + \Sigma_2 \eta_1 + e_y
+  (\#eq:mraid2)
+\end{equation}
+
+The estimated marginal effects for the $p$ SNP instruments on the exposure are denoted by $\hat \beta_X$, while 
+$\hat\beta_Y$ represents the estimated effects of the same SNPs on the outcome. The $p$ by $p$ SNP instrument correlation matrices for the exposure and outcome 
+are written as $\Sigma_1$ and $\Sigma_2$, respectively. $\Sigma_1$ and $\Sigma_2$ can be estimated with 1000 Genomes 
+Project data, for example, by choosing a subset of 1000 Genomes Project subjects that have similar ancestry 
+[@the_1000_genomes_project_consortium_1000_2012]. The $p$-vector error terms, $e_x$ and $e_y$, follow multivariate normal 
+distributions with mean zero and variances $\Sigma_1\frac{\sigma_X^2}{n_1 - 1}$ 
+and $\Sigma_2\frac{\sigma_Y^2}{n_2 - 1}$, 
+respectively.
 @yuan_likelihood_2021 then constructs a Gibbs sampler to make likelihood-based inferences [@geman1984stochastic].
@@
     4. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.
+
 By encoding these assumptions in prior distributions, @yuan_likelihood_2021 enables the inferential procedures to accommodate the observed data in the context of the assumptions on SNPs.

### 2022-04-27 reply-to-review1/mr-review.Rmd
added practice recs

@@
 
-## Practice recommendations
-
 ## Binary, Count, and Time-to-event outcomes and exposures
@@
 ## Recent Findings
+
 ### Omnigenic MR
+
 Recent research on the omnigenic hypothesis, which posits that every SNP’s effect on a trait is nonzero, has informed MR methods development [@boyle2017].
@@
 These causal relationships highlight the important roles of plasma lipids, blood pressure, and the immune system in CAD susceptibility and those of obesity and the immune system in asthma development.
+
 ### Sample overlap in biobanks
+
 The issue of overlapping samples in the two-sample MR with GWAS summary data has recently received attention in the published literature.
@@
 ### Bi-directional causal inference with MR to account for reverse causation
+
 Reverse causation occurs when the outcome at an early time point has a causal effect on the exposure variable.
@@
 
+## Practice recommendations
+
+We now turn attention to the question of how to choose among the MR methods for a given analysis.
+We focus on the two-sample setting where only GWAS summary statistics are available.
+The large number of two-sample MR methods with summary statistics may challenge  a practitioner seeking to develop
+an analysis plan. Early considerations include:
+
+1. assessing the plausibility of the MR assumptions  
+1. determining which questions to address with sensitivity analysis  
+1. assessing causal directionality between exposure and outcome variables
+
+In some settings, a practitioner may address these questions by a careful study of the 
+relevant scientific literature. For example, there may be experimental studies that 
+suggest a causal direction for the relationship between exposure and outcome. 
+If both causal directions are plausible, then a bidirectional MR analysis may be needed. 
+If the bidirectional MR analysis (and supporting scientific evidence) is consistent with a single causal direction, then
+a practitioner may proceed with unidirectional MR methods.
+Given the pervasive nature of horizontal pleiotropy and its potential impact on MR, we recommend using one or more MR methods explicitly model horizontal pleiotropy.
+Recent methods, such as CAUSE and MRAID, go a step further by modeling correlated horizontal pleiotropy [@morrison_mendelian_2020;@yuan_likelihood_2021].
+
+Other MR methods can be used in sensitivity analysis to assess robustness of findings to 
+possible violations of MR assumptions.
+Additionally, methods like MR-clust have the potential to reveal additional relationships among genes that share a biological pathway [@foley_mr-clust_2021].
+The STROBE-MR statement has additional practice recommendations [@skrivankova_strengthening_2021].
+
 ## Summary and Outlook

### 2022-04-28 reply-to-review1/mr-review.Rmd
inserted paragraph breaks

@@
 date: "`r lubridate::now()`"
-bibliography: ../exported.bib
 csl: ../elsevier-with-titles.csl
@@
 linkcolor: blue
-always_allow_html: true
+always_allow_html: yes
+bibliography: ../exported.bib
 output:
-  - bookdown::pdf_document2
-  - bookdown::html_document2
-  - bookdown::word_document2
+  bookdown::word_document2:
+    number_sections: false
 ---
@@
 MR analysis is a form of instrumental variable analysis in which genetic markers, typically single nucleotide polymorphisms (SNPs), serve as instruments, or proxies, for inferring causal effects of an exposure variable on an outcome variable [@wright1928tariff;@wright1921correlation;@wright1923theory;@stock2003retrospectives;@katan1986apolipoprotein;@katan2004commentary;@didelez2007mendelian].
+
 MR analysis is facilitated by the development of genome-wide association studies (GWAS), which present unique opportunities for discovery of causal relationships via MR [@swerdlow_selecting_2016].
@@
 
-```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for keyword searches: 1. Mendelian randomization and 2. genome-wide association study."}
+```{r, article-counts, message = FALSE, fig.cap= "Upward trend in article counts by year for Google Scholar keyword searches: 1. Mendelian randomization and 2. genome-wide association study."}
 library(magrittr)
@@
   ggplot2::xlab("Year") + 
-  ggplot2::ggtitle("Published journal articles per year for Google Scholar key word searches: 1. Mendelian randomization and 2. Genome-wide association study") + 
+  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = -90, 
+                                                     vjust = 0.5, 
+                                                     hjust=1)) +
+ # ggplot2::ggtitle("Published journal articles per year") + 
   ggplot2::guides(fill = ggplot2::guide_legend("Search terms")) + 
@@
 The overall trend of the methodology development is in the direction of increasingly sophisticated modeling of horizontal pleiotropy including both independent and correlated horizontal pleiotropy while attempting to maintain scalability and computational efficiency in the presence of multiple correlated SNPs.
+
 Here, we present a comprehensive review on 47 MR methods, primarily in the context of GWAS summary statistics, to help practitioners to choose which MR methods to use in applied data analysis.
@@
 We focus on the two-sample MR design before considering the one-sample design and the partial two-sample design where samples are partially overlapped between the two cohorts.
+
 MR aims to infer the causal effect of the exposure variable on the outcome variable in observational studies.
@@
 The third assumption ensures that the SNP instruments are independent of all confounders of the exposure-outcome relationship.
+
 When the three MR assumptions hold, one may use instrumental variable statistical methods, with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].
@@
 Sensitivity analysis is highly recommended as a way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014].
-We will discuss these sensitivity analyses in more detail in a late section and consider their evaluation in MR studies [@vanderweele_methodological_2014].
+We will discuss these sensitivity analyses in more detail in a later section and consider their evaluation in MR studies [@vanderweele_methodological_2014].
+
 The causal assumptions of MR also help one to understand the origin of the term “Mendelian randomization”.
-Recall Mendel’s inheritance laws [mendel1865versuche], under the assumption that alleles segregate randomly from parent to offspring, the offspring genotypes are unlikely to be associated with confounders of the exposure-outcome relationship.
+Recall Mendel’s inheritance laws [@mendel1865versuche], under the assumption that alleles segregate randomly from parent to offspring, the offspring genotypes are unlikely to be associated with confounders of the exposure-outcome relationship.
 Additionally, reverse causation from the outcome or exposure to the genotypes is unanticipated since germ-line genotypes are fixed at conception and, thus, precede realization of other observed variables.
@@
 ## Statistical models and methods for MR with one instrument
+
 We begin our discussion of Mendelian randomization in GWAS data by considering the simplest case, where there is a single SNP instrument and a single outcome variable.
@@
 ### Ratio of coefficients method
+
 The ratio of coefficients method, also known as the Wald method, estimates the causal effect of the exposure X on the outcome Y by using a single SNP instrument [@wald_fitting_1940].
@@
 ### Two-stage methods
+
 A two-stage statistical model involves two regression models [@burgess_review_2017;@sawa_exact_1969].
@@
 Note that the uncertainty in the fitted values from the first regression is not considered when performing the second regression.
-For this reason, the variance of the coefficient estimator,  is incorrect in two-stage calculations [@burgess_review_2017;@angrist2008mostly].
+For this reason, the variance of the estimator is incorrect in two-stage calculations [@burgess_review_2017;@angrist2008mostly].
 This and other observations led researchers to develop likelihood-based MR methods.
@@
 ### Likelihood-based methods
+
 Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood estimates with their many desirable properties [@fisher1925theory].
@@
 Additionally, the limited information maximum likelihood framework can accommodate  more than one SNP instrument by replacing $\alpha_1z_{i1}$ with the sum $\sum_{k = 1}^K\alpha_k z_{ik}$.
+
 One may also use Bayesian methods to obtain likelihood-based estimators [@kleibergen_bayesian_2003].
@@
 ### Semiparametric methods
+
 Semiparametric instrumental variable methods, which feature parametric and nonparametric components, typically assume a parametric model connecting the exposure and outcome, but don’t impose distributional assumptions on the errors [@burgess_review_2017].
@@
 Another semiparametric estimation method is G-estimation of a structural mean model [@burgess_review_2017;@robins_new_1986;@greenland_estimating_2008].
-We follow [@burgess_review_2017] by defining the potential outcome  as the outcome value that we would have observed had we set the exposure value  to .
-For example,  denotes the observed outcome had we set the exposure to zero instead of its observed value .
-The structural mean model for a continuous outcome is displayed in Equation (3.8).
-
-$Y(x)$ as the outcome value that we would 
-have observed had we set the exposure value $X$ to $x$. For example, $Y(0)$ denotes the observed outcome had 
+We follow [@burgess_review_2017] by defining the potential outcome $Y(x)$ as the outcome value that we would have observed had we set the exposure value $X$ to $x$.
+For example, $Y(0)$ denotes the observed outcome had
 we set the exposure to zero instead of its observed value $x$. The structural mean model for a continuous 
@@
 ## Mendelian Randomization with multiple independent SNP instruments
+
 While using a single SNP instrument for MR analysis can be effective, @smith_mendelian_2004 recognized that many genetic variants individually explain a small proportion of the variation in a trait, and, thus, sufficient statistical power of MR analysis with a single SNP instrument would require sample sizes in the tens of thousands [@burgess_mendelian_2013;@schatzkin2009mendelian].
@@
 Recognition of complex traits’ diverse genetic architectures has thus fueled development of two-sample MR methods with multiple SNP instruments [@bowden_mendelian_2015;@hormozdiari_leveraging_2018;@watanabe_global_2019;@pierce_power_2011].
+
 There are two important considerations for MR analysis with multiple SNP instruments.
@@
 ## Statistical models and methods for MR with multiple instruments
+
 ### Methods with Individual-level data
+
 One approach to integrating multiple SNP instruments into a MR framework is through calculation of allele scores [@burgess_use_2013].
@@
 @burgess_use_2013 found that when weights are obtained from external data or from cross-validation or jackknife approaches applied to the analysis data, the allele score functions as a single instrumental variable and greatly diminishes the bias compared to that of the two-stage least squares estimator [@burgess_review_2017;@angrist1999jackknife].
+
 @kang_instrumental_2016 developed a LASSO-based method, sisVIVE, to identify invalid SNP instruments [@tibshirani_regression_1996].
@@
 Simulations and data analysis results reveal that sisVIVE is robust to possibly invalid instruments.
+
 @windmeijer_use_2019, building on the research from @kang_instrumental_2016, implemented an adaptive LASSO-based estimator after recognizing that sisVIVE misclassifies valid SNP instruments as invalid when the invalid SNP instruments have strong effects on the exposure.
@@
 One important limitation of this thread of research is that both @kang_instrumental_2016 and @windmeijer_use_2019 require individual-level data.
+
 @jiang_constrained_2019, working with individual-level GWAS data, share naive and smoothed constrained instrumental variable methods.
@@
 @jiang_constrained_2019 find that both the smoothed version of their estimator and the allele score method are unbiased in their simulation settings.
+
 @spiller_detecting_2018 and @spiller_interaction-based_2020 present a method, MR-GxE, that exploits gene by environment interactions to detect and to adjust for bias due to horizontal pleiotropy.
@@
 For an even number, say $2K$, number of SNP instruments, choose the midpoint between the two middle-ranking ordered ratio estimates, $\frac{\tilde \beta_K + \tilde \beta_{K + 1}}{2}$.
+
 Due to the inefficiency of the simple median estimator, @bowden_consistent_2016 define the weighted median estimator.
@@
 In this setting, the simple median estimator is seen to be the weighted median estimator when all weights are equal.
+
 @bowden_consistent_2016 also studied the penalized weighted median estimator. They defined the 
@@
 Consequently, many MR methods have been recently developed to model multiple correlated instruments.
+
 @burgess_combining_2016 developed methods for using correlated SNPs in IVW MR by inserting a covariance matrix into the standard IVW formulation.
@@
 Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently."}
-knitr::include_graphics("../figures/dags.pdf")
+knitr::include_graphics("../figures/all_dags-two.png")
 ```
@@
 While absence of such associations doesn’t guarantee satisfaction of the assumption, presence of SNP-covariate associations must be investigated carefully, as they may constitute assumption violations [@burgess_sensitivity_2017].
+
 @burgess_sensitivity_2017 considered a collection of sensitivity analysis methods when working with GWAS summary statistics.
@@
 While they can’t rule out the possibility of unmeasured confounding, they can study the possibility of measured covariates serving as confounders.
+
 To illustrate their approach, @burgess_sensitivity_2017 shared a case study in which they examined the causal effect of C-reactive protein (CRP) levels on coronary artery disease risk with four genetic variants in the CRP gene region and 17 other genetic variants that affect coronary artery disease risk.
@@
 Additionally, Egger regression can be useful in this setting [@egger_bias_1997;@bowden_mendelian_2015;@burgess_interpreting_2017;@bowden_framework_2017].
+
 @spiller_interaction-based_2020 point out that MR-GxE can be used as a sensitivity analysis tool to choose a set of valid SNP instruments from a collection of SNPs in settings where there may be a violation of the constant pleiotropy assumption.
+
 MR-GENIUS is a framework for two-sample MR analysis when individual-level data are available [@tchetgen_tchetgen_genius_2021].
@@
 ## Explicit modeling of horizontal pleiotropy
+
 Besides sensitivity analysis, several methods have been developed to directly model horizontal pleiotropy to ensure the validity of MR assumptions.
@@
 We discuss approaches to modeling and accounting for horizontal pleiotropy below.
+
 In the short time since publication of @verbanck_detection_2018, researchers have recognized two types of horizontal pleiotropy [@yuan_likelihood_2021].
@@
 Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
+
 @bowden_mendelian_2015 adapted Egger regression, originally developed to detect bias in meta-analyses, to detect bias from horizontal pleiotropy in two-sample MR studies with GWAS summary statistics or individual-level data [@egger_bias_1997].
@@
 @hemani_orienting_2017 applied their test in a study of DNA methylation and gene expression, where either direction of causality is plausible, and found that many methylation traits cause changes in gene expression.
+
 @hartwig_robust_2017 developed a weighted mode-based MR causal effect estimation method for two-sample MR analysis with GWAS summary data.
@@
 In this manner, the mode-based estimators demonstrated a greater robustness to horizontal pleiotropy than did the other estimators.
+
 @rees_robust_2019 approached the challenge of horizontal pleiotropy by developing three new methods for two-sample MR analysis with GWAS summary data: robust regression, penalized weights, and LASSO penalization.
@@
 They then work with profile log likelihood over a grid of causal parameter values to make inferences.
+
 @zhao_statistical_2020 presented MR-RAPS, a method that leverages a robust adjusted profile score to accomplish statistical inference in two-sample MR analyses with GWAS summary data.
@@
 They then demonstrate the properties of their estimator by analyzing simulated and real data [@zhao_statistical_2020].
+
 @bowden_improving_2019, working with a collection of SNP instruments, reasoned that if all three MR assumptions are satisfied, then the collection of Wald ratio estimates should be homogeneous; thus, any departures from homogeneity may signal a violation of one or more assumptions.
@@
 Their modified weights lead to an estimator that is useful for quantifying heterogeneity and detecting outliers.
+
 @cho_exploiting_2020 developed the MR-TRYX framework for exploiting horizontal pleiotropy to identify alternative causal pathways.
@@
 They follow up promising trait associations to account for their observations with multiple causal pathways.
+
 @fang2021evaluating recognized that in some MR studies, the Wald ratio estimates for different SNP instruments may form clusters, and these clusters may represent distinct causal mechanisms and may identify the SNPs that are involved in them.
 They devised a method to detect these clusters in the collection of Wald ratio estimates and found SNP instruments that point to distinct causal pathways of biological importance [@fang2021evaluating].
+
 @cheng_mr-ldp_2020 sought a probabilistic model that accounts for both linkage disequilibrium among SNP instruments and horizontal pleiotropy.
@@
 is its reliance on the InSIDE assumption and resulting inability to accommodate correlated horizontal pleiotropy.
+
 @zhu_iterative_2021 present IMRP, a method for causal effect estimation and horizontal pleiotropy detection.
@@
 Finally, the authors note that IMRP is three orders of magnitude faster than simulations-based MR-PRESSO.
+
 @grant_efficient_2020 build on research from @lin2015regularization to present a method that applies a $\ell_1$ penalty to coefficients of covariates in two-sample MR with GWAS summary statistics.
@@
 In so doing, they extend work from @burgess_multivariable_2015 and @burgess_re_2015 to get a causal effect estimator that needs no valid SNP instruments and is robust to measured vertical pleiotropy and more statistically efficient than previous MR methods.
+
 @xu_robust_2021 consider two classes of horizontal pleiotropy with their method, MRCIP.
@@
 One direction for future study is to augment MRCIP to permit unbalanced horizontal pleiotropy.
+
 @iong_latent_2020 build on other research using mixture models in the two-sample MR setting with GWAS summary statistics.
@@
 They use an expectation maximization algorithm to fit the model parameters (the component weights and the parameters for each of the $J$ normal distributions) and suggest a Bayesian information criterion-based approach to choosing $J$ [@schwarz1978estimating].
+
 @wang_weakinstrument_2021 propose causal effect test statistics that are robust under weak instrument asymptotics [@staiger1994instrumental].
@@
 @wang_weakinstrument_2021 extend their methods to demonstrate conditions under which their estimators are equivalent to those in works from @zhao_statistical_2020, @windmeijer_use_2019, and @bowden_improving_2019.
+
 @bucur_inferring_2020 developed BayesMR, a Bayesian framework for two-sample MR with individual-level data and discuss an approximation for use with
@@
 After specifying the model, they present a nested sampling scheme [@skilling2006nested;@handley2015polychord1;@handley2015polychord] that simultaneously computes the model evidence for both causal directions (i.e., exposure causes outcome and outcome causes exposure) and acquires samples from the posterior distribution for Bayesian statistical inferences.
+
 @schmidt_mendelian_2018 present BMRE, a Bayesian implementation of MR Egger.
@@
 The authors emphasize a potential role for BMRE in MR sensitivity analysis.
+
 @sanderson_examination_2019 found that multivariable MR, where multiple exposures jointly cause an outcome, consistently estimates the effect of the exposures on the outcome.
@@
 Importantly, the authors note that the estimands for multivariable MR and MR differ; multivariable MR estimates the direct causal effects of each exposure on the outcome, while MR estimates the total causal effect of an exposure on the outcome.
+
 MR-link, reported by @van_der_graaf_mendelian_2020, aims to account for unobserved pleiotropy and linkage disequilibrium by using individual-level data on the outcome variable and summary statistics for the exposure.
 MR-link uses a three-step procedure to estimate causal effects.
+
 1. Use GCTA-COJO to obtain SNP instruments [@genetic_investigation_of_anthropometric_traits_giant_consortium_conditional_2012]  
+2. Prune all SNPs in linkage disequilibrium with the SNP instruments to obtain a set of "tag" SNPs with correlations less than 0.95  
+3. Solve for the causal effect estimate with ridge regression of the outcome on the matrix resulting from concatenation of the SNP instrument genotypes matrix and the tag SNPs genotypes matrix.
 
-    2. Prune all SNPs in linkage disequilibrium with the SNP instruments to obtain a set of “tag” SNPs with correlations less than 0.95
-
-    3. Solve for the causal effect estimate with ridge regression of the outcome on the matrix resulting from concatenation of the SNP instrument genotypes matrix and the tag SNPs genotypes matrix.
 Note that in step 3, ridge regression is preferred to ordinary least squares due to the limited statistical power of ordinary least squares in this setting.
@@
 Specifically, they estimate the difference in the expected log pointwise posterior density for the model with causal effect fixed at zero and the model that permits a nonzero causal effect [@vehtari_practical_2017].
+
 @yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is biased.
@@
 To facilitate computations, the investigators make several assumptions about the collection of SNPs:
+
 1. Relatively small proportion of SNPs have nonzero effects on the exposure  
-
 2. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
-
 3. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs  
-
 4. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.
@@
 They implemented an iterative optimization algorithm to simultaneously infer the causal effect parameter and other parameters.
+
 Researchers have also leveraged other generalized linear models for analysis in MR studies [@nelder1972generalized].
@@
 With growing interest in modeling counts in biomedical data, where molecular phenotyping technologies now acquire RNA molecule counts and protein abundances, there may be opportunities for methods innovations in a generalized linear models framework for MR studies [@hwang_single-cell_2018;@ozsolak_rna_2011;@domon_mass_2006;@bantscheff_quantitative_2007].
+
 Recently, multiple teams of investigators have analyzed time-to-event traits in MR studies [@burgess_commentary_2015;@arvanitis_linear_2021;@tikkanen_body_2019;@he_causal_2018;@guo_body_2017].
@@
 
+Open research questions remain in the MR field.
+We have highlighted above three recent findings: omnigenic MR, sample overlap in biobanks, and bidirectional causal inference.
+While we discussed initial findings in these three areas, many opportunities for methods enhancements remain.
+For example, as biobank sizes grow in number of subjects and number of measured traits, there is increasing demand for MR methods that scale efficiently with large sample sizes.
+Development of methods for two-sample MR with overlapping samples opens opportunities to study rich biobank data, and we anticipate that this question - how to accommodate overlapping samples - will continue to be an active area of research.
+
 In summary, we have presented a comprehensive review on 47 methods for MR analysis in GWAS (Figure 4).

### 2022-04-28 reply-to-review1/old-eqns.Rmd
inserted paragraph breaks

@@
 
-# Assumption violations and Sensitivity Analysis
+# Assumption violations and sensitivity analysis
 
@@
 Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
-must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
+must affect the outcome only through the exposure. The causal diagram in Figure 3B illustrates 
 a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 

### 2022-04-28 reply-to-review1/mr-review.Rmd
fixed title

@@
 ---
-title: "Mendelian randomization with genome-wide association studies data"
+title: "Statistical methods for Mendelian randomization in genome-wide association studies: A review"
 author: "Frederick J. Boehm & Xiang Zhou"
-date: "`r lubridate::now()`"
+date: "`r lubridate::today()`"
 csl: ../elsevier-with-titles.csl
@@
 The GMM estimate is the value of the vector $\beta$ that satisfies Equation \@ref(eq:gmm), where $f(x_i; \beta) = \mathbb{E}(Y|\text{do}(X = x))$.
-@palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equations \@ref(eq:gmm).
+@palmer_instrumental_2011 developed numerical methods for obtaining estimates from Equation \@ref(eq:gmm).
 
@@
 Let’s first consider the exclusion restriction assumption (assumption 2 above).
-It states that the SNP instrument must affect the outcome only through the exposure.
-The causal diagram in Figure 3B illustrates a scenario that violates this assumption [@vanderweele_methodological_2014].
+It states that the SNP instrument must affect the outcome only through the exposure (Figure 3B).
+The causal diagram in Figure 3A illustrates a scenario that violates this assumption [@vanderweele_methodological_2014].
 Bias in causal effect estimation would result.
@@
 A.
-Scenario where the genetic variant affects an exposure, which in turn affects an outcome.
-B.
 Scenario where the genetic variant affects an intermediate variable on the pathway to the exposure.
 Because the intermediate affects the outcome through a pathway that doesn’t involve exposure, this scenario violates the exclusion restriction assumption.
+B.
+Scenario where the genetic variant affects an exposure, which in turn affects an outcome, possibly in the presence of unmeasured confounding.
 C.
 Correlated horizontal pleiotropy occurs when the genetic variant affects the exposure, which in turn affects the outcome, and the genetic variant affects the unobserved confounder, which in turn affects the exposure and the outcome independently."}
-knitr::include_graphics("../figures/all_dags-two.png")
+knitr::include_graphics("../figures/all_dags.png")
 ```
@@
 possible violations of MR assumptions.
-Additionally, methods like MR-clust have the potential to reveal additional relationships among genes that share a biological pathway [@foley_mr-clust_2021].
+Methods like MR-clust have the potential to reveal additional relationships among genes that share a biological pathway [@foley_mr-clust_2021].
 The STROBE-MR statement has additional practice recommendations [@skrivankova_strengthening_2021].
@@
 Open research questions remain in the MR field.
-We have highlighted above three recent findings: omnigenic MR, sample overlap in biobanks, and bidirectional causal inference.
+We have highlighted above three recent findings: omnigenic MR, methods for sample overlap in biobanks, and methods for bidirectional causal inference.
 While we discussed initial findings in these three areas, many opportunities for methods enhancements remain.
